### 2021-01-21 draft_paper.Rmd
kmeans first run done

---
title: 'Local geography of the COVID-19 crime drop: the first six months'
output: pdf_document
header-includes:
  - \usepackage{float}
---

```{r setup, include=FALSE, echo = F}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

- An evidence-base has begun to emerge around the impact of the COVID-19 pandemic on crime and anti-social behaviour.
- Findings have tended to demonstrate a widespread decline in police-recorded crime, although the extent to which this holds true varies by crime type.
- But, most studies have focused on macro-level units of analysis, such as cities or countries.
- A recent study took a 'look back' on the first six months of lockdown using police-recorded crime data aggregated to England and Wales.
- The macro-level longitudinal patterns observed were consistent with opportunity perspectives on crime.
- Here, we disentangle this macro-level trend using localized data aggregated to neighbourhood units across England and Wales. 
- We focus on the extent to which localized areas remained stable, or were subject to volatility, amidst the nationwide change in police-recorded crime.
- This includes a descriptive analysis of concentration, an identification of the 'drivers' of the lockdown crime drop, and the characteristics of localized areas driving the change.
- In doing so, we disentangle the macro-level crime trends observed during lockdown in England and Wales.

## Background

- Overview of findings on lockdown crime to date.
- Note that a key component missing relates to spatial concentrations, and the extent to which macro-level trends mask underlying, localized variation.
- These discussions have proved vital to crime and place research, and our understanding of criminal opportunity and routine activities theory (e.g. law of crime concentration).
- Localized analysis may also shed light on policing and resource allocation during lockdown.
- In England and Wales, we already have a reasonable understanding of the macro-level trends.

```{r crimeoverall, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Macro-level crime trends in England and Wales during the lockdown period in 2020."}
knitr::include_graphics("visuals/raw_counts_gg.png")
```

## Data and Method

- Generalized Gini coefficient for all offence categories.
- Focus on two broad categories: notifiable offences and anti-social behaviour (ASB).
- Decile change to establish whether previously high-crime/ASB areas remained, ceased or became increasingly problematic during lockdown.
- Non-parametric longitudinal clustering (k-means) used to disentangle the macro-level trends observed during the lockdown, and identify the 'drivers' of the dramatic drop and subsequent resurgence.
- This speaks to existing literature and methods deployed in crime and place research.

## Results

### Gini

```{r crimegini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Measure of concentration using the generalized Gini coefficient during the lockdown period."}
knitr::include_graphics("visuals/gini_gg.png")
```

### Deciles

```{r decilenotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Decile changes in notifiable offences."}
knitr::include_graphics("visuals/deciles_tot_an_gg.png")
```

```{r decileasb, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Decile changes in anti-social behaviour."}
knitr::include_graphics("visuals/deciles_asb_an_gg.png")
```

### Longitudinal clustering

```{r kmeansnotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA notifiable offences.", out.width= "90%"}
knitr::include_graphics("visuals/kmeans_violin_gg.png")
```

```{r kmeansasb, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA ASB.", out.width= "90%"}
knitr::include_graphics("visuals/kmeans_violin_asb_gg.png")
```

### Spatial distribution of clusters

### Characteristics of clusters

## Discussion

### 2021-01-21 scripts/asb_kmeans.R

@@ -1,0 +1,94 @@
+# Use longitudinal k-means to unpick the total noficiable crimes (excl. drugs trend) at LSOA level.
+tc_asb_kmeans_df <- sub_data_agg_1920_df %>%
+  filter(crime_type == "Anti-social behaviour",
+         month != "2019-12") %>% 
+  group_by(lsoa_code, month) %>% 
+  summarise(ew_crime_count = sum(crime_count)) %>% 
+  ungroup() 
+
+# For now, only keep the lockdown period in the first visual: March to August.
+tc_asb_kmeans_sub_df <- tc_asb_kmeans_df %>% 
+  filter(month == "2020-03" | month == "2020-04" | month == "2020-05" | month == "2020-06" |
+         month == "2020-07" | month == "2020-08")
+
+# Check for those LSOAs which were essentially crime-free during the study period.
+# We keep in wide because that's what kml likes.
+tc_asb_kmeans_sub_clean_df <- tc_asb_kmeans_sub_df %>% 
+  pivot_wider(id_cols = lsoa_code, names_from = "month", values_from = "ew_crime_count") %>%
+  mutate(total_months = rowSums(.[2:7])) %>% 
+  filter(total_months != 0) %>% # N = 111
+  select(-total_months) %>% 
+  rename(march = `2020-03`, april = `2020-04`, may = `2020-05`, june = `2020-06`,
+         july = `2020-07` , august = `2020-08`)
+
+# Perform kmeans
+n <- 3:6
+
+tc_asb_kml_mat <- as.matrix(tc_asb_kmeans_sub_clean_df[2:7])
+tc_asb_traj    <- clusterLongData(traj = tc_asb_kml_mat) 
+kml(tc_asb_traj, nbClusters = n, toPlot = "criterion", nbRedrawing = 20)
+
+# Apppend clusters back with data frame.
+tc_asb_clusters <- cbind.data.frame(lsoa_code   = tc_asb_kmeans_sub_clean_df$lsoa_code,
+                                traj       = getClusters(tc_asb_traj, 3)) # We select 4 due to CH value.
+
+tc_asb_clusters_df <- tc_asb_kmeans_sub_clean_df %>% 
+  left_join(tc_asb_clusters) %>% 
+  pivot_longer(cols = c(-lsoa_code, -traj), names_to = "month", values_to = "ew_crime_counts") %>% 
+  mutate(month_fac = as.factor(month),
[…]

### 2021-01-24 scripts/asb_kmeans.R
draft start

@@ -82,8 +82,8 @@ tc_asb_clusters_props_df <- tc_asb_clusters_df %>%
 
 # Plot stacks
 tc_asb_props_gg <- ggplot(data = tc_asb_clusters_props_df) +
-  geom_area(mapping = aes(x = month_fac, y = sum_monthly_props, group = traj_titles, fill = traj_titles),
-            alpha = 0.8) +
+  geom_bar(mapping = aes(x = month_fac, y = sum_monthly_props, group = traj_titles, fill = traj_titles),
+            alpha = 0.8, stat = "identity") +
   scale_x_discrete(labels = c(str_extract(month.name[3:8], "^.{3}"), character(1))) +
   labs(x = NULL, y = "% total ASB", fill = NULL) +
   theme_bw() +

### 2021-01-24 scripts/police_data_handling.R

@@ -394,6 +394,11 @@ tc_clusters_df <- tc_clusters_df %>%
 # Check.
 table(tc_clusters_df$traj_titles)
 
+# Save for Anthony
+# tc_clusters_df %>% 
+#   distinct(lsoa_code, traj, traj_titles) %>% 
+#   write_csv(path = "results/tc_clusters_df.csv")
+
 # Plot (violin plot).
 kmeans_violin_gg <- ggplot(data = tc_clusters_df,
                           mapping = aes(x = month_fac, y = ew_crime_counts,#
@@ -423,8 +428,8 @@ tc_clusters_props_df <- tc_clusters_df %>%
 
 # Plot stacks
 tc_props_gg <- ggplot(data = tc_clusters_props_df) +
-  geom_area(mapping = aes(x = month_fac, y = sum_monthly_props, group = traj_titles, fill = traj_titles),
-            alpha = 0.8) +
+  geom_bar(mapping = aes(x = month_fac, y = sum_monthly_props, group = traj_titles, fill = traj_titles),
+            alpha = 0.8, stat = "identity") +
   scale_x_discrete(labels = c(str_extract(month.name[3:8], "^.{3}"), character(1))) +
   labs(x = NULL, y = "% total crime", fill = NULL) +
   theme_bw() +
@@ -436,9 +441,9 @@ ggsave(plot = tc_props_gg, filename = "visuals/tc_props_gg.png",
 
 # Save and load workspace as appropriate.
 # save.image(file = "data_handling.RData")
-# load(file = "data_handling.RData")
+load(file = "data_handling.RData")
 
-march = "march",# We know that the most change occurred in April, across all crimes types.
+# We know that the most change occurred in April, across all crimes types.
 # But what areas were driving this change? Was it previously criminal areas?
 # Here, we calculate which LSOAs are driving the change observed in April.
 
@@ -453,23 +458,6 @@ aprils_lsoa_df <- sub_data_agg_1920_df %>%
   mutate(decile = ntile(cc_total_plus, 10)) %>%
   ungroup() %>%
   arrange(decile)
-
-# How much have the deciles changed?
-aprils_lsoa_wide_df <- aprils_lsoa_df %>%
-  pivot_wider(id_cols = lsoa_code, names_from = "month", values_from = "decile") %>% 
-  rename(april19 = `2019-04`,
[…]

### 2021-01-24 scripts/surplus_material.R

@@ -1,0 +1,316 @@
+
+# Calculate Spearman's Rank correlation *between months of different years*.
+table(sub_data_agg_1920_df$month)
+
+# Remove December 2019 (until December 2020 is released), and split into list of df by crime type.
+cor_prep_list <- sub_data_agg_1920_df %>% 
+  filter(month != "2019-12") %>% 
+  select(crime_type, month, lsoa_code, crime_count) %>% 
+  group_split(crime_type) 
+
+# Add names.
+names(cor_prep_list) <- unique(sub_data_agg_1920_df$crime_type)
+
+# Function to pivot each df in list long to wide.
+cor_prep_fun <- function(x) {
+  x %>% 
+    select(-crime_type) %>% 
+    pivot_wider(id_cols = lsoa_code, names_from = month, values_from = crime_count) %>% 
+    select(-lsoa_code)
+}
+
+# Run function through list of df.
+cor_output_list <- lapply(cor_prep_list, cor_prep_fun)
+
+# Spearman's rank correlation on each.
+cor_results_list <- lapply(cor_output_list, function(x){as_tibble(round(cor(x = x, method = "spearman"), 2))})
+
+# Create month variable for each correlation table.
+cor_results_pull_list <- lapply(cor_results_list, function(x){x %>%
+    mutate(month_lag = names(.)) %>% 
+    relocate(month_lag, .before = `2019-01`) %>% 
+    slice(12:22) %>% 
+    select(1:12) })
+
+# Save each.
+names(cor_results_pull_list) 
+for(i in seq_along(cor_results_pull_list))
+  write_csv(x = cor_results_pull_list[[i]], path = paste0("results/", names(cor_results_pull_list)[i], "_month_cor", ".csv"))
+
+# # Retrieve the diagnal, which compared the same months of different years.
+# cor_results_sub_list <- lapply(cor_results_list, function(x){diag(as.matrix(x[, -1]))})
+# 
+# # Stick together and create id column. Rows are in the order of months.
+# cor_results_sub_df <- cor_results_sub_list %>% 
+#   bind_rows() %>%
+#   mutate(cor_compare = names(cor_results_list[[1]]))
+# 
+# 
[…]

### 2021-01-27 scripts/police_data_handling.R
crime chars and perc change gg

@@ -416,6 +416,9 @@ kmeans_violin_gg <- ggplot(data = tc_clusters_df,
 ggsave(plot = kmeans_violin_gg, filename = "visuals/kmeans_violin_gg.png",
        height = 20, width = 20, unit = "cm", dpi = 200)
 
+# Add 2019 years to it.
+
+
 # Create proportion contribution to total crime in each month.
 tc_clusters_props_df <- tc_clusters_df %>%
   group_by(month_fac) %>% 
@@ -439,11 +442,108 @@ tc_props_gg <- ggplot(data = tc_clusters_props_df) +
 ggsave(plot = tc_props_gg, filename = "visuals/tc_props_gg.png",
        height = 14, width = 16, unit = "cm", dpi = 200)
 
+# Calculate the % of month-on-month change each cluster contributed to.
+change_df <- tc_clusters_df %>% 
+  select(lsoa_code, month, traj_titles, ew_crime_counts) %>% 
+  group_by(month, traj_titles) %>%
+  summarise(traj_crimes = sum(ew_crime_counts)) %>%
+  ungroup() %>%
+  group_by(month) %>%
+  mutate(total_crimes = sum(traj_crimes)) %>% 
+  pivot_wider(id_cols = traj_titles, names_from = month, values_from = c(traj_crimes, total_crimes)) %>% 
+  mutate(tot_mar_apr  = total_crimes_april-total_crimes_march,
+         tot_apr_may  = total_crimes_may-total_crimes_april,
+         tot_may_jun  = total_crimes_june-total_crimes_may,
+         tot_jun_jul  = total_crimes_july-total_crimes_june,
+         tot_jul_aug  = total_crimes_august-total_crimes_july,
+         traj_mar_apr = traj_crimes_april-traj_crimes_march,
+         traj_apr_may = traj_crimes_may-traj_crimes_april,
+         traj_may_jun = traj_crimes_june-traj_crimes_may,
+         traj_jun_jul = traj_crimes_july-traj_crimes_june,
+         traj_jul_aug = traj_crimes_august-traj_crimes_july,
+         prop_mar_apr = traj_mar_apr/tot_mar_apr,
+         prop_apr_may = traj_apr_may/tot_apr_may,
+         prop_may_jun = traj_may_jun/tot_may_jun,
+         prop_jun_jul = traj_jun_jul/tot_jun_jul,
[…]

### 2021-01-27 scripts/police_data_handling.R
kmean cluster visual 2019 lines

@@ -417,7 +417,47 @@ ggsave(plot = kmeans_violin_gg, filename = "visuals/kmeans_violin_gg.png",
        height = 20, width = 20, unit = "cm", dpi = 200)
 
 # Add 2019 years to it.
+traj_names_df <- tc_clusters_df %>% 
+  distinct(lsoa_code, traj, traj_titles)
+  
+tc_clusters_2019_df <- sub_data_agg_1920_df %>%
+  filter(year == "2019",
+         crime_type != "Anti-social behaviour" & crime_type != "Drugs") %>% 
+  group_by(month, lsoa_code) %>% 
+  summarise(ew_crime_counts = sum(crime_count)) %>% 
+  ungroup() %>% 
+  left_join(traj_names_df) %>% 
+  drop_na(traj_titles) %>% # the LSOA dropped for clustering
+  filter(month != "2019-02", month != "2019-01", month != "2019-09", month != "2019-10", month != "2019-11", month != "2019-12") %>% 
+  mutate(month = as_factor(month),
+         month_fac = fct_recode(month, march  = "2019-03",
+                                       april  = "2019-04",
+                                       may    = "2019-05",
+                                       june   = "2019-06",
+                                       july   = "2019-07",
+                                       august = "2019-08"))
 
+kmeans_violin_2019_gg <- ggplot(data = tc_clusters_df,
+       mapping = aes(x = month_fac, y = ew_crime_counts,#
+                     fill = traj_titles)) +
+  geom_violin(alpha = 0.3, colour = "transparent", adjust = 2) + # to match asb - makes no difference.
+  stat_summary(aes(group = traj_titles), fun = "median", colour = "black", size = 0.5, geom = "line") +
+  stat_summary(aes(group = traj_titles), fun = "mean", colour = "black", linetype = "dotted", size = 0.5, geom = "line") +
+  facet_wrap(~traj_titles, ncol = 2, scales = "free_y") +
+  stat_summary(data = tc_clusters_2019_df,
+               mapping = aes(x = month_fac, y = ew_crime_counts, group = traj_titles),
+               fun = "median", colour = "red", size = 0.5, geom = "line") +
+  stat_summary(data = tc_clusters_2019_df,
+               mapping = aes(x = month_fac, y = ew_crime_counts, group = traj_titles),
+               fun = "mean", colour = "red", linetype = "dotted", size = 0.5, geom = "line") +
[…]

### 2021-01-27 draft_paper.Rmd
draft paper edits

@@
 - Overview of findings on lockdown crime to date.
-- Note that a key component missing relates to spatial concentrations, and the extent to which macro-level trends mask underlying, localized variation.
+- Note that a key missing component relates to spatial concentrations, and the extent to which macro-level trends mask underlying, localized variation.
 - These discussions have proved vital to crime and place research, and our understanding of criminal opportunity and routine activities theory (e.g. law of crime concentration).
@@
 
-```{r decileasb, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Decile changes in anti-social behaviour."}
-knitr::include_graphics("visuals/deciles_asb_an_gg.png")
+### Longitudinal clustering
+
+#### Cluster trends
+
+- General conclusion: most of the lockdown crime drop can be attributed to a small number LSOAs.
+- Most localized areas actually experienced minimal crime change during the pandemic.
+- The proportion of total crime attributable to very low and very high crime clusters fell into April, while mid-crime areas increased their proportion.
+- All clusters converged back to 'normality' by the end of August.
+- Remarkable stability given the absolute change that occurred during this study period.
+- 2020 mean (dotted line) and median (solid line) of each cluster are shown in black.
+- 2019 mean (dotted line) and median (solid line) of each cluster are shown in red.
+- This demonstrates that the clusters identified using the 2020 data were distinct and meaningful even when also applied to the 2019 data.
+
+```{r kmeansnotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA notifiable offences.
+Mean (dotted line) and median (solid line) are shown for each cluster for 2020 (black) and 2019 (red) respectively.
+Violin plots refer to 2020 only.", out.width= "90%"}
+knitr::include_graphics("visuals/kmeans_violin_2019_gg.png")
 ```
 
-### Longitudinal clustering
+#### Contribution of each cluster to change
 
-```{r kmeansnotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA notifiable offences.", out.width= "90%"}
-knitr::include_graphics("visuals/kmeans_violin_gg.png")
+```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Proportion of month-on-month change in crime attributable to each cluster", out.width= "90%"}
+knitr::include_graphics("visuals/change_gg.png")
 ```
 
-```{r kmeansasb, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA ASB.", out.width= "90%"}
-knitr::include_graphics("visuals/kmeans_violin_asb_gg.png")
-```
+- The above summarises the proportion of month-to-month change in crime counts attributable to each cluster.
+- In other words, we try and use this to identify which clusters drove the initial decline and subsequent resurgence nationwide.
+- For example, the red cluster (N = 23,924), despite consisting of 72% of total LSOAs, contributed to around 30% of the drop in notifiable offences between March and April.
+- While, the pink cluster (N = 83) despite comprising only 0.25% of LSOAs, contributed to 12.5% of the fall in notifiable offences between March and April.
+- July to August is weird, because although total crime increased slightly between those months, counts actually decreased for the red and green groupings, and increased for blue and purple.
+So, the blue and purple groupings drove this final nationwide increase on their own, countering the decline observed in red and green LSOAs.
+I think this is why the proportional figures do not add up to 1 for this final monthly change, so I need to think of a better way of summarizing this.
+- I haven't worked out how to visualize things yet - it needs revisiting, but hopefully gives you an idea.
 
@@
 
+- List major city composition?
+- Maps of major city case studies.
+
 ### Characteristics of clusters
 
+Descriptive statistics about the LSOAs in each cluster solution.
+
+#### Crime type profile
+
+- What crime types tend to comprise each cluster?
+
+```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type characteristics of each cluster solution.", out.width= "90%"}
+knitr::include_graphics("visuals/traj_crimes_gg.png")
+```
+
+### Deprivation
+
+- What is the level of deprivation across each cluster?
+- Note that deciles for England and Wales have been allocated separately, because they have a different IMD measure.
+
+```{r deprivclus, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Deprivation decile distribution of each cluster solution.", out.width= "90%"}
+knitr::include_graphics("visuals/decile_gg.png")
+```
+
+#### Urban-rural distinction
+
+- What is the urban-rural distinction of each cluster?
+
+```{r urbanrural, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Urban-rural classification distribution of each cluster solution.", out.width= "90%"}
+knitr::include_graphics("visuals/urban_gg.png")
+```
+
 ## Discussion

### 2021-02-02 draft_paper.Rmd
change ggplot

@@
 
-#### Contribution of each cluster to change
+#### Contribution of each cluster
 
-```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Proportion of month-on-month change in crime attributable to each cluster", out.width= "90%"}
+```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Count and percentage change between months attributable to each cluster.", out.width= "90%"}
 knitr::include_graphics("visuals/change_gg.png")
@@
 
-- The above summarises the proportion of month-to-month change in crime counts attributable to each cluster.
-- In other words, we try and use this to identify which clusters drove the initial decline and subsequent resurgence nationwide.
-- For example, the red cluster (N = 23,924), despite consisting of 72% of total LSOAs, contributed to around 30% of the drop in notifiable offences between March and April.
-- While, the pink cluster (N = 83) despite comprising only 0.25% of LSOAs, contributed to 12.5% of the fall in notifiable offences between March and April.
-- July to August is weird, because although total crime increased slightly between those months, counts actually decreased for the red and green groupings, and increased for blue and purple.
-So, the blue and purple groupings drove this final nationwide increase on their own, countering the decline observed in red and green LSOAs.
-I think this is why the proportional figures do not add up to 1 for this final monthly change, so I need to think of a better way of summarizing this.
-- I haven't worked out how to visualize things yet - it needs revisiting, but hopefully gives you an idea.
+- The above summarises absolute counts and proportional month-on-month change attributable to each cluster.
+- We use this to identify which clusters drove the initial decline and subsequent resurgence nationwide.
+- Percentages represent the proportion of total nationwide month-on-month change attributable to each cluster.
+- For example, the red cluster (N = 23,924), despite consisting of 72% of total LSOAs, only contributed to 32% of the drop in notifiable offences between March and April.
+- While, the pink cluster (N = 83) despite comprising only 0.25% of LSOAs, contributed to 13% of the fall in notifiable offences between March and April.
+- This 'high crime big drop' cluster would stabilise between April and May, contributing little to the initial nationwide increase -- rather, this change is due to the other clusters increasing and 'resurging' back to normality.
+- Between July and August, red and green clusters would actually experience a drop in notifiable offences, accounting for 17% and 9% of total change respectively.
+This would partially offset the continued increase among the blue and pink clusters, which accounted for 46% and 28% of total change respectively.
+- The macro-level outcome of this July to August change was a marginal increase in crime.
+- The nationwide trend of 'decline and resurgence' masks underlying volatility: most of the initial 'lockdown crime drop' can be attributed to a disproportionately small number of areas, and the subsequent resurgence back to normality was not uniform, varying considerably at localised spatial scales.
 
@@
 
-### Deprivation
-
-- What is the level of deprivation across each cluster?
-- Note that deciles for England and Wales have been allocated separately, because they have a different IMD measure.
-
-```{r deprivclus, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Deprivation decile distribution of each cluster solution.", out.width= "90%"}
-knitr::include_graphics("visuals/decile_gg.png")
-```
-
-#### Urban-rural distinction
-
-- What is the urban-rural distinction of each cluster?
-
-```{r urbanrural, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Urban-rural classification distribution of each cluster solution.", out.width= "90%"}
-knitr::include_graphics("visuals/urban_gg.png")
-```
-
 ## Discussion

### 2021-02-02 scripts/osm_handling.R

@@ -1,0 +1,486 @@
+library(dplyr)
+library(osmdata)
+library(sf)
+library(stringr)
+library(ggplot2)
+library(tidyr)
+library(readr)
+
+# Load E&W.
+lsoa_sf <- st_read("data/ukds_infuse_lsoa/infuse_lsoa_lyr_2011_clipped.shp")
+
+# Subset LSOAs for E&W and then remove Greater Manchester.
+lsoa_ew_sf <- lsoa_sf %>% 
+  mutate(country_cd = str_extract(geo_code, "^.{1}")) %>% 
+  filter(country_cd == "E" | country_cd == "W",
+         !str_detect(geo_label, "Manchester|Bolton|Oldham|Rochdale|Stockport|Tameside|Salford|Bury|Trafford|Wigan"))
+
+# Validity checks to be done once we get to the aggregation stage.
+
+# Get bounding box of E&W.
+bb_sf <- lsoa_ew_sf %>% 
+  st_transform(crs = 4326) %>% 
+  st_bbox() 
+
+ew_bb <- c(bb_sf[[1]], bb_sf[[2]], bb_sf[[3]], bb_sf[[4]])
+
+# Pull pubs in bb
+pubs_sf <- opq(bbox = ew_bb, timeout = 100) %>%
+  add_osm_feature(key = "amenity", value = "pub") %>% 
+  osmdata_sf()
+
+# Pull points only.
+pubs_pts_sf <- pubs_sf$osm_points
+
+# We drop any pub that does not have a name allocated to it. This based on an overview of OSM
+# data, some LSOAs had upto ~400 pubs, but most were not named. If they have no name, I doubt
+# it's a legitimate to widely accepted definition of a pub. This actually what OverPass Turbo does
+# automatically (see https://overpass-turbo.eu/).
+# Doing this drops data from ~174k to ~21k which is closer to other estimates of the N pubs in E&W.
+pubs_pts_complete_sf <- pubs_pts_sf %>%
+  drop_na(name)
+
+# Remove the original query to save memory.
+rm(pubs_sf)
+
+# Check duplicates (none).
+length(unique(pubs_pts_complete_sf$geometry))
+
+# Project to BNG
+pubs_pts_complete_sf <- pubs_pts_complete_sf %>% 
+  st_transform(crs = 27700)
+
[…]

### 2021-02-02 scripts/police_data_handling.R

@@ -490,48 +490,73 @@ change_df <- tc_clusters_df %>%
   summarise(traj_crimes = sum(ew_crime_counts)) %>%
   ungroup() %>%
   group_by(month) %>%
-  mutate(total_crimes = sum(traj_crimes)) %>% 
+  mutate(total_crimes = sum(traj_crimes)) %>% # sum crimes across all trajectories (i.e. count per month in e&w)
   pivot_wider(id_cols = traj_titles, names_from = month, values_from = c(traj_crimes, total_crimes)) %>% 
+  select(traj_titles,
+         traj_crimes_march, traj_crimes_april, traj_crimes_may, 
+         traj_crimes_june, traj_crimes_july, traj_crimes_august,
+         total_crimes_march, total_crimes_april, total_crimes_may, 
+         total_crimes_june, total_crimes_july, total_crimes_august) %>% 
   mutate(tot_mar_apr  = total_crimes_april-total_crimes_march,
          tot_apr_may  = total_crimes_may-total_crimes_april,
          tot_may_jun  = total_crimes_june-total_crimes_may,
          tot_jun_jul  = total_crimes_july-total_crimes_june,
          tot_jul_aug  = total_crimes_august-total_crimes_july,
+         
          traj_mar_apr = traj_crimes_april-traj_crimes_march,
          traj_apr_may = traj_crimes_may-traj_crimes_april,
          traj_may_jun = traj_crimes_june-traj_crimes_may,
          traj_jun_jul = traj_crimes_july-traj_crimes_june,
-         traj_jul_aug = traj_crimes_august-traj_crimes_july,
-         prop_mar_apr = traj_mar_apr/tot_mar_apr,
-         prop_apr_may = traj_apr_may/tot_apr_may,
-         prop_may_jun = traj_may_jun/tot_may_jun,
-         prop_jun_jul = traj_jun_jul/tot_jun_jul,
-         prop_jul_aug = traj_jul_aug/tot_jul_aug) %>% 
-  select(traj_titles, prop_mar_apr:prop_jul_aug) %>% 
-  pivot_longer(cols = -traj_titles, names_to = "month_change", values_to = "percentage_change") %>% 
-  mutate(month_change = fct_relevel(month_change,
-                                    "prop_mar_apr",
-                                    "prop_apr_may",
-                                    "prop_may_jun",
-                                    "prop_jun_jul",
-                                    "prop_jul_aug")) 
[…]

### 2021-02-04 scripts/osm_handling.R
6 kmeans update

@@ -611,6 +611,35 @@ osm_transport_df <- lsoa_ew_valid_sf %>%
   as_tibble() %>% 
   select(-geometry, -geo_labelw, -country_cd)
 
+# conveniencercycle car parks.
+convenience_sf <- opq(bbox = ew_bb, timeout = 300) %>%
+  add_osm_feature(key = "shop", value = "convenience") %>% 
+  osmdata_sf()
+
+# Points.
+convenience_pts_sf <- convenience_sf$osm_points
+
+# Deplicates,
+length(unique(convenience_pts_sf$geometry)) 
+
+# Complete.
+convenience_pts_complete_sf <- convenience_pts_sf %>% 
+  distinct(geometry)
+
+# Project
+convenience_pts_complete_sf <- convenience_pts_complete_sf %>% 
+  st_transform(crs = 27700)
+
+# Aggregate bar points to LSOA.
+lsoa_ew_valid_sf <- lsoa_ew_valid_sf %>% 
+  mutate(conveniences = lengths(st_intersects(lsoa_ew_valid_sf, convenience_pts_complete_sf)))
+
+# subset for join
+osm_transport_df <- lsoa_ew_valid_sf %>% 
+  as_tibble() %>% 
+  select(-geometry, -geo_labelw, -country_cd)
+
+
 # Open OSM data.
 osm_df <- read_csv("data/osm_full.csv")
 
@@ -621,12 +650,12 @@ osm_full_df <- left_join(osm_df, osm_transport_df)
 write_csv(x = osm_full_df, path = "data/osm_full.csv")
 
 # Subset stats table for paper.
-osm_df <- read_csv("data/osm_stats_rounded.csv")
+osm_df <- read_csv("data/osm_stats_rounded_k6.csv")
 
 osm_df <- osm_df %>% 
-  select(traj_titles, starts_with("median"), starts_with("mean")) %>% 
-  rename(Trajectory              = traj_titles,
-         `Nightlife (median)`    = median_nightlife,
+  mutate(Cluster = LETTERS[1:6]) %>% 
+  select(Cluster, starts_with("median"), starts_with("mean")) %>% 
+  rename(`Nightlife (median)`    = median_nightlife,
          `Nightlife (mean)`      = mean_nightlife,
          `Shops (median)`        = median_shops,
          `Shops (mean)`          = mean_shops,
@@ -635,4 +664,5 @@ osm_df <- osm_df %>%
[…]

### 2021-02-04 scripts/police_data_handling.R

@@ -282,11 +282,12 @@ raw_counts_gg <- sub_data_agg_1920_df %>%
   geom_line(mapping = aes(x = month, y = ew_crime_count, group = year, colour = year), size = 0.8) +
   geom_vline(xintercept = 1.7, linetype = "dotted") +
   facet_wrap(~ crime_type, ncol = 3, scales = "free_y") +
-  scale_x_discrete(labels = c(str_extract(month.name[3:8], "^.{3}"), character(1))) +
+  scale_x_discrete(labels = str_extract(month.name[2:9], "^.{3}")) +
+  # scale_x_discrete(labels = c(str_extract(month.name[3:8], "^.{3}"), character(1))) +
   scale_color_manual(values = rev(c("black", "darkgrey"))) +
   labs(x = NULL, y = NULL, colour = NULL) +
   theme_bw() +
-  theme(axis.text.x = element_text(size = 6, hjust = -0.4),
+  theme(axis.text.x = element_text(size = 6), #, hjust = -0.4
         axis.ticks = element_line(size = 0.3, lineend = "round"),
         axis.text.y = element_text(size = 6),
         strip.text = element_text(size = 8),
@@ -323,10 +324,11 @@ gini_gg <- ggplot(data = gini_ct_1920_df) +
   facet_wrap(~ crime_type, ncol = 3) +
   ylim(0, 1) +
   labs(x = NULL, y = "Generalized Gini Coefficient", colour = NULL) +
-  scale_x_discrete(labels = c(str_extract(month.name[3:8], "^.{3}"), character(1))) +
+  # scale_x_discrete(labels = c(str_extract(month.name[3:8], "^.{3}"), character(1))) +
+  scale_x_discrete(labels = str_extract(month.name[2:9], "^.{3}")) +
   scale_color_manual(values = rev(c("black", "darkgrey"))) +
   theme_bw() +
-  theme(axis.text.x = element_text(size = 6, hjust = -0.4),
+  theme(axis.text.x = element_text(size = 6), #, hjust = -0.4
         axis.ticks = element_line(size = 0.3, lineend = "round"),
         axis.text.y = element_text(size = 6),
         strip.text = element_text(size = 8),
@@ -346,73 +348,76 @@ tc_kmeans_df <- sub_data_agg_1920_df %>%
 
 # For now, only keep the lockdown period in the first visual: March to August.
 tc_kmeans_sub_df <- tc_kmeans_df %>% 
[…]

### 2021-02-05 draft_paper.Rmd
convenience stories and begin paper

@@
 ---
-title: 'Local geography of the COVID-19 crime drop: the first six months'
+title: 'Instability and inequality in the COVID-19 crime drop'
 output: pdf_document
+bibliography: refs.bib
+csl: apa.csl
 header-includes:
@@
 knitr::opts_chunk$set(echo = FALSE)
+library(readr);library(kableExtra)
 ```
@@
 
-- An evidence-base has begun to emerge around the impact of the COVID-19 pandemic on crime and anti-social behaviour.
-- Findings have tended to demonstrate a widespread decline in police-recorded crime, although the extent to which this holds true varies by crime type.
-- But, most studies have focused on macro-level units of analysis, such as cities or countries.
-- A recent study took a 'look back' on the first six months of lockdown using police-recorded crime data aggregated to England and Wales.
-- The macro-level longitudinal patterns observed were consistent with opportunity perspectives on crime.
-- Here, we disentangle this macro-level trend using localized data aggregated to neighbourhood units across England and Wales. 
-- We focus on the extent to which localized areas remained stable, or were subject to volatility, amidst the nationwide change in police-recorded crime.
-- This includes a descriptive analysis of concentration, an identification of the 'drivers' of the lockdown crime drop, and the characteristics of localized areas driving the change.
-- In doing so, we disentangle the macro-level crime trends observed during lockdown in England and Wales.
+- An evidence-base has begun to emerge around the impact of the COVID-19 pandemic on crime.
+- Most studies have focused on macro-level units of analysis, such as cities or countries.
+- Findings have tended to demonstrate a widespread decline in police-recorded crime.
+- The longitudinal patterns observed are largely consistent with opportunity perspectives on crime.
+- As such, we are beginning to garner a reasonable evidence-base around the impact of the pandemic lockdown on crime at the macro-level.
+- That said, less is known about the localized drivers of the fall in police-recorded crime, and the degree of which areas have experienced the lockdown crime drop in unison.
+- This challenge speaks to two strands of literature.
+Firstly, crime concentration research examining localized instability during crime drops [@andresen2017trajectories] and secondly, examinations of inequality in the exposure to crime in a drop scenario [@ignatans2016whom; @akmed2021mon; @mcvie2020increasing].
+- Here, we disentangle the macro-level drop in police-recorded notifiable offences (excluding drugs) in England and Wales (see Langton et al., 2021) between February and August 2020, aggregating data to neighbourhood units (N = 33,075).
+- We do this using non-parametric longitudinal clustering, which identifies areas which remained stable, even amidst nationwide volatility, and disentangles (small number of) the areas which drove the decline and subsequent resurgence.
+- Findings are augmented with a breakdown of the facilities present in each cluster, using data sourced from Open Street Map, and the crime type profile of each cluster.
+- Results are discussed in the context of geographic inequalities and opportunity theories of crime.
 
@@
 
-- Overview of findings on lockdown crime to date.
-- Note that a key missing component relates to spatial concentrations, and the extent to which macro-level trends mask underlying, localized variation.
-- These discussions have proved vital to crime and place research, and our understanding of criminal opportunity and routine activities theory (e.g. law of crime concentration).
-- Localized analysis may also shed light on policing and resource allocation during lockdown.
-- In England and Wales, we already have a reasonable understanding of the macro-level trends.
+- Recent months have seen a wave of interest in describing and explaining changes in  police-recorded crime in response to restrictions on mobility brought about by efforts to limit the spread of COVID-19.
+- Generally speaking, findings have been consistent with theoretical expectations from an opportunity perspective, namely, that lockdown restrictions disrupted the convergence of motivated offenders, (in)capable guardians and suitable targets, bringing about widespread, macro-level declines in crime.
+- Some notable cases demonstrate that changes may be specific to crime type, although some studies using longer time periods have shown that change has been relatively uniform across crime types, with some exceptions (Langton et al., 2021).
+- Findings hold important practical and theoretical value for crime science research, and yet, less is known about the localized drivers of macro-level trends. 
+- Questions around the drivers of macro-level crime drops have been asked since the early 2000s, during which time much of the world experienced a considerably fall in crime [@weisburd2012criminology; @weisburd2004trajectories; @bannister2018local].
+- Numerous studies have deployed longitudinal clustering methods to unmask localized variation in the crime drop.
+Findings have consistently found that macro-level declines in crime over many years (decades) tend to be driven by a disproportionately small number of places, with most areas remaining fairly stable.
+- Here, we are presented with a unique scenario.
+Rather than a gradual decline in police-recorded crime over many decades, we have witnessed an overnight decline over a matter of months.
+And yet, comparable questions remain.
+- To what extent have areas experienced this decline in unison? Have the dividends of the lockdown crime drop been shared equitably across space [@akmed2021mon]?
+- Here, we use 7-months of police-recorded crime in England and Wales, aggregated to meso-level local units, to disentangle the nationwide lockdown crime drop.
+
+## Data and Method
+
+### Data
+
+- Police-recorded crime data for England and Wales (excluding Greater Manchester) between February and August, covering the first six months of lockdown March to August, and the preceding month (February), for comparison.
+- Data is available for thirteen notifiable offences, and anti-social behaviour (ASB).
+- As demonstrated in Langton et al., 2021 most notifiable offences experienced dramatic declines during April, the first full month of lockdown nationwide. 
+- Exceptions are drugs and ASB, which we exclude, and then aggregate to create a measure of 'total crime', representing all notifiable offences excluding drugs.
+- On the local level, data is aggregated to Lower Super Output Area level, a neighborhood unit designed to be uniform by resident population (~ 1500 people), of which there are around 33,000 in England and Wales.
+- Data on facilities to describe LSOAs are obtained from the Open Street Map API using the `osmdata` package [@osmrpack] in R [@rman].
+
+### Method 
+- We set the background with descriptive statistics on macro-level trends (raw counts) and concentration (generalized Gini coefficient) by crime type, including the total crime category.
+- This provides the 'global' picture of raw change and concentration at the national level, setting the reference point from which we can unmask the local variability underpinning the macro-level change.
+- For the main analysis, we deploy k-means, a non-parametric longitudinal clustering technique, on the notifiable offences measure, aggregated to LSOA in England and Wales.
+- K-means has been used in crime concentration research to disentangle local variation from macro-level trends [@andresen2017trajectories; @curman2015crime].
+- Its deployment was largely inspired by a popular technique, group-based trajectory modelling (GBTM) [@weisburd2004trajectories].
+- K-means has a number of benefits over GBTM (detailed in @curman2015crime) but in this scenario, it has a unique advantage, in that it is highly capable of identifying short-term fluctuation, unlimited by polynomial terms [@akmed2021mon].
+- In this way, it suitable for usage in this 'natural experiment' scenario, whereby we expect change to occur rapidly and dramatically from one month to the next, rather than smoothed over many years.
+- Cluster solution is determined by the Calinski-Criterion [@calinski1974dendrite].
+
+## Results
+
+### Nationwide trends
+
+- Raw counts of crime types and notifiable offences confirm what has been elsewhere and existing (more detailed) accounts of lockdown crime in England and Wales (Langton et al., 2021).
 
@@
 
-## Data and Method
+### Nationwide concentration
 
-- Generalized Gini coefficient for all offence categories.
-- Focus on two broad categories: notifiable offences and anti-social behaviour (ASB).
-- Decile change to establish whether previously high-crime/ASB areas remained, ceased or became increasingly problematic during lockdown.
-- Non-parametric longitudinal clustering (k-means) used to disentangle the macro-level trends observed during the lockdown, and identify the 'drivers' of the dramatic drop and subsequent resurgence.
-- This speaks to existing literature and methods deployed in crime and place research.
-
-## Results
-
-### Gini
+- Here, we note that most crime types experienced remarkably stable global concentration, even amidst dramatic volatility in crime counts.
+- Exceptions include bicycle theft, theft from the person, other theft, robbery, and to some extent vehicle crime.
+- These experienced a fall in concentration (i.e. crimes were less concentrated, and more spread out) compared to 2019.
+- Overall, however, notifiable offences experienced minimal change, with these crime types driving a marginal dip in concentration during the first two months of lockdown, quickly returning to normal levels of concentration and crime surged back upwards during summer.
 
@@
 
-### Deciles
-
-```{r decilenotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Decile changes in notifiable offences."}
-knitr::include_graphics("visuals/deciles_tot_an_gg.png")
-```
+- Based on these raw counts and Gini concentration trends over time, we might expect localized variability in notifiable offences. 
+- Have neighborhoods experienced these trends collectively and uniformly?
+Or have we attribute most of the macro-level trends to a handful of areas?
+To what extent have the dividends of the crime drop been shared equitably across the country [@akmed2021mon]?
+- This is what we can answer and explore using k-means to cluster the longitudinal trajectories of LSOAs.
 
@@
 ```{r kmeansnotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA notifiable offences.
-Mean (dotted line) and median (solid line) are shown for each cluster for 2020 (black) and 2019 (red) respectively.
-Violin plots refer to 2020 only.", out.width= "90%"}
-knitr::include_graphics("visuals/kmeans_violin_2019_gg.png")
+Mean (dotted line) and median (solid line) are shown for each cluster for 2020 (black), 2019 (red) and 2018 (blue) respectively.
+Distributions refer to 2020 only.", out.width= "90%"}
+knitr::include_graphics("visuals/kmeans_violin_k6_1819_gg.png")
 ```
@@
 
-```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Count and percentage change between months attributable to each cluster.", out.width= "90%"}
-knitr::include_graphics("visuals/change_gg.png")
+```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Counts and percentage of nationwide change between months attributable to each cluster.", out.width= "100%"}
+knitr::include_graphics("visuals/change_k6_gg.png")
 ```
@@
 - Maps of major city case studies.
+- Link to interactive map.
 
@@
 Descriptive statistics about the LSOAs in each cluster solution.
+
+```{r thresht, warning=F, message = F}
+osm_stats_df <- read_csv("data/osm_stats_sub_rounded_k6.csv")
+
+osm_table <- kable(osm_stats_df, caption = "Descriptive statistics of facilities in each cluster. Sourced from Open Street Map.",
+      booktabs = T, longtable = T, linesep = "", format = "latex") %>%
+    kable_styling(latex_options = c("basic"),
+                  font_size = 8, full_width = F)
+
+column_spec(osm_table, c(1:9), width = c("0.7cm", "1.5cm", "1.5cm", "1.5cm", "1.5cm", "1cm", "1cm", "1.5cm", "1cm"))
+```
 
@@
 ```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type characteristics of each cluster solution.", out.width= "90%"}
-knitr::include_graphics("visuals/traj_crimes_gg.png")
+knitr::include_graphics("visuals/traj_crimes_monthly_k6_gg.png")
 ```
@@
 ## Discussion
+
+## Appendix

### 2021-02-05 scripts/osm_handling.R

@@ -635,7 +635,7 @@ lsoa_ew_valid_sf <- lsoa_ew_valid_sf %>%
   mutate(conveniences = lengths(st_intersects(lsoa_ew_valid_sf, convenience_pts_complete_sf)))
 
 # subset for join
-osm_transport_df <- lsoa_ew_valid_sf %>% 
+osm_conv_df <- lsoa_ew_valid_sf %>% 
   as_tibble() %>% 
   select(-geometry, -geo_labelw, -country_cd)
 
@@ -644,7 +644,7 @@ osm_transport_df <- lsoa_ew_valid_sf %>%
 osm_df <- read_csv("data/osm_full.csv")
 
 # Join.
-osm_full_df <- left_join(osm_df, osm_transport_df)
+osm_full_df <- left_join(osm_df, osm_conv_df)
 
 # Save.
 write_csv(x = osm_full_df, path = "data/osm_full.csv")
@@ -652,17 +652,21 @@ write_csv(x = osm_full_df, path = "data/osm_full.csv")
 # Subset stats table for paper.
 osm_df <- read_csv("data/osm_stats_rounded_k6.csv")
 
-osm_df <- osm_df %>% 
-  mutate(Cluster = LETTERS[1:6]) %>% 
-  select(Cluster, starts_with("median"), starts_with("mean")) %>% 
-  rename(`Nightlife (median)`    = median_nightlife,
+names_df <- distinct(traj_names_osm_df, traj, traj_titles)
+
+osm_table_df <- osm_df %>% 
+  left_join(names_df) %>% 
+  select(traj, starts_with("median"), starts_with("mean")) %>% 
+  rename(Cluster                 = traj,
+         `Nightlife (median)`    = median_nightlife,
          `Nightlife (mean)`      = mean_nightlife,
          `Shops (median)`        = median_shops,
          `Shops (mean)`          = mean_shops,
          `Public Transport (median)` = median_trans,
          `Public Transport (mean)`   = mean_trans,
          `Bicycle parking (median)`  = median_bikes,
-         `Bicycle parking (mean)`    = mean_bikes)
+         `Bicycle parking (mean)`    = mean_bikes) %>% 
+  arrange(Cluster)
 
 
-write_csv(x = osm_df, path = "data/osm_stats_sub_rounded_k6.csv")
+write_csv(x = osm_table_df, path = "data/osm_stats_sub_rounded_k6.csv")

### 2021-02-05 scripts/police_data_handling.R

@@ -283,7 +283,6 @@ raw_counts_gg <- sub_data_agg_1920_df %>%
   geom_vline(xintercept = 1.7, linetype = "dotted") +
   facet_wrap(~ crime_type, ncol = 3, scales = "free_y") +
   scale_x_discrete(labels = str_extract(month.name[2:9], "^.{3}")) +
-  # scale_x_discrete(labels = c(str_extract(month.name[3:8], "^.{3}"), character(1))) +
   scale_color_manual(values = rev(c("black", "darkgrey"))) +
   labs(x = NULL, y = NULL, colour = NULL) +
   theme_bw() +
@@ -324,7 +323,6 @@ gini_gg <- ggplot(data = gini_ct_1920_df) +
   facet_wrap(~ crime_type, ncol = 3) +
   ylim(0, 1) +
   labs(x = NULL, y = "Generalized Gini Coefficient", colour = NULL) +
-  # scale_x_discrete(labels = c(str_extract(month.name[3:8], "^.{3}"), character(1))) +
   scale_x_discrete(labels = str_extract(month.name[2:9], "^.{3}")) +
   scale_color_manual(values = rev(c("black", "darkgrey"))) +
   theme_bw() +
@@ -367,7 +365,7 @@ length(unique(tc_kmeans_sub_clean_df$lsoa_code))
 
 
 # Perform kmeans
-n <- 3:6
+n <- 3:6 # max 6 for simplicity.
 
 tc_kml_mat <- as.matrix(tc_kmeans_sub_clean_df[2:8])
 tc_traj    <- clusterLongData(traj = tc_kml_mat) 
@@ -375,7 +373,7 @@ kml(tc_traj, nbClusters = n, toPlot = "criterion", nbRedrawing = 20)
 
 # Apppend clusters back with data frame.
 tc_clusters <- cbind.data.frame(lsoa_code   = tc_kmeans_sub_clean_df$lsoa_code,
-                                 traj       = getClusters(tc_traj, 6)) # We select 4 due to CH value.
+                                 traj       = getClusters(tc_traj, 6)) # We select 6 due to CH value.
 
 tc_clusters_df <- tc_kmeans_sub_clean_df %>% 
   left_join(tc_clusters) %>% 
@@ -514,8 +512,6 @@ tc_clusters_2018_df <- sub_ew_agg_2018 %>%
                                 july     = "2018-07",
                                 august   = "2018-08"))
 
-
-
 # Violin plot with 2018 added.
[…]

### 2021-02-08 draft_paper.Rmd
fig 3 and write results

@@
 - As demonstrated in Langton et al., 2021 most notifiable offences experienced dramatic declines during April, the first full month of lockdown nationwide. 
-- Exceptions are drugs and ASB, which we exclude, and then aggregate to create a measure of 'total crime', representing all notifiable offences excluding drugs.
-- On the local level, data is aggregated to Lower Super Output Area level, a neighborhood unit designed to be uniform by resident population (~ 1500 people), of which there are around 33,000 in England and Wales.
-- Data on facilities to describe LSOAs are obtained from the Open Street Map API using the `osmdata` package [@osmrpack] in R [@rman].
+- Exceptions are drugs and ASB.
+- Here, we provide a background of nationwide trends by crime type (counts, Gini) before proceeding with a broader 'notifiable offences (excluding drugs)' category.
+- Data is aggregated to Lower Super Output Area level, a neighborhood unit designed to be uniform by resident population (~ 1500 people), of which there are around ~33,000 in England and Wales.
+- Due to some LSOA not having snap points within them, we end up with a final population size of N = 33,075.
+- Data on facilities to describe the LSOA clusters obtained are retrieved from the Open Street Map API using the `osmdata` package [@osmrpack] in R [@rman].
 
@@
 - This provides the 'global' picture of raw change and concentration at the national level, setting the reference point from which we can unmask the local variability underpinning the macro-level change.
-- For the main analysis, we deploy k-means, a non-parametric longitudinal clustering technique, on the notifiable offences measure, aggregated to LSOA in England and Wales.
+- For the main analysis, we deploy k-means [@genolini2011kml; @genolini2015kml], a non-parametric longitudinal clustering technique, on notifiable offences counts trajectories at LSOA level.
 - K-means has been used in crime concentration research to disentangle local variation from macro-level trends [@andresen2017trajectories; @curman2015crime].
@@
 - In this way, it suitable for usage in this 'natural experiment' scenario, whereby we expect change to occur rapidly and dramatically from one month to the next, rather than smoothed over many years.
-- Cluster solution is determined by the Calinski-Criterion [@calinski1974dendrite].
+- Cluster solution is determined by the Calinski-Criterion [@calinski1974dendrite] in alignment with existing studies in crime concentration research [@@andresen2017trajectories; @akmed2021mon].
 
@@
 
-- Raw counts of crime types and notifiable offences confirm what has been elsewhere and existing (more detailed) accounts of lockdown crime in England and Wales (Langton et al., 2021).
+- Raw counts by crime type confirm what has been demonstrated in existing accounts of lockdown crime in England and Wales (Langton et al., 2021).
+- Here, we augment these existing crimes with a 'total crime (excluding drugs)' category, representing all notifiable offences with the exception of drugs.
+- Drug crimes are thought to largely reflect police proactivity rather than a genuine shift in criminal behavioral.
+-  We accept that granularity is lost with examining notifiable offences in their totality. We address this using a crime type breakdown, and return to it in the more detail for the discussion.
 
@@
 
-- Here, we note that most crime types experienced remarkably stable global concentration, even amidst dramatic volatility in crime counts.
+- We note from the Gini coefficient that most crime types experienced remarkably stable 'global' concentration, even amidst dramatic volatility in crime counts.
 - Exceptions include bicycle theft, theft from the person, other theft, robbery, and to some extent vehicle crime.
-- These experienced a fall in concentration (i.e. crimes were less concentrated, and more spread out) compared to 2019.
-- Overall, however, notifiable offences experienced minimal change, with these crime types driving a marginal dip in concentration during the first two months of lockdown, quickly returning to normal levels of concentration and crime surged back upwards during summer.
+- These experienced a fall in concentration (i.e. crimes were less concentrated, more spread out) compared to 2019.
+- Overall, however, notifiable offences (excluding drugs) experienced minimal change.
+There is a marginal dip in concentration during the first two months of lockdown, quickly returning to normal levels of concentration as crime surged back upwards during summer.
 
@@
 To what extent have the dividends of the crime drop been shared equitably across the country [@akmed2021mon]?
-- This is what we can answer and explore using k-means to cluster the longitudinal trajectories of LSOAs.
+- What is the crime type profile of the area driving nationwide trends?
+- We can answer these questions using k-means to cluster the longitudinal trajectories of LSOA.
 
@@
 
-- General conclusion: most of the lockdown crime drop can be attributed to a small number LSOAs.
-- Most localized areas actually experienced minimal crime change during the pandemic.
-- The proportion of total crime attributable to very low and very high crime clusters fell into April, while mid-crime areas increased their proportion.
-- All clusters converged back to 'normality' by the end of August.
-- Remarkable stability given the absolute change that occurred during this study period.
-- 2020 mean (dotted line) and median (solid line) of each cluster are shown in black.
-- 2019 mean (dotted line) and median (solid line) of each cluster are shown in red.
-- This demonstrates that the clusters identified using the 2020 data were distinct and meaningful even when also applied to the 2019 data.
+- Figure 3 visualizes the cluster solution generated by k-means, with 6-clusters identified as optimal based on the Calinski-Criterion [@calinski1974dendrite].
+- Black solid line is the median count of the cluster at any given time point, with the black dotted line showing the mean.
+- Violin plots have been added to convey the variation in counts at each time point.
+Here, we note that distributions were fairly uniform, with the mean and median representing the cluster solutions fairly well, and little overlap between clusters.
+- For reference to what we trends we may have expected in a 'typical' year without the pandemic, we have added the mean and median trend lines for each cluster during 2018 (in blue) and 2019 (in red).
+- This demonstrates that the clusters identified using the 2020 data were distinct and meaningful even in previous years, and highlights the gap between previous years and what has been observed during the pandemic.
+- Overall, we note that most LSOA were remarkably stable during the pandemic.
+Clusters A and B can be described and 'low crime and stable', exhibiting fairly low counts throughout the study period and across years.
+Together, these clusters comprise 89% of LSOAs in England and Wales.
+- That said, there is some deviation from previous years in April, in alignment with expectations, there is a slight dip in the mean and median crime count, particularly in Cluster B.
+- The third largest cluster, cluster C, comprises 10% of all LSOAs in the country.
+This cluster experienced a more prominent dip in crime, along with cluster D (1.6% of LSOAs).
+Together, we might describe these LSOAs as 'mid-crime, mid-drop'.
+- However, the 'action' appears to be occurring amongst a small subset of LSOAs, contained within clusters E and F.
+Together, these clusters comprise only 0.33% of all LSOA in the country (N = 110), however they are 'high crime, major drop'.
+- Interestingly, clusters E and F represent the only clusters which did not return back to normality by August, even we assume that 2018 and 2019 represent a trend which we might have otherwise expected.
+- Overall, we note that despite the dramatic nationwide trend in notifiable offences (excluding drugs), there has been remarkable stability in local areas, with most change being driven by a small number of areas.
 
@@
 
-- The above summarises absolute counts and proportional month-on-month change attributable to each cluster.
+- To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the monthly change in counts of each cluster.
+- Alongside these raw counts, we also include the percentage of total absolute change (i.e. positive or negative) nationwide attributable to each cluster.
 - We use this to identify which clusters drove the initial decline and subsequent resurgence nationwide.
-- Percentages represent the proportion of total nationwide month-on-month change attributable to each cluster.
-- For example, the red cluster (N = 23,924), despite consisting of 72% of total LSOAs, only contributed to 32% of the drop in notifiable offences between March and April.
-- While, the pink cluster (N = 83) despite comprising only 0.25% of LSOAs, contributed to 13% of the fall in notifiable offences between March and April.
-- This 'high crime big drop' cluster would stabilise between April and May, contributing little to the initial nationwide increase -- rather, this change is due to the other clusters increasing and 'resurging' back to normality.
-- Between July and August, red and green clusters would actually experience a drop in notifiable offences, accounting for 17% and 9% of total change respectively.
-This would partially offset the continued increase among the blue and pink clusters, which accounted for 46% and 28% of total change respectively.
-- The macro-level outcome of this July to August change was a marginal increase in crime.
-- The nationwide trend of 'decline and resurgence' masks underlying volatility: most of the initial 'lockdown crime drop' can be attributed to a disproportionately small number of areas, and the subsequent resurgence back to normality was not uniform, varying considerably at localised spatial scales.
+- As expected, the vast majority of change occurred between March and April.
+- However, Figure 4 demonstrates that this substantial decline did not occur equitably across the country.
+Consistent with the cluster solutions trends in Figure 3, it is clear that a small number of areas drove the trend.
+- For instance, clusters A and B, which comprise 89% of LSOAs in the country, accounted for only 47% of the decline between March and April.
+- By contrast, clusters E and F, which together comprised only 0.33% of LSOA (N = 110) contributed to 15% of the nationwide drop between months.
+This a disproportionately large impact of the macro-level trend, and further supports the observation that most areas remained fairly stable.
+- Interestingly, this picture shifts upon the beginning of the resurgence between April and May.
+- Cluster A continues to contribute a disproportionately small amount to nationwide change, but cluster B contributes to 42% of the increase, despite only comprising 31% of LSOA.
+- The contribution of the smallest clusters E and F is remarkably small compared to the initial decline, 2% and 0% respectively.
+This can be largely attributed to the flattening of their crime count trends between these months.
+In other words, following their initially steep decline between March and April, crime remained reasonably unchanged between April and May in these areas.
+- Overall, this pattern continues throughout the spring convergence back to normality, with some notable changes. 
+- Following the period of stabilization, clusters E and F gather pace during the resurgence, as does cluster D.
+For instance, between June and July, these three clusters account for nearly a third of the nationwide increase, despite comprising less than 2% of total LSOA.
+- July to August represents an interesting period.
+Until that point, each cluster, despite their vastly differing relative contributions to change, with directionally homogeneous.
+In other words, every cluster declines between February and April, and then increased between April and July.
+- Between July and August, clusters began to diverge. Nationwide, there was a marginal increase in notifiable offence counts (see Figure 1), but this masks a great deal of localized variability.
+- The two large, low crime clusters (A and B) actually declined again between July and August, while the higher crime clusters continued to increase.
+This is suggestive of a local change in opportunities as lockdown rules were relaxed - change which has been aggregated away at the macro-level.
+- In other words, low crime areas mitigated against further nationwide increases which may, if they had continued, actually resulted in a higher crime count in August than observed in 2019.
 

### 2021-02-08 scripts/police_data_handling.R
writingggg

@@ -573,8 +573,8 @@ ggsave(plot = kmeans_violin_1819_gg, filename = "visuals/kmeans_violin_k6_1819_g
 #        height = 20, width = 20, unit = "cm", dpi = 200)
 
 # Save and load workspace as appropriate.
-save.image(file = "data_handling_6kmean.RData")
-# load(file = "data_handling_6kmean.RData")
+# save.image(file = "data_handling_6kmean.RData")
+load(file = "data_handling_6kmean.RData")
 
 
 # Create proportion contribution to total crime in each month.
@@ -771,7 +771,37 @@ traj_names_df <- tc_clusters_df %>%
 # Join OSM data.
 traj_names_osm_df <- left_join(traj_names_df, osm_df, by = c("lsoa_code" = "geo_code"))
 
-# Descriptives.
+# Overall descriptives.
+
+osm_total_stats_df <- traj_names_osm_df %>%
+  summarise(Median_nightlife = median(nightlife_total),
+            Mean_nightlife   = mean(nightlife_total),
+            SD_nightlife     = sd(nightlife_total),
+            Min_nightlife    = min(nightlife_total),
+            Max_nightlife    = max(nightlife_total),
+            Median_shops     = median(shops_total),
+            Mean_shops       = mean(shops_total),
+            SD_shops         = sd(shops_total),
+            Min_shops        = min(shops_total),
+            Max_shops        = max(shops_total),
+            Median_transport     = median(transport_total),
+            Mean_transport       = mean(transport_total),
+            SD_transport         = sd(transport_total),
+            Min_transport        = min(transport_total),
+            Max_transport        = max(transport_total),
+            Median_bikes     = median(bikes),
+            Mean_bikes       = mean(bikes),
+            SD_bikes         = sd(bikes),
+            Min_bikes        = min(bikes),
+            Max_bikes        = max(bikes)) %>% 
+  mutate_if(is.numeric, round, 2) %>%
+  mutate(id = 1) %>% 
+  pivot_longer(cols = -id, names_to = "Facility", values_to = "value") %>% 
+  separate(col = Facility, sep = "_", into = c("statistic", "Facility")) %>% 
+  pivot_wider(id_cols = Facility, names_from = statistic, values_from = value) %>% 
+  mutate_if(is.character, str_to_title)
+
+# Cluster descriptives.
 osm_stats_df <- traj_names_osm_df %>%
   group_by(traj_titles) %>% 
   summarise(median_nightlife = median(nightlife_total),

### 2021-02-09 draft_paper.Rmd
bullet point draft 1

@@
 - Here, we disentangle the macro-level drop in police-recorded notifiable offences (excluding drugs) in England and Wales (see Langton et al., 2021) between February and August 2020, aggregating data to neighbourhood units (N = 33,075).
-- We do this using non-parametric longitudinal clustering, which identifies areas which remained stable, even amidst nationwide volatility, and disentangles (small number of) the areas which drove the decline and subsequent resurgence.
+- We do this using non-parametric longitudinal clustering, which identifies areas which remained stable, even amidst nationwide volatility, and disentangles the areas which drove the initial decline and subsequent resurgence.
 - Findings are augmented with a breakdown of the facilities present in each cluster, using data sourced from Open Street Map, and the crime type profile of each cluster.
@@
 - Findings hold important practical and theoretical value for crime science research, and yet, less is known about the localized drivers of macro-level trends. 
-- Questions around the drivers of macro-level crime drops have been asked since the early 2000s, during which time much of the world experienced a considerably fall in crime [@weisburd2012criminology; @weisburd2004trajectories; @bannister2018local].
-- Numerous studies have deployed longitudinal clustering methods to unmask localized variation in the crime drop.
+- Questions around the drivers of macro-level crime drops have been of interest for some time, following the widespread fall in crime since the 1990s [@farrell2014crime; @weisburd2012criminology; @weisburd2004trajectories; @bannister2018local].
+- At the city level, numerous studies have deployed longitudinal clustering methods to unmask localized variation in the crime drop.
 Findings have consistently found that macro-level declines in crime over many years (decades) tend to be driven by a disproportionately small number of places, with most areas remaining fairly stable.
 - Here, we are presented with a unique scenario.
-Rather than a gradual decline in police-recorded crime over many decades, we have witnessed an overnight decline over a matter of months.
+Rather than a gradual decline in police-recorded crime over many decades, we have witnessed a rapid decline over a matter of weeks and months.
 And yet, comparable questions remain.
@@
 - Here, we use 7-months of police-recorded crime in England and Wales, aggregated to meso-level local units, to disentangle the nationwide lockdown crime drop.
+- We describe areas in terms of their opportunity structure and crime type profile.
 
@@
 
-### Data
+### Crime data
 
@@
 - Exceptions are drugs and ASB.
-- Here, we provide a background of nationwide trends by crime type (counts, Gini) before proceeding with a broader 'notifiable offences (excluding drugs)' category.
-- Data is aggregated to Lower Super Output Area level, a neighborhood unit designed to be uniform by resident population (~ 1500 people), of which there are around ~33,000 in England and Wales.
-- Due to some LSOA not having snap points within them, we end up with a final population size of N = 33,075.
+- We provide context and a summary of nationwide trends decomposed by crime type in terms of raw trends (counts) and concentration (Generalized Gini coefficient).
+- We then proceed with a broader 'notifiable offences (excluding drugs)' category.
+
+### Unit of analysis
+
+- Data is aggregated to Lower Super Output Area level, a neighborhood unit designed to be uniform by resident population (~ 1500 people), of which there are around 33,000 in England and Wales.
+- Due to some LSOA not having police.uk snap points, and one LSOA not having any police-recorded notifiable offences during the study period, we end up with a final sample/population size of N = 33,075.
+
+### Open Street Map
+
 - Data on facilities to describe the LSOA clusters obtained are retrieved from the Open Street Map API using the `osmdata` package [@osmrpack] in R [@rman].
+- Four count measures were chosen based on established crime attractors and generators, and availability and feasibility of obtaining nationwide data from Open Street Map, namely: nightlife, shops, public transport and bicycle parking.
+- Measure simply represent node counts.
+- Nightlife: pubs, nightclubs, restaurants.
+- Shops: convenience stores, malls, shoe shops, department stores, clothes shops, electrical shops, supermarkets, chemists, greengrocers.
+- Public transport: bus stops and railway stations.
+- Bicycle parking: bicycle parking.
+- Descriptive statistics on these counts at LSOA level are reported for each cluster solution.
+- To supplement these statistics, which as discussed later, are imperfect, we provide access to an interactive map for readers to explore independently.
 
@@
 - Its deployment was largely inspired by a popular technique, group-based trajectory modelling (GBTM) [@weisburd2004trajectories].
-- K-means has a number of benefits over GBTM (detailed in @curman2015crime) but in this scenario, it has a unique advantage, in that it is highly capable of identifying short-term fluctuation, unlimited by polynomial terms [@akmed2021mon].
-- In this way, it suitable for usage in this 'natural experiment' scenario, whereby we expect change to occur rapidly and dramatically from one month to the next, rather than smoothed over many years.
-- Cluster solution is determined by the Calinski-Criterion [@calinski1974dendrite] in alignment with existing studies in crime concentration research [@@andresen2017trajectories; @akmed2021mon].
+- K-means has a number of benefits over GBTM for crime and place research (detailed in @curman2015crime) but in this scenario, it has a unique advantage, in that it is highly capable of identifying short-term fluctuation, unlimited by polynomial terms [@akmed2021mon].
+- In this way, it is suitable for usage in this 'natural experiment' scenario, whereby we expect change to occur rapidly and dramatically from one month to the next, rather than smoothed over many years.
+- Cluster solution is determined by the Calinski-Criterion [@calinski1974dendrite] in alignment with existing studies in crime concentration research [@andresen2017trajectories; @akmed2021mon].
 
@@
 - Drug crimes are thought to largely reflect police proactivity rather than a genuine shift in criminal behavioral.
--  We accept that granularity is lost with examining notifiable offences in their totality. We address this using a crime type breakdown, and return to it in the more detail for the discussion.
+-  We accept that granularity is lost with examining notifiable offences in their totality. We address this using a crime type breakdown, and return to it in more detail for the discussion.
 
@@
 
-- We note from the Gini coefficient that most crime types experienced remarkably stable 'global' concentration, even amidst dramatic volatility in crime counts.
+- We note from the Generalized Gini coefficients that most crime types experienced remarkably stable 'global' concentration, even amidst this dramatic volatility in crime counts.
 - Exceptions include bicycle theft, theft from the person, other theft, robbery, and to some extent vehicle crime.
@@
 - Overall, however, notifiable offences (excluding drugs) experienced minimal change.
-There is a marginal dip in concentration during the first two months of lockdown, quickly returning to normal levels of concentration as crime surged back upwards during summer.
+There is a marginal dip in concentration during the first month of lockdown, quickly returning to normal levels of concentration as crime surged back upwards during summer.
 
@@
 - Have neighborhoods experienced these trends collectively and uniformly?
-Or have we attribute most of the macro-level trends to a handful of areas?
+Can we attribute most of the macro-level count trends to a handful of areas?
 To what extent have the dividends of the crime drop been shared equitably across the country [@akmed2021mon]?
-- What is the crime type profile of the area driving nationwide trends?
-- We can answer these questions using k-means to cluster the longitudinal trajectories of LSOA.
+- What is the oppportunity structure of the areas driving nationwide trends?
+- What is the crime type profile of the areas driving nationwide trends?
 
@@
 - Violin plots have been added to convey the variation in counts at each time point.
-Here, we note that distributions were fairly uniform, with the mean and median representing the cluster solutions fairly well, and little overlap between clusters.
-- For reference to what we trends we may have expected in a 'typical' year without the pandemic, we have added the mean and median trend lines for each cluster during 2018 (in blue) and 2019 (in red).
+Here, we note that distributions were fairly uniform, with the mean and median representing the cluster solutions fairly well.
+There is little overlap between clusters.
+- For reference to what trends we may have expected in a 'typical' year without the pandemic, we have added the mean and median trend lines for each cluster during 2018 (in blue) and 2019 (in red).
 - This demonstrates that the clusters identified using the 2020 data were distinct and meaningful even in previous years, and highlights the gap between previous years and what has been observed during the pandemic.
@@
 Together, these clusters comprise 89% of LSOAs in England and Wales.
-- That said, there is some deviation from previous years in April, in alignment with expectations, there is a slight dip in the mean and median crime count, particularly in Cluster B.
+- That said, there is some deviation from previous years in April, in alignment with expectations. There is a slight dip in the mean and median crime count, particularly in Cluster B.
 - The third largest cluster, cluster C, comprises 10% of all LSOAs in the country.
@@
 Together, these clusters comprise only 0.33% of all LSOA in the country (N = 110), however they are 'high crime, major drop'.
-- Interestingly, clusters E and F represent the only clusters which did not return back to normality by August, even we assume that 2018 and 2019 represent a trend which we might have otherwise expected.
+- Interestingly, clusters E and F represent the only clusters which did not return back to normality by August (if we assume that 2018 and 2019 represent a trend which we might have otherwise expected).
 - Overall, we note that despite the dramatic nationwide trend in notifiable offences (excluding drugs), there has been remarkable stability in local areas, with most change being driven by a small number of areas.
@@
 
-```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Counts and percentage of nationwide change between months attributable to each cluster.", out.width= "100%"}
-knitr::include_graphics("visuals/change_k6_gg.png")
-```
-
 - To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the monthly change in counts of each cluster.
-- Alongside these raw counts, we also include the percentage of total absolute change (i.e. positive or negative) nationwide attributable to each cluster.
+- Alongside these raw counts, we also include the percentage of total absolute change (i.e. positive or negative) attributable to each cluster.
 - We use this to identify which clusters drove the initial decline and subsequent resurgence nationwide.
@@
 - July to August represents an interesting period.
-Until that point, each cluster, despite their vastly differing relative contributions to change, with directionally homogeneous.
-In other words, every cluster declines between February and April, and then increased between April and July.
+Until that point, each cluster, despite their vastly differing relative contributions to change, were directionally homogeneous.
+In other words, every cluster declined between February and April, and then increased between April and July.
 - Between July and August, clusters began to diverge. Nationwide, there was a marginal increase in notifiable offence counts (see Figure 1), but this masks a great deal of localized variability.
@@
 
+```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Counts and percentage of nationwide change between months attributable to each cluster.", out.width= "100%"}
+knitr::include_graphics("visuals/change_k6_gg.png")
+```
+
 ### Spatial distribution of clusters
@@
 - List major city composition?
-- Maps of major city case studies.
-- Link to interactive map.
+- Maps of major city case studies?
+- Link to interactive map?
 
@@
 
-Descriptive statistics about the LSOAs in each cluster solution.
+- We have now identified that, amidst dramatic nationwide change in notifiable offences during lockdown, and stability in overall crime concentrations, there has been considerable underlying volatility, with a small number of local areas driving the macro-level trends.
+- From an opportunity perspective, we expect the clusters to have differing opportunity structures.
+- For instance, LSOA which were typically 'high crime' in 2018 and 2019, and then experienced a dramatic fall in crime upon lockdown commencement, are likely to have a large volume of criminal opportunities which were closed off following restrictions on mobility.
+- On the other hand, areas which were typically low crime, and remained low crime, are likely to have a vastly different opportunity structure: places with minimal attractors or generators to begin with are unlikely to have changed dramatically during lockdown.
+- Table 1 demonstrates this point using the data obtained from Open Street Map.
+- Clusters A and B, the 'low crime and stable' LSOAs, tend to have few major opportunities for crime.
+Few pubs, nightclubs or restaurants to generate a violence and theft-prone nightlife, minimal shopping opportunities (and thus less chance of shoplifting), limited public transport (less theft) and less bicycle parking (less bicycle theft).
+- By contrast, the 'high crime major drop' clusters (e.g.
+E and F) are major nightlife areas, have considerable shopping opportunities, often have expansive public transport networks, and have many bicycle parking facilities.
+- These findings are consistent with what is observed using the interactive map made available as part of this paper (link here).
 
@@
 
-- What crime types tend to comprise each cluster?
+- For the purposes of simplicity and setting a benchmark, we have been using a measure of 'notifiable offences excluding drugs' for analysis.
+- That said, we might expect the clusters identified to have different crime type profiles.
+- Given the differing opportunity structures between clusters, the 'typical' crime profiles of each cluster may change considerably as lockdown commences.
+- Alternatively, the pervasiveness of lockdown on mobility may have impacted on crime types uniformly within each cluster, in which case, we may see stability in crime type profiles before and after lockdown.
+- Figure 5 summarises the profile of each cluster by calculating the proportion of total crime in each cluster.
+- We observe considerable stability in the distribution of crime types both within and between clusters.
+- Violence and sexual offences represent the most prevalent crime type across all clusters.
+The proportion of total crime in each cluster attributable to violence and sexual offences increases in the first two months of lockdown.
+- This by no means implies that violence and sexual offences increased, but rather, that violence and sexual offences declined at a slower rate than other crime types, increasing its share of all crimes.
+- In cluster F, we see a unique pattern, whereby theft from the person and other theft are squeezed considerably into April and May, while the share of public order increased.
+- This might reflect the opportunity structures of LSOA in cluster F being weighted heavily towards theft (e.g. lots of public transport nodes).
+- At the same time, the increase share in public order may reflect other activity, such as the protests around BLM and anti-lockdown movements, which take place in city centres for coverage.
 
@@
 
+- We have investigated the degree of stability and localised inequality in the lockdown crime drop between February and August in England and Wales using police-recorded crime data aggregated to local areas (LSOA).
+- Findings are largely consistent with broader findings around inequalities in the crime drop and the notion that a small number of fine-grained geographic areas tend to drive nationwide trends [@andresen2017trajectories; @weisburd2012criminology; @mcvie2020increasing].
+- Specifically, we find that most LSOA remained remarkably stable during the initial, major lockdown crime drop between March and April.
+- The subsequent resurgence among low and mid-crime areas was limited by 'high crime' areas stabilizing into May.
+- But, the resurgence of high crime areas was lagged, and began towards the end of spring.
+- Further macro-level increases in crime at the end of summer were limited by the low crime and stable areas, mitigating the impact of increases in high crime clusters.
+- The LSOA which drove the lockdown crime drop tended to have considerable crime attractors and generators, such as public transport nodes, an active nightlife, and plenty of shops and bicycle parking.
+- Thus, while the opportunity structures of these areas makes them high crime areas in typical times, in a lockdown scenario, these are the areas which are most impacted by changes in mobility, driving crime down.
+- The crime type profile of clusters demonstrates considerable stability.
+
 ## Appendix

### 2021-02-23 scripts/police_data_handling_cleaned.R
intro draft and data start

@@ -609,6 +609,9 @@ lsoa_ew_valid_sf <- sub_data_agg_full_df %>%
   right_join(lsoa_ew_valid_sf, by = c("lsoa_code" = "geo_code")) %>% 
   st_as_sf()
 
+# Save.
+st_write(obj = lsoa_ew_valid_sf, dsn = "data/lsoa_ew_valid_k6_sf.shp")
+
 # Selection of cities. Top 5 by population in England minus London plus Cardiff.
 birm_sf <- lsoa_ew_valid_sf %>% 
   filter(str_detect(string = lsoa_ew_valid_sf$geo_label, pattern = "Birmingham"))
@@ -638,11 +641,13 @@ cardiff_sf <- cardiff_sf %>%
 cities_list <- list(birm_sf, leeds_sf, sheff_sf, brad_sf, liver_sf, cardiff_sf)
 
 # Create map plot function.
+# Get colours manually, because not all cities have all clusters.
 map_fun <- function(x) {
   ggplot(data = x) +
     geom_sf(mapping = aes(fill = traj_titles), size = 0.0001, colour = "black", alpha = 0.8) +
     theme_void() +
-    theme(legend.position = "none")
+    theme(legend.position = "none") +
+    scale_fill_manual(values = scales::hue_pal()(6), drop = FALSE)
 }
 
 # Create plots.
@@ -653,15 +658,16 @@ maps_gg <- plot_grid(plotlist = maps_list, ncol = 2, labels = c("Birmingham", "L
                                                                 "Bradford", "Liverpool", "Cardiff"),
                      label_size = 7, scale = c(1, 1, 1.1, 1.05, 1, 1.1))
 
-# Create legend (Cardiff just used as example - any city will work).
-temp_gg <- cardiff_sf %>% 
+# Create legend (Birmingham needed as it has all 6 clusters).
+temp_gg <- birm_sf %>% 
   mutate(traj_titles = str_extract(traj_titles, "^.{3}")) %>% 
   ggplot() +
   geom_sf(mapping = aes(fill = traj_titles)) +
   theme_void() +
   labs(fill = NULL) +
   theme(legend.position = "bottom") +
-  guides(fill = guide_legend(nrow = 1)) 
+  guides(fill = guide_legend(nrow = 1)) +
+  scale_fill_manual(values = scales::hue_pal()(6), drop = FALSE)
   
 legend_gg <- get_legend(temp_gg)
 

### 2021-02-25 scripts/police_data_handling_cleaned.R
method writing and cluster legend add

@@ -255,10 +255,6 @@ length(unique(sub_data_agg_2020$lsoa_code))  # 33076
 length(unique(sub_data_agg_2019$lsoa_code))  # 33076
 length(unique(sub_data_agg_2018$lsoa_code))  # 33076
 
-# Save workspace.
-# save.image(file = "data_handling_cleaning.RData")
-load(file = "data_handling_cleaning.RData")
-
 # Combine 2018, 2019 and 2020 datasets for next steps.
 sub_data_agg_full_df <- bind_rows(sub_data_agg_2018, sub_data_agg_2019, sub_data_agg_2020)
 
@@ -273,7 +269,7 @@ total_crime_agg_df <- sub_data_agg_full_df %>%
   group_by(month, year) %>% 
   summarise(ew_crime_count = sum(crime_count)) %>% 
   ungroup() %>% 
-  mutate(crime_type = "Total crime (excl. drugs)") %>% 
+  mutate(crime_type = "Notifiable offences (excl. drugs)") %>% 
   select(crime_type, year, month, ew_crime_count) %>% 
   separate(month, into = c("year", "month"), sep = "-") %>% 
   filter(month != "01", month != "09", month != "10", month != "11", month != "12") 
@@ -312,7 +308,7 @@ gini_total_crime_df <- sub_data_agg_full_df %>%
   summarise(gini_coef = gini(crime_count, generalized = TRUE, unbiased = TRUE)) %>% 
   ungroup() %>% 
   separate(month, into = c("year","month"), sep = "-") %>% 
-  mutate(crime_type = "Total crime (excl. drugs)") %>% 
+  mutate(crime_type = "Notifiable offences (excl. drugs)") %>% 
   select(crime_type, gini_coef, year,  month)
 
 # Do the same but by crime type and add the total crime category on.
@@ -344,6 +340,25 @@ gini_gg <- ggplot(data = gini_crime_type_df) +
 
 # Save.
 ggsave(plot = gini_gg, filename = "visuals/gini_gg.png", width = 14, height = 20, units = "cm", dpi = 600)
+
+# Long-term trends for public order.
+long_term_gg <- sub_data_agg_full_df %>% 
+  filter(crime_type == "Public order" | crime_type == "Violence and sexual offences" |
+         crime_type == "Other crime",
+         month != "2020-09" & month != "2020-10" & month != "2020-11") %>% 
+  group_by(crime_type, month, year) %>% 
+  summarise(ew_crime_count = sum(crime_count)) %>% 
+  ungroup() %>% 
[…]

### 2021-02-25 draft_paper.Rmd
cluster legend

@@
 ---
-title: 'Instability and inequality in the COVID-19 crime drop'
+title: 'Instability and inequality in the COVID-19 lockdown crime drop'
 output: pdf_document
@@
 
-## Introduction
+# Introduction
 
-- An evidence-base has begun to emerge around the impact of the COVID-19 pandemic on crime.
-- Most studies have focused on macro-level units of analysis, such as cities or countries.
-- Findings have tended to demonstrate a widespread decline in police-recorded crime.
-- The longitudinal patterns observed are largely consistent with opportunity perspectives on crime.
-- As such, we are beginning to garner a reasonable evidence-base around the impact of the pandemic lockdown on crime at the macro-level.
-- That said, less is known about the localized drivers of the fall in police-recorded crime, and the degree of which areas have experienced the lockdown crime drop in unison.
-- This challenge speaks to two strands of literature.
-Firstly, crime concentration research examining localized instability during crime drops [@andresen2017trajectories] and secondly, examinations of inequality in the exposure to crime in a drop scenario [@ignatans2016whom; @akmed2021mon; @mcvie2020increasing].
-- Here, we disentangle the macro-level drop in police-recorded notifiable offences (excluding drugs) in England and Wales (see Langton et al., 2021) between February and August 2020, aggregating data to neighbourhood units (N = 33,075).
-- We do this using non-parametric longitudinal clustering, which identifies areas which remained stable, even amidst nationwide volatility, and disentangles the areas which drove the initial decline and subsequent resurgence.
-- Findings are augmented with a breakdown of the facilities present in each cluster, using data sourced from Open Street Map, and the crime type profile of each cluster.
-- Results are discussed in the context of geographic inequalities and opportunity theories of crime.
+Efforts to slow the spread of COVID-19 through social distancing and restrictions on citizen mobility have induced dramatic and unprecedented changes in police-recorded crime.
+Studies from international case study sites including the United States, United Kingdom and Australia have reported a decline in most crime types following the imposition of "stay at home" (or equivalent) orders [@estevez2020crime].
+Generally speaking, findings have aligned with expectations from opportunity perspectives on crime.
+Lockdown mobility patterns disrupted the typical convergence of motivated offenders, suitable targets and capable guardians, reducing opportunities for victimization [@stickle2020crime].
+As lockdown restrictions were slowly repealed, routine activities converged back to pre-pandemic levels and we witnessed a resurgence in criminal activity [@langton2021cs].
 
-## Background
+While existing studies have provided insight into the opportunity structures of crime, and the nuances of using police data to study lockdown as a natural experiment [@stickle2020crime; @langton2021cs] research has almost exclusively been undertaken using macro-level units of analysis, such as cities or countries.
+Less is known about the localized drivers of the lockdown crime drop or the degree to which macro-level trends are masking geographic inequalities in victimization.
+Pre-pandemic research examining the (in)stability of crime at fine-grained spatial scales [@weisburd2012criminology; @curman2015crime] and inequalities in the crime drop [@mcvie2020increasing; @ignatans2016whom; @akmed2021mon] would certainly suggest that local areas are unlikely to have experienced the macro-level lockdown crime trends in unison.
+Rather, we might expect specific places, typically associated with high ambient populations and plentiful opportunities for crime, to have driven the wider trend, with most localized areas remaining fairly stable.
+
+Examining the localized drivers of the lockdown crime drop represents the primary motivation of this paper.
+We disentangle the macro-level drop in police-recorded crime observed in England and Wales between February and August 2020.
+We deploy non-parametric longitudinal clustering to identify clusters of meso-level units which contributed disproportionately to the nationwide drop and subsequent resurgence in crime during the nationwide lockdown.
+The spatial patterning, opportunity structure and crime type profile of these local areas are quantified and summarized for their consistency with expectations for opportunity and mobility theories of crime [@halford2020crime].
+
+# Background
 
@@
 And yet, comparable questions remain.
-- To what extent have areas experienced this decline in unison? Have the dividends of the lockdown crime drop been shared equitably across space [@akmed2021mon]?
+- To what extent have local areas experienced this decline in unison?
+- Can we attribute most of the macro-level count trends to a handful of meso-level areas? 
+- Have the dividends of the lockdown crime drop been shared equitably across space [@akmed2021mon]?
 - Here, we use 7-months of police-recorded crime in England and Wales, aggregated to meso-level local units, to disentangle the nationwide lockdown crime drop.
-- We describe areas in terms of their opportunity structure and crime type profile.
+- We describe areas in terms of their opportunity structure, spatial patterning and crime type profile.
 
-## Data and Method
+# Data and Method
+
+To examine localized instability in the lockdown crime drop, we make use of three data sources, namely, open police-recorded crime data, geographic boundaries from Ordnance Survey and the Office for National Statistics (ONS), and data sourced from the Open Street Map API.
+Each of these are now outlined in turn, followed by an outline of the methods deployed.
+Code to replicate the data downloads, handling, analyses and visualization reported here is openly available (https://github.com/langtonhugh/covid_spatial).
 
@@
 
-- Police-recorded crime data for England and Wales (excluding Greater Manchester) between February and August, covering the first six months of lockdown March to August, and the preceding month (February), for comparison.
-- Data is available for thirteen notifiable offences, and anti-social behaviour (ASB).
-- As demonstrated in Langton et al., 2021 most notifiable offences experienced dramatic declines during April, the first full month of lockdown nationwide. 
-- Exceptions are drugs and ASB.
-- We provide context and a summary of nationwide trends decomposed by crime type in terms of raw trends (counts) and concentration (Generalized Gini coefficient).
-- We then proceed with a broader 'notifiable offences (excluding drugs)' category.
+Open police-recorded data on crime and anti-social behavior in England and Wales is published through an online web portal (https://data.police.uk/).
+Individual records are released on a month-by-month basis for each of the 43 police forces comprising England and Wales.
+We used a study period spanning February to August 2020 in order to capture the first six months of the nationwide lockdown (March to August) and the one month preceding the change (February).
+For reference and comparison to historical trends, we obtained data for the same months in 2018 and 2019.
+Individual records are time-stamped by month -- the temporal scale of this study.
+Due to ongoing data issues with Greater Manchester Police, we excluded data from this region, and therefore collated data from 42 out of 43 forces in England and Wales.
+
+Individual open records categorize crime according to thirteen different notifiable offence categories (e.g.
+burglary, violence and sexual offences, theft from the person, vehicle crime).^[Further information on these crime categories can be obtained from https://www.police.uk/pu/contact-the-police/what-and-how-to-report/what-report/.] Records also include anti-social behaviour (ASB) which usually captures less serious offences such as nuisance behavior and is not a notifiable offence.
+Individual records were aggregated to create count measures by month for the thirteen notifiable offences and ASB.
+These are used in the initial stages of analyses to provide a background overview of nationwide trends in crime and concentration before and after lockdown.
+For the primary cluster analysis, we use a count measure comprised of notifiable offences excluding drug crimes.
+This is due to the widespread recognition that drug crime trends, particularly during the COVID-19 lockdown, largely reflect policing proactivity rather than meaningful shifts in criminal behavior [@langton2021cs].
+Recognizing that aggregating data across crime types can mask variation [@andresen2017trajectories] we later decompose our findings according to the twelve remaining notifiable offences.
 
@@
 
-- Data is aggregated to Lower Super Output Area level, a neighborhood unit designed to be uniform by resident population (~ 1500 people), of which there are around 33,000 in England and Wales.
-- Due to some LSOA not having police.uk snap points, and one LSOA not having any police-recorded notifiable offences during the study period, we end up with a final sample/population size of N = 33,075.
+For localized monthly analysis, notifiable offences (excluding drugs) were aggregated by Lower Super Output Area (LSOA).
+LSOA are a meso-level geographic units designed for the reporting of official statistics at small geographies [@onslsoa].
+England and Wales is comprised of 32,844 LSOA designed to be uniform by resident population size.
+In 2019, LSOA housed an average of 1,700 people.
+Data obtained from the open police data portal (see previous section) includes a pre-assigned field stating the LSOA in which the crime occurred, as recorded by the police.
+Due to the spatial anonymization method used prior to data release, LSOA are the lowest level of aggregation at which we can reasonably assume spatial accuracy across multiple crime types [@tompson2015uk].
+Boundaries reflect 'on the ground' characteristics such as roads and railways [@onslsoachange], and as such, we consider LSOA to represent meaningful behavioral spaces to study the opportunity structures of crime.
+
+After removing crimes recorded by Greater Manchester Police, crimes recorded to have occurred within the Greater Manchester region, and four LSOAs which contained no crime between 2018 and 2020 (likely due to the spatial anonymization process), our final sample population for the meso-level analysis comprised 33,075 LSOAs.
 
@@
 
-- Data on facilities to describe the LSOA clusters obtained are retrieved from the Open Street Map API using the `osmdata` package [@osmrpack] in R [@rman].
-- Four count measures were chosen based on established crime attractors and generators, and availability and feasibility of obtaining nationwide data from Open Street Map, namely: nightlife, shops, public transport and bicycle parking.
-- Measure simply represent node counts.
+To summarize the opportunity structure of local areas we required a nationwide dataset of theoretically relevant facilities and urban features which could be aggregated at the LSOA level.
+To this end, we obtained point-level data from the Application Programming Interface (API) for Open Street Map via the `osmdata` package [@padgham2017osmdata] in `R` [@team2013r].
+Open Street Map is a crowdsourced geospatial database containing a vast array of features which can be used for explaining the temporal and spatial patterning of crime [@langton2020open].
+Geographic features are identified by pairs of keys and values which can be used to computationally query the API for data.
+Based on existing research examining the opportunity structures of fine-grained spatial scales, we collated the coordinate locations of the following facilities:
+
 - Nightlife: pubs, nightclubs, restaurants.
@@
 - Public transport: bus stops and railway stations.
-- Bicycle parking: bicycle parking.
-- Descriptive statistics on these counts at LSOA level are reported for each cluster solution.
-- To supplement these statistics, which as discussed later, are imperfect, we provide access to an interactive map for readers to explore independently.
+- Bicycle parking: bicycle parking lots.
 
-### Method 
-- We set the background with descriptive statistics on macro-level trends (raw counts) and concentration (generalized Gini coefficient) by crime type, including the total crime category.
-- This provides the 'global' picture of raw change and concentration at the national level, setting the reference point from which we can unmask the local variability underpinning the macro-level change.
-- For the main analysis, we deploy k-means [@genolini2011kml; @genolini2015kml], a non-parametric longitudinal clustering technique, on notifiable offences counts trajectories at LSOA level.
-- K-means has been used in crime concentration research to disentangle local variation from macro-level trends [@andresen2017trajectories; @curman2015crime].
-- Its deployment was largely inspired by a popular technique, group-based trajectory modelling (GBTM) [@weisburd2004trajectories].
-- K-means has a number of benefits over GBTM for crime and place research (detailed in @curman2015crime) but in this scenario, it has a unique advantage, in that it is highly capable of identifying short-term fluctuation, unlimited by polynomial terms [@akmed2021mon].
-- In this way, it is suitable for usage in this 'natural experiment' scenario, whereby we expect change to occur rapidly and dramatically from one month to the next, rather than smoothed over many years.
-- Cluster solution is determined by the Calinski-Criterion [@calinski1974dendrite] in alignment with existing studies in crime concentration research [@andresen2017trajectories; @akmed2021mon].
+The point-level data on these features were aggregated to create counts for each facility by LSOA.
+For simplicity, and due to issues of data sparsity, we sum the counts for each facility according to their overarching description (i.e.
+nightlife, shops, public transport, bicycle parking).
+We hypothesize that LSOAs containing a high number of facility counts across each domain will have higher pre-pandemic levels of crime, due to the plentiful opportunities for criminality, and in turn, steeper declines in crime following lockdown due to such opportunities suddenly becoming unavailable.
+We expect areas with _low_ counts across these domains to have similarly low crime levels pre-lockdown, and thus will remain low and stable following lockdown commencement.
 
-## Results
+## Method
 
-### Nationwide trends
+Analyses to examine the localized variation in the lockdown crime drop are conducted in three principal stages.
+First, an overview of nationwide (macro) trend is provided by crime type in terms of both absolute counts and the degree of concentration.
+Second, the macro-level trend is disentangled using non-parametric clustering techniques notifiable offences (excluding drugs) on the LSOA (meso) units (N = 33,075).
+Third, the characteristics of each cluster will be summarized in terms of their opportunity structures, spatial patterning and crime type profile.
+We conclude with a reflection on the findings given the crime profiles observed.
 
-- Raw counts by crime type confirm what has been demonstrated in existing accounts of lockdown crime in England and Wales (Langton et al., 2021).
-- Here, we augment these existing crimes with a 'total crime (excluding drugs)' category, representing all notifiable offences with the exception of drugs.
-- Drug crimes are thought to largely reflect police proactivity rather than a genuine shift in criminal behavioral.
--  We accept that granularity is lost with examining notifiable offences in their totality. We address this using a crime type breakdown, and return to it in more detail for the discussion.
+### Macro-level descriptives
+
+Macro-level descriptives of count trends and concentration by offence type are reported for provide context from which we will subsequently unmask local (meso) variation.
+First, we visualize observed crime counts between February and August 2020 relative to the same periods in 2018 and 2019 by offence type.
+In doing so, we can observe how nationwide crime trends changed in the face of lockdown measures [see also @langton2021cs].
+Second, we visualize generalized Gini coefficients [@bernasco2017more] to describe the degree of which crimes became more (or less) concentrated during lockdown at LSOA level.
+Again, we report figures for 2018 and 2019 as a point of reference.
+We use the generalized version of the Gini coefficient to account for data sparsity, as in some cases, the number of crimes fell below the number of LSOAs in the study region.
+Measures of counts and concentration set scene from which we can disentangle the underlying, meso-level variation.
+
+### Meso-level clustering
+
+To quantify the degree of meso-level uniformity underpinning the macro-level trend, and identify the potential drivers of the lockdown crime drop, we deploy a longitudinal variant of k-means clustering [@genolini2011kml; @genolini2015kml].
+This non-parametric clustering technique has an established role in crime and place research for examining the longitudinal trajectories of local areas in a macro-level crime drop scenario [@curman2015crime; @andresen2017trajectories].
+
+Here, we note that the experimental conditions of the COVID-19 lockdown make the usage of k-means particularly suitable.
+Existing research adopting the method has tended to investigate _long-term_ change over years or decades, focusing on the directional homogeneity (e.g.
+increasing, decreasing or stable) of clusters, rather than short-term volatility [@andresen2017trajectories].
+And yet, one key strength of k-means is that it is capable of identifying short-term fluctuation in longitudinal trends [@akmed2021mon].
+In this lockdown scenario, the availability of criminal opportunities were withdrawn quite literally overnight, and thus we might expect a similarly rapid and short-term change in crime at fine-grained spatial scales.
+This sets k-means apart from equivalent model-based methods restricted by polynomial terms, such as group-based trajectory modelling [@weisburd2004trajectories] and non-parametric techniques designed for long-term rather than short-term change [@akmed2021mon].
+
+We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOAs in the study region (N = 33,075) between February and August 2020.
+To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions using three, four, five and six clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite].
+For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable (see Appendix).
+This indicated that the optimal cluster solution contained six clusters.
+
+We visualize the final cluster solution in a manner which conveys the underlying distribution of observations comprising each cluster at each time point, rather than reporting a summary statistic (e.g.
+the mean trajectory) in isolation, as has been in the case in existing research.
+For each cluster, we also overlay the equivalent trajectories for 2018 and 2019 as a reference point for comparison to a 'typical' year.
+In doing so, we aim to not only identify localized (in)stability in the lockdown crime drop, but also assess the extent to which the trends observed have deviated from previous years.
+
+### Cluster characteristics
+
+We expect that the meso-level areas driving the lockdown crime drop will be those with plentiful opportunities for crime.
+That is, a disproportionately large volume of the decline (and subsequent resurgence) will be attributable to a handful of places which had pre-existing high crime levels as a result of their opportunity structure.
+Using the measures for opportunity generated from Open Street Map (i.e.
+nightlife, shops, public transport and bicycle parking), we report descriptive statistics on facility counts for each of the clusters obtained from the k-means analysis.
+In doing so, we expect to unpick a meaningful pattern which is consistent with the opportunity perspective of crime.
+
+To supplement this, we visualize the spatial patterning of the cluster solutions.
+For brevity and simplicity, we focus on Birmingham, Liverpool, Leeds, Bradford, Sheffield and Cardiff.
+We have excluded Greater Manchester due to the lack of police data, and given its size, we determined Greater London to warrant an individual case study for future research.
+Given these exclusions, the six cities we report represent the five most populous cities in England, and the most populous city in Wales.
+
+Finally, recognizing the unique opportunity structure of specific crimes, we summarize the crime type profile of clusters.
+For each cluster, we report the percentage breakdown of crimes types.
+We suspect that clusters will have differing crime type profiles according to the opportunity structures of each grouping.
+
+# Results
+
+## Nationwide trends
+
+To set the context for the localized analysis, Figure 2 visualizes crime counts between February and August 2020 for the thirteen notifiable offences, anti-social behavior, and notifiable offences excluding drugs.
+This descriptive analysis confirms what has been demonstrated using model-based forecasting analyses elsewhere (Langton et al., 2021).
+In April, the first full month of lockdown in England and Wales, we observe a widespread decline across most notifiable offences in comparison to previous years.
+One key exception is drug crimes, which as noted, is thought to have increased as a result of policing proactivity.
+Public order and violence and sexual offences experienced a dramatic decline in April, but appear to deviate little from previous years.
+This is due to long-term increase in these crime types (see Appendix A).
+The aggregate crime category, notifiable offences (excluding drugs) demonstrates a clear decline in March and April, followed by a resurgence to normality by August.
+This represents the 'global' trend which we will subsequently disentangle using localized analyses.
+
+Figure 3 visualizes the overall level of concentration across these offence types before and after lockdown, and in comparison to previous years, using the generalized Gini coefficient.
+Here, we note that most crime types experience remarkably stable concentration, even amidst volatile and sharply changing levels in absolute counts (see Figure 2).
+For many crime types, the degree of concentration is almost indistinguishable compared to 2018 and 2019.
+Exceptions include bicycle theft, theft from the person, other theft, robbery, and (to a lesser extent) vehicle crime.
+These experienced a fall in concentration (i.e.
+crimes were less concentrated, more spread out) compared to previous years.
+Overall, notifiable offences (excluding drugs) experienced minimal change.
+There is a marginal dip in concentration during the first month of lockdown, quickly returning to normal levels of concentration as crime surged back upwards during summer.
+
+Based on the dramatic decline and subsequent resurgence of notifiable offences (excluding drugs) during the study period, we might expect some localized variability.
+Measures of concentration indicate that there has been stability at the 'global' level, but the evidence-base surrounding the law of crime concentration and micro-place (in)stability would certainly give reason to suspect that this masks underlying volatility [@weisburd2015law; @andresen2017trajectories].
 
@@
 
-### Nationwide concentration
-
-- We note from the Generalized Gini coefficients that most crime types experienced remarkably stable 'global' concentration, even amidst this dramatic volatility in crime counts.
-- Exceptions include bicycle theft, theft from the person, other theft, robbery, and to some extent vehicle crime.
-- These experienced a fall in concentration (i.e. crimes were less concentrated, more spread out) compared to 2019.
-- Overall, however, notifiable offences (excluding drugs) experienced minimal change.
-There is a marginal dip in concentration during the first month of lockdown, quickly returning to normal levels of concentration as crime surged back upwards during summer.
-
 ```{r crimegini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Measure of concentration using the generalized Gini coefficient during the lockdown period."}
@@
 
-- Based on these raw counts and Gini concentration trends over time, we might expect localized variability in notifiable offences. 
-- Have neighborhoods experienced these trends collectively and uniformly?
-Can we attribute most of the macro-level count trends to a handful of areas?
-To what extent have the dividends of the crime drop been shared equitably across the country [@akmed2021mon]?
-- What is the oppportunity structure of the areas driving nationwide trends?
-- What is the crime type profile of the areas driving nationwide trends?
+## Longitudinal clustering
 
-### Longitudinal clustering
-
-#### Cluster trends
+### Cluster trends
 
@@
 - That said, there is some deviation from previous years in April, in alignment with expectations. There is a slight dip in the mean and median crime count, particularly in Cluster B.
-- The third largest cluster, cluster C, comprises 10% of all LSOAs in the country.
+- The third largest cluster, cluster C, comprises 10% of all LSOA in the country.
 This cluster experienced a more prominent dip in crime, along with cluster D (1.6% of LSOAs).
@@
 
-```{r kmeansnotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA notifiable offences.
-Mean (dotted line) and median (solid line) are shown for each cluster for 2020 (black), 2019 (red) and 2018 (blue) respectively.
+```{r kmeansnotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA notifiable offences (excluding drugs).
 Distributions refer to 2020 only.", out.width= "90%"}
-knitr::include_graphics("visuals/kmeans_violin_k6_1819_gg.png")
+knitr::include_graphics("visuals/kmeans_violin_full_gg.png")
 ```
 
-#### Contribution of each cluster
+### Contribution of each cluster
 
@@
 
-### Spatial distribution of clusters
+## Spatial distribution of clusters
 
-- List major city composition?
-- Maps of major city case studies?
-- Link to interactive map?
+- Link to interactive map
 
-### Characteristics of clusters
+```{r  maps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Spatial distribution of clusters by major urban conurbation.", out.width= "100%"}
+knitr::include_graphics("visuals/maps_gg.png")
+```
+
+## Characteristics of clusters
 
@@
 ```{r thresht, warning=F, message = F}
-osm_stats_df <- read_csv("data/osm_stats_sub_rounded_k6.csv")
+osm_stats_df <- read_csv("data/osm_stats_long.csv")
 
-osm_table <- kable(osm_stats_df, caption = "Descriptive statistics of facilities in each cluster. Sourced from Open Street Map.",
+osm_table <- kable(osm_stats_df, caption = "Descriptive statistics of facilities and features in each cluster. Sourced from Open Street Map.",
       booktabs = T, longtable = T, linesep = "", format = "latex") %>%
@@
 
-column_spec(osm_table, c(1:9), width = c("0.7cm", "1.5cm", "1.5cm", "1.5cm", "1.5cm", "1cm", "1cm", "1.5cm", "1cm"))
+column_spec(osm_table, c(1:7), width = c("4cm", "1.5cm", "1.5cm", "1.5cm", "1.5cm", "1.5cm", "1.5cm"))
 ```
 
-#### Crime type profile
+### Crime type profile
 
@@
 
-```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type characteristics of each cluster solution.", out.width= "90%"}
-knitr::include_graphics("visuals/traj_crimes_monthly_k6_gg.png")
-```
-
-## Discussion
+# Discussion
 
@@
 
-## Appendix
+# Appendix
+
+```{r longterm, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Select crime types with long-term increases in counts.", out.width= "90%"}
+knitr::include_graphics("visuals/long_term_gg.png")
+```
+
+```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type characteristics of each cluster solution.", out.width= "90%"}
+knitr::include_graphics("visuals/traj_crimes_monthly_k6_gg.png")
+```
+
+# References

### 2021-03-02 scripts/police_data_handling_cleaned.R
commit before going to NO only

@@ -273,7 +273,25 @@ total_crime_agg_df <- sub_data_agg_full_df %>%
   select(crime_type, year, month, ew_crime_count) %>% 
   separate(month, into = c("year", "month"), sep = "-") %>% 
   filter(month != "01", month != "09", month != "10", month != "11", month != "12") 
-  
+
+# Single plot for notifiable offences.
+no_raw_counts_gg <- ggplot(data = total_crime_agg_df) +
+  geom_line(mapping = aes(x = month, y = ew_crime_count, group = year, colour = year), size = 0.8) +
+  geom_vline(xintercept = 1.7, linetype = "dotted") +
+  scale_x_discrete(labels = str_extract(month.name[2:9], "^.{3}")) +
+  scale_color_manual(values = rev(c("black", "darkgrey", "lightgrey"))) +
+  labs(x = NULL, y = "Count", colour = NULL) +
+  theme_bw() +
+  theme(axis.text.x = element_text(size = 6), #, hjust = -0.4
+        axis.ticks = element_line(size = 0.3, lineend = "round"),
+        axis.text.y = element_text(size = 6),
+        strip.text = element_text(size = 8),
+        strip.background = element_rect(fill = "transparent"),
+        legend.position = "bottom")
+
+# Save.
+ggsave(plot = no_raw_counts_gg, filename = "visuals/no_raw_counts_gg.png", width = 10, height = 8, unit = "cm")
+
 # Then calculate these counts by crime type, bind the total crime data frame to it, and visualise.
 raw_counts_gg <- sub_data_agg_full_df %>% 
   group_by(crime_type, month, year) %>% 
@@ -309,7 +327,28 @@ gini_total_crime_df <- sub_data_agg_full_df %>%
   ungroup() %>% 
   separate(month, into = c("year","month"), sep = "-") %>% 
   mutate(crime_type = "Notifiable offences (excl. drugs)") %>% 
-  select(crime_type, gini_coef, year,  month)
+  select(crime_type, gini_coef, year,  month) %>% 
+  filter(month != "01", month != "09", month != "10", month != "11", month != "12") %>% 
+  mutate(year = recode_factor(year, "2018" = "2018", "2019" = "2019", "2020" = "2020")) 
+
+# Single plot for notifiable offences.
+no_gini_gg <- ggplot(data = gini_total_crime_df) +
+  geom_line(mapping = aes(x = month, y = gini_coef, group = year, colour = year), size = 0.8) +
[…]

### 2021-03-03 draft_paper.Rmd
commit before going to NO only
osm para and crime prof gg

@@
 burglary, violence and sexual offences, theft from the person, vehicle crime).^[Further information on these crime categories can be obtained from https://www.police.uk/pu/contact-the-police/what-and-how-to-report/what-report/.] Records also include anti-social behaviour (ASB) which usually captures less serious offences such as nuisance behavior and is not a notifiable offence.
-Individual records were aggregated to create count measures by month for the thirteen notifiable offences and ASB.
-These are used in the initial stages of analyses to provide a background overview of nationwide trends in crime and concentration before and after lockdown.
-For the primary cluster analysis, we use a count measure comprised of notifiable offences excluding drug crimes.
-This is due to the widespread recognition that drug crime trends, particularly during the COVID-19 lockdown, largely reflect policing proactivity rather than meaningful shifts in criminal behavior [@langton2021cs].
-Recognizing that aggregating data across crime types can mask variation [@andresen2017trajectories] we later decompose our findings according to the twelve remaining notifiable offences.
+Individual records were aggregated to create a count measure for notifiable offences (excluding drugs) by month either at the nationwide (macro) level, or the localized (meso) level, as detailed in the next section.
+There widespread recognition that drug crime trends, particularly during the COVID-19 lockdown, largely reflect policing proactivity rather than meaningful shifts in criminal behavior [@langton2021cs].
+Recognizing that aggregating data across crime types can mask variation [@andresen2017trajectories] we later decompose our main findings according to the twelve remaining notifiable offences, and refer to additional analyses in the Appendix.
 
@@
 
-For localized monthly analysis, notifiable offences (excluding drugs) were aggregated by Lower Super Output Area (LSOA).
+For nationwide (macro) descriptive statistics, which set the background for the main analyses, we use the geographic region of England and Wales, noting the exclusion of Greater Manchester.
+For localized (meso) analysis, we aggregate offences to Lower Super Output Area (LSOA).
 LSOA are a meso-level geographic units designed for the reporting of official statistics at small geographies [@onslsoa].
@@
 Boundaries reflect 'on the ground' characteristics such as roads and railways [@onslsoachange], and as such, we consider LSOA to represent meaningful behavioral spaces to study the opportunity structures of crime.
-
-After removing crimes recorded by Greater Manchester Police, crimes recorded to have occurred within the Greater Manchester region, and four LSOAs which contained no crime between 2018 and 2020 (likely due to the spatial anonymization process), our final sample population for the meso-level analysis comprised 33,075 LSOAs.
+After removing crimes recorded by Greater Manchester Police, crimes recorded to have occurred within the Greater Manchester region, and four LSOAs which contained no crime between 2018 and 2020 (likely due to the spatial anonymization process), our final sample population for the meso-level analysis comprised 33,075 LSOA.
 
@@
 Analyses to examine the localized variation in the lockdown crime drop are conducted in three principal stages.
-First, an overview of nationwide (macro) trend is provided by crime type in terms of both absolute counts and the degree of concentration.
-Second, the macro-level trend is disentangled using non-parametric clustering techniques notifiable offences (excluding drugs) on the LSOA (meso) units (N = 33,075).
+First, an overview of nationwide (macro) trend is provided in terms of both absolute counts and the degree of concentration.
+Second, the macro-level trend is disentangled using non-parametric clustering techniques on the LSOA (meso) units (N = 33,075).
 Third, the characteristics of each cluster will be summarized in terms of their opportunity structures, spatial patterning and crime type profile.
-We conclude with a reflection on the findings given the crime profiles observed.
 
@@
 
-Macro-level descriptives of count trends and concentration by offence type are reported for provide context from which we will subsequently unmask local (meso) variation.
-First, we visualize observed crime counts between February and August 2020 relative to the same periods in 2018 and 2019 by offence type.
+Macro-level descriptives of count trends and concentration for notifiable offences (excluding drugs) are reported for provide context from which we will subsequently unmask local (meso) variation.
+First, we visualize observed counts between February and August 2020 relative to the same periods in 2018 and 2019.
 In doing so, we can observe how nationwide crime trends changed in the face of lockdown measures [see also @langton2021cs].
-Second, we visualize generalized Gini coefficients [@bernasco2017more] to describe the degree of which crimes became more (or less) concentrated during lockdown at LSOA level.
+Second, we visualize generalized Gini coefficients [@bernasco2017more] to describe the degree of which offences became more (or less) concentrated during lockdown at LSOA level.
 Again, we report figures for 2018 and 2019 as a point of reference.
-We use the generalized version of the Gini coefficient to account for data sparsity, as in some cases, the number of crimes fell below the number of LSOAs in the study region.
+We use the generalized version of the Gini coefficient to account for data sparsity, as for some crime types reported in the Appendix, the number of crimes falls below the number of LSOAs in the study region.
 Measures of counts and concentration set scene from which we can disentangle the underlying, meso-level variation.
@@
 
-We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOAs in the study region (N = 33,075) between February and August 2020.
+We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOA in the study region (N = 33,075) between February and August 2020.
 To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions using three, four, five and six clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite].
 For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable (see Appendix).
-This indicated that the optimal cluster solution contained six clusters.
 
@@
 Given these exclusions, the six cities we report represent the five most populous cities in England, and the most populous city in Wales.
+Study regions are defined by based on the city names appearing in LSOA name.
 
@@
 
-To set the context for the localized analysis, Figure 2 visualizes crime counts between February and August 2020 for the thirteen notifiable offences, anti-social behavior, and notifiable offences excluding drugs.
-This descriptive analysis confirms what has been demonstrated using model-based forecasting analyses elsewhere (Langton et al., 2021).
-In April, the first full month of lockdown in England and Wales, we observe a widespread decline across most notifiable offences in comparison to previous years.
-One key exception is drug crimes, which as noted, is thought to have increased as a result of policing proactivity.
-Public order and violence and sexual offences experienced a dramatic decline in April, but appear to deviate little from previous years.
-This is due to long-term increase in these crime types (see Appendix A).
-The aggregate crime category, notifiable offences (excluding drugs) demonstrates a clear decline in March and April, followed by a resurgence to normality by August.
-This represents the 'global' trend which we will subsequently disentangle using localized analyses.
+To set the context for the localized analysis, Figure 2 visualizes crime counts between February and August 2020 for notifiable offences excluding drugs before and after lockdown.
+This descriptive analysis broadly confirms what has been demonstrated using model-based forecasting analyses elsewhere (Langton et al., 2021) although we note that this descriptive comparison does not account for long-term trends.
+In April, the first full month of lockdown in England and Wales, we observe a nationwide decline in notifiable offences in comparison to previous years.
+Upon the relaxation of lockdown rules, crime resurged back to normality, and by August, crime had returned to within a range we might have expected without the nationwide lockdown.
+This trend represents the 'global' trend which we will subsequently disentangle using localized analyses.
 
-Figure 3 visualizes the overall level of concentration across these offence types before and after lockdown, and in comparison to previous years, using the generalized Gini coefficient.
-Here, we note that most crime types experience remarkably stable concentration, even amidst volatile and sharply changing levels in absolute counts (see Figure 2).
-For many crime types, the degree of concentration is almost indistinguishable compared to 2018 and 2019.
-Exceptions include bicycle theft, theft from the person, other theft, robbery, and (to a lesser extent) vehicle crime.
-These experienced a fall in concentration (i.e.
-crimes were less concentrated, more spread out) compared to previous years.
-Overall, notifiable offences (excluding drugs) experienced minimal change.
-There is a marginal dip in concentration during the first month of lockdown, quickly returning to normal levels of concentration as crime surged back upwards during summer.
-
-Based on the dramatic decline and subsequent resurgence of notifiable offences (excluding drugs) during the study period, we might expect some localized variability.
-Measures of concentration indicate that there has been stability at the 'global' level, but the evidence-base surrounding the law of crime concentration and micro-place (in)stability would certainly give reason to suspect that this masks underlying volatility [@weisburd2015law; @andresen2017trajectories].
-
-```{r crimeoverall, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Macro-level crime trends in England and Wales during the lockdown period in 2020."}
-knitr::include_graphics("visuals/raw_counts_gg.png")
+```{r nocounts, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Notitifable offence (excluding drugs) counts in England and Wales during the lockdown period.", fig.align='center'}
+knitr::include_graphics("visuals/no_raw_counts_gg.png")
 ```
 
-```{r crimegini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Measure of concentration using the generalized Gini coefficient during the lockdown period."}
-knitr::include_graphics("visuals/gini_gg.png")
+Figure 3 visualizes the overall level of concentration before and after lockdown, and in comparison to previous years, using the generalized Gini coefficient.
+Here, it is clear that notifiable offences, even amidst volatile and sharply changing levels in absolute counts (see Figure 2), experienced remarkably stable levels of concentration.
+There is a marginal dip in concentration during April, the first month of lockdown.
+This suggests that as crimes fell, those were occurring were _less_ concentrated.
+That is, crimes occurring during the height of lockdown were more spread out that we might have otherwise expected.
+That said, concentration quickly returned to normal levels.
+
+```{r nogini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Monthly generalized Gini cofficient trends for notitifable offences (excluding drugs) in England and Wales during the lockdown period.", fig.align='center'}
+knitr::include_graphics("visuals/no_gini_gg.png")
 ```
+
+Based on the dramatic decline and subsequent resurgence of notifiable offences during the study period, we might expect some localized variability.
+Measures of concentration indicate that there has been stability at the 'global' level, but the evidence-base surrounding the law of crime concentration and micro-place (in)stability would certainly give reason to suspect that this masks underlying volatility [@weisburd2015law; @andresen2017trajectories].
+With this in mind, we now report the localized, meso-level analysis which unpicks the macro-level trend.
 
@@
 
-- Figure 3 visualizes the cluster solution generated by k-means, with 6-clusters identified as optimal based on the Calinski-Criterion [@calinski1974dendrite].
-- Black solid line is the median count of the cluster at any given time point, with the black dotted line showing the mean.
-- Violin plots have been added to convey the variation in counts at each time point.
-Here, we note that distributions were fairly uniform, with the mean and median representing the cluster solutions fairly well.
-There is little overlap between clusters.
-- For reference to what trends we may have expected in a 'typical' year without the pandemic, we have added the mean and median trend lines for each cluster during 2018 (in blue) and 2019 (in red).
-- This demonstrates that the clusters identified using the 2020 data were distinct and meaningful even in previous years, and highlights the gap between previous years and what has been observed during the pandemic.
-- Overall, we note that most LSOA were remarkably stable during the pandemic.
-Clusters A and B can be described and 'low crime and stable', exhibiting fairly low counts throughout the study period and across years.
-Together, these clusters comprise 89% of LSOAs in England and Wales.
-- That said, there is some deviation from previous years in April, in alignment with expectations. There is a slight dip in the mean and median crime count, particularly in Cluster B.
-- The third largest cluster, cluster C, comprises 10% of all LSOA in the country.
-This cluster experienced a more prominent dip in crime, along with cluster D (1.6% of LSOAs).
+We deploy non-parametric k-means clustering on LSOA (meso) level geographic units (N = 33,075) comprising England and Wales to decompose the macro-level 'decline and resurgence' observed between February and August, 2020.
+Figure 3 visualizes the resulting 6-cluster solution determined by the optimal Calinski-Criterion statistic [@calinski1974dendrite].
+Solid black lines represent the median count for each cluster at any given time point, with the black dotted line showing the mean.
+Violin plots have been added to convey variation around these points at each time point.
+These suggest that the mean and median point statistics summarize the underlying data reasonably well, and indicate that clusters are distinct from one another, with little overlap.
+Additional lines have been added to convey each clusters' mean and median trend in 2018 and 2019 respectively.
+These trends suggest that the clusters identified using the 2020 study period were distinct and meaningful even in previous years, and provide a relative baseline from which we can compare lockdown trends.
+
+Overall, we note that most LSOA were remarkably stable during the pandemic.
+Clusters A and B could be described and 'low crime and stable', exhibiting fairly low counts throughout the study period and across years.
+Together, these clusters comprise 89% of LSOA in England and Wales.
+Even amidst the stark macro-level decline in notifiable offences (see Figure 2) LSOA in these clusters only experienced marginal average dips in crime.
+The third largest cluster, cluster C, comprises 10% of LSOA in England and Wales.
+LSOA in this cluster experienced a more prominent dip in crime, along with cluster D (1.6% of LSOA).
 Together, we might describe these LSOAs as 'mid-crime, mid-drop'.
-- However, the 'action' appears to be occurring amongst a small subset of LSOAs, contained within clusters E and F.
-Together, these clusters comprise only 0.33% of all LSOA in the country (N = 110), however they are 'high crime, major drop'.
-- Interestingly, clusters E and F represent the only clusters which did not return back to normality by August (if we assume that 2018 and 2019 represent a trend which we might have otherwise expected).
-- Overall, we note that despite the dramatic nationwide trend in notifiable offences (excluding drugs), there has been remarkable stability in local areas, with most change being driven by a small number of areas.
+In both cases, there is a clear deviation from previous years.
+Notifiable offences fall between March and April, and then began to converge back to levels observed in previous years.
 
 ```{r kmeansnotoff, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "K-means cluster solutions for LSOA notifiable offences (excluding drugs).
-Distributions refer to 2020 only.", out.width= "90%"}
+Distributions refer to 2020 only.", out.width= "90%", fig.align='center'}
 knitr::include_graphics("visuals/kmeans_violin_full_gg.png")
@@
 
+That said, most 'action' appears to be occurring amongst a small subset of LSOA.
+Clusters E and F collectively comprise only 0.33% (N = 110) of LSOA in England and Wales.
+Yet, their crime counts are much higher, and the decline between March and April is considerable.
+We might therefore describe these clusters as 'high crime, major drop': LSOA with plentiful opportunities for crime in typical times, and in turn, LSOA which are most sensitive to the restriction in opportunities which followed after the imposition of lockdown.
+Interestingly, clusters E and F represent the only clusters which remained below pre-pandemic levels by August.
+In both cases, the average counts in February 2020 were _higher_ than those observed in previous years, but following the sharp decline and subsequent increase by August, these averages were _lower_ than 2018 and 2019.
+
 ### Contribution of each cluster
 
-- To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the monthly change in counts of each cluster.
-- Alongside these raw counts, we also include the percentage of total absolute change (i.e. positive or negative) attributable to each cluster.
-- We use this to identify which clusters drove the initial decline and subsequent resurgence nationwide.
-- As expected, the vast majority of change occurred between March and April.
-- However, Figure 4 demonstrates that this substantial decline did not occur equitably across the country.
-Consistent with the cluster solutions trends in Figure 3, it is clear that a small number of areas drove the trend.
-- For instance, clusters A and B, which comprise 89% of LSOAs in the country, accounted for only 47% of the decline between March and April.
-- By contrast, clusters E and F, which together comprised only 0.33% of LSOA (N = 110) contributed to 15% of the nationwide drop between months.
-This a disproportionately large impact of the macro-level trend, and further supports the observation that most areas remained fairly stable.
-- Interestingly, this picture shifts upon the beginning of the resurgence between April and May.
-- Cluster A continues to contribute a disproportionately small amount to nationwide change, but cluster B contributes to 42% of the increase, despite only comprising 31% of LSOA.
-- The contribution of the smallest clusters E and F is remarkably small compared to the initial decline, 2% and 0% respectively.
+To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the month-by-month change in counts.
+Alongside these raw counts, we also include the percentage of total absolute change (i.e.
+positive or negative) attributable to each cluster.
+We can use this visual to identify which clusters drove the initial decline and subsequent resurgence nationwide.
+
+As expected, the vast majority of change across all clusters occurred between March and April.
+Between these months, notifiable offences experienced a dramatic fall nationwide.
+That said, the figure demonstrates that this decline did not occur equitably across local areas.
+Consistent with the cluster solutions trends in Figure 3, a small number of LSOA (meso) units appear to have contributed disproportionately to the nationwide (macro) trend.
+For instance, clusters A and B, which comprise 58% and 31% of LSOA in the country, accounted for only 20% and 27% of the total decline between March and April.
+By contrast, clusters E and F, which together comprised only 0.33% of LSOA (N = 110) contributed to 15% of the nationwide drop in these months.
+Again, this suggests that most areas remarkably stable, given the nationwide volatility, with just a handful of areas driving the decline the nationwide lockdown crime drop.
+
+This picture shifts slightly upon the beginning of the resurgence between April and May.
+Clusters A and B continue to contribute a disproportionately _small_ amount to nationwide change, but the proportion increases, with each contributing 24% and 42% to the increase respectively.
+The contribution of the smallest clusters E and F is small compared to the initial decline, 2% and 0% (when rounded), respectively.
 This can be largely attributed to the flattening of their crime count trends between these months.
-In other words, following their initially steep decline between March and April, crime remained reasonably unchanged between April and May in these areas.
-- Overall, this pattern continues throughout the spring convergence back to normality, with some notable changes. 
-- Following the period of stabilization, clusters E and F gather pace during the resurgence, as does cluster D.
-For instance, between June and July, these three clusters account for nearly a third of the nationwide increase, despite comprising less than 2% of total LSOA.
-- July to August represents an interesting period.
+In other words, following their initially steep decline between March and April, crime remained reasonably unchanged between April and May in the 'high crime, major drop' areas.
+
+Following the period of stabilization, clusters E and F gather pace during the resurgence, as does cluster D.
+For instance, between June and July, these three clusters collectively account for 27% of the nationwide increase, despite comprising less than 2% of LSOA in the country.
+July to August represents an interesting period.
 Until that point, each cluster, despite their vastly differing relative contributions to change, were directionally homogeneous.
 In other words, every cluster declined between February and April, and then increased between April and July.
-- Between July and August, clusters began to diverge. Nationwide, there was a marginal increase in notifiable offence counts (see Figure 1), but this masks a great deal of localized variability.
-- The two large, low crime clusters (A and B) actually declined again between July and August, while the higher crime clusters continued to increase.
-This is suggestive of a local change in opportunities as lockdown rules were relaxed - change which has been aggregated away at the macro-level.
-- In other words, low crime areas mitigated against further nationwide increases which may, if they had continued, actually resulted in a higher crime count in August than observed in 2019.
+Between July and August, clusters began to diverge.
+Nationwide, there was a marginal increase in notifiable offence counts (see Figure 1), but this clearly masks a great deal of localized variability.
 
-```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Counts and percentage of nationwide change between months attributable to each cluster.", out.width= "100%"}
+The 'low crime and stable' clusters (A and B) actually _declined again_ between July and August, and the higher crime clusters continued to increase.
+The change is almost imperceptible when viewing average counts in these LSOA (see Figure 3) but the sheer size of these units, collectively comprising 89% of England and Wales, has a major impact on nationwide trends.
+This is suggestive of highly localized change in opportunities as lockdown rules were relaxed: change which is aggregated away at the macro-level.
+Stable, low crime areas mitigated against further nationwide increases which may, if they had continued, actually resulted in a higher crime count in August than observed in previous years.
+
+```{r changeline, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Counts and percentage of nationwide change between months attributable to each cluster.", out.width= "100%", fig.align='center'}
 knitr::include_graphics("visuals/change_k6_gg.png")
-```
-
-## Spatial distribution of clusters
-
-- Link to interactive map
-
-```{r  maps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Spatial distribution of clusters by major urban conurbation.", out.width= "100%"}
-knitr::include_graphics("visuals/maps_gg.png")
 ```
@@
 
-- We have now identified that, amidst dramatic nationwide change in notifiable offences during lockdown, and stability in overall crime concentrations, there has been considerable underlying volatility, with a small number of local areas driving the macro-level trends.
-- From an opportunity perspective, we expect the clusters to have differing opportunity structures.
-- For instance, LSOA which were typically 'high crime' in 2018 and 2019, and then experienced a dramatic fall in crime upon lockdown commencement, are likely to have a large volume of criminal opportunities which were closed off following restrictions on mobility.
-- On the other hand, areas which were typically low crime, and remained low crime, are likely to have a vastly different opportunity structure: places with minimal attractors or generators to begin with are unlikely to have changed dramatically during lockdown.
-- Table 1 demonstrates this point using the data obtained from Open Street Map.
-- Clusters A and B, the 'low crime and stable' LSOAs, tend to have few major opportunities for crime.
-Few pubs, nightclubs or restaurants to generate a violence and theft-prone nightlife, minimal shopping opportunities (and thus less chance of shoplifting), limited public transport (less theft) and less bicycle parking (less bicycle theft).
-- By contrast, the 'high crime major drop' clusters (e.g.
-E and F) are major nightlife areas, have considerable shopping opportunities, often have expansive public transport networks, and have many bicycle parking facilities.
-- These findings are consistent with what is observed using the interactive map made available as part of this paper (link here).
+# Spatial distribution
+
+The provide local context to the cluster solutions identified through the k-means clustering, we visualize the spatial patterning of groupings for six major cities in England and Wales (see Figure 5).
+We find distinct geographic pattern to the clusters identified.
+Here, it is worth noting that the Bradford region contains the city of Bradford, but also the satellite town of Keighley to the north west.
+The cluster map thus reflects these two urban centers.
+Without exception, the city centres and commercial districts of each city are characterised by 'high crime major drop' clusters E and F, which in turn often neighbor the 'mid crime mid drop' cluster D.
+By contrast, the 'low crime and stable' clusters tend to sit outside of the city centres: cluster B in the suburbs and the sparsely populated cluster A LSOA on the periphery of urban conurbations.
+For a detailed investigation of the spatial clusters identified for the whole of England and Wales, we refer readers to an openly available online map (anonymized link here).
+
+```{r  maps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Spatial distribution of clusters by major urban conurbation.", out.width= "100%", fig.align='center'}
+knitr::include_graphics("visuals/maps_gg.png")
+```
+
+# Opportunity structure
+
+We have now identified that, amidst dramatic nationwide change in notifiable offences during lockdown, and stability in overall crime concentrations, there has been considerable underlying volatility, with a small number of local areas driving the macro-level trend.
+From a theoretical perspective, we expect the clusters to have differing opportunity structures.
+Table 1 reports a series of descriptive statistics based on the opportunity structure of each cluster using the nationwide data obtained from the Open Street Map API.
+These findings largely support our expectations, namely, that large, stable clusters have few opportunities for crime, while high crime areas which were responsible for a disproportionately large amount of the crime drop have plentiful features which facilitate crime.
+
+By way of example, there are sparse opportunities for crime in the largest and most stable cluster A.
+Most LSOA in this cluster have no nightlife facilities, no shops and no bicycle parking.
+Public transport is available but it is not common: the median LSOA in cluster A only contained four bus stops or railways stations.
+As we move along to medium and high crime clusters, which had higher pre-existing crime levels, and thus steeper declines during lockdown, these counts markedly increase.
+For instance, LSOA in cluster E contain a median of 21 nightlife facilities, 28 shops, 24 public transport nodes and 13 bicycle parking spaces.
+These counts increase further for cluster F, although we note that the cluster contains only nine LSOA.
 
@@
 
-### Crime type profile
+# Crime type profile
 
@@
 - At the same time, the increase share in public order may reflect other activity, such as the protests around BLM and anti-lockdown movements, which take place in city centres for coverage.
+
+```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type characteristics of each cluster solution.", out.width= "100%", fig.align='center'}
+knitr::include_graphics("visuals/traj_crimes_monthly_k6_gg.png")
+```
 
@@
 
-```{r longterm, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Select crime types with long-term increases in counts.", out.width= "90%"}
+```{r crimeoverall, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Counts by offence type in England and Wales during the lockdown period.", fig.align='center'}
+knitr::include_graphics("visuals/raw_counts_gg.png")
+```
+
+```{r crimegini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Monthly generalized Gini cofficient trends by offence type in England and Wales during the lockdown period.", fig.align='center'}
+knitr::include_graphics("visuals/gini_gg.png")
+```
+
+```{r longterm, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Select crime types with long-term increases in counts.", out.width= "90%", fig.align='center'}
 knitr::include_graphics("visuals/long_term_gg.png")
@@
 
-```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type characteristics of each cluster solution.", out.width= "90%"}
-knitr::include_graphics("visuals/traj_crimes_monthly_k6_gg.png")
-```
-
 # References

### 2021-03-03 scripts/police_data_handling_cleaned.R

@@ -631,18 +631,28 @@ traj_crime_types_df <- traj_crime_types_df %>%
   left_join(traj_names_df)
 
 # Plot.
-traj_crimes_gg <- ggplot(data = traj_crime_types_df) +
+traj_crimes_gg <- ggplot(data = traj_crime_types_df) + 
+  theme_bw() +
   geom_bar(mapping = aes(x = month, y = traj_ct_prop, group = crime_type, fill = crime_type),
             stat = "identity", colour = "black", size = 0.2) +
-  facet_wrap(~traj_titles, ncol = 2) +
+  facet_wrap(~traj_titles, ncol = 6) +
   labs(y = "Proportion comprising total crime", x = NULL, fill = NULL) +
   scale_x_discrete(labels = c(str_extract(month.name[2:8], "^.{3}"), character(1))) +
-  guides(fill = guide_legend(ncol = 3)) +
-  theme_bw() +
-  theme(legend.position = "bottom", strip.background = element_rect(fill = "transparent"))
+  guides(fill = guide_legend(nrow = 2)) +
+  theme(legend.position = "bottom",
+        legend.text = element_text(size = 5),
+        strip.background = element_rect(fill = "transparent", size = 0.2),
+        strip.text = element_text(size = 6),
+        axis.text.y = element_text(size = 5),
+        axis.text.x = element_text(size = 5, angle = 90, vjust = 0.2),
+        axis.title = element_text(size = 6),
+        axis.ticks = element_line(size = 0.3),
+        panel.border = element_rect(size = 0.3),
+        legend.key.width = unit(0.3, "cm"),
+        legend.key.height = unit(0.3, "cm"))
 
 # Save.
-ggsave(plot = traj_crimes_gg, filename = "visuals/traj_crimes_monthly_k6_gg.png", width = 16, height = 24, unit = "cm")
+ggsave(plot = traj_crimes_gg, filename = "visuals/traj_crimes_monthly_k6_gg.png", width = 16, height = 5, unit = "cm")
 
 # Save and load workspace as appropriate.
 # save.image(file = "data_handling_cleaning.RData")

### 2021-03-04 draft_paper.Rmd
crime type profiles

@@
 
-### Crime data
+## Crime data
 
@@
 
-### Unit of analysis
+## Unit of analysis
 
@@
 
-### Open Street Map
+## Open Street Map
 
@@
 
-## Method
+# Method
 
@@
 
-### Macro-level descriptives
+## Macro-level descriptives
 
@@
 
-### Meso-level clustering
+## Meso-level clustering
 
@@
 We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOA in the study region (N = 33,075) between February and August 2020.
-To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions using three, four, five and six clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite].
+To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite].
 For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable (see Appendix).
@@
 
-### Cluster characteristics
+## Cluster characteristics
 
@@
 
-# Spatial distribution
+### Spatial distribution
 
@@
 
-# Opportunity structure
+### Opportunity structure
 
@@
 These counts increase further for cluster F, although we note that the cluster contains only nine LSOA.
+This is consistent with earlier findings which suggest that the crime decline most in these clusters: areas previously rich in opportunity, in which crime was pervasive, became the drivers of the lockdown crime drop.
 
@@
 
-# Crime type profile
+### Crime type profile
 
-- For the purposes of simplicity and setting a benchmark, we have been using a measure of 'notifiable offences excluding drugs' for analysis.
-- That said, we might expect the clusters identified to have different crime type profiles.
-- Given the differing opportunity structures between clusters, the 'typical' crime profiles of each cluster may change considerably as lockdown commences.
-- Alternatively, the pervasiveness of lockdown on mobility may have impacted on crime types uniformly within each cluster, in which case, we may see stability in crime type profiles before and after lockdown.
-- Figure 5 summarises the profile of each cluster by calculating the proportion of total crime in each cluster.
-- We observe considerable stability in the distribution of crime types both within and between clusters.
-- Violence and sexual offences represent the most prevalent crime type across all clusters.
-The proportion of total crime in each cluster attributable to violence and sexual offences increases in the first two months of lockdown.
-- This by no means implies that violence and sexual offences increased, but rather, that violence and sexual offences declined at a slower rate than other crime types, increasing its share of all crimes.
-- In cluster F, we see a unique pattern, whereby theft from the person and other theft are squeezed considerably into April and May, while the share of public order increased.
-- This might reflect the opportunity structures of LSOA in cluster F being weighted heavily towards theft (e.g. lots of public transport nodes).
-- At the same time, the increase share in public order may reflect other activity, such as the protests around BLM and anti-lockdown movements, which take place in city centres for coverage.
+Given the unique longitudinal trends, spatial patterning and opportunity structure of the clusters identified using the aggregate notifiable offences measure, we might expect the clusters identified to have distinct crime type profiles.
+That is, some crime types may be more or less prevalent in certain clusters to begin with and / or there may be variation in the extent to which lockdown impeded opportunities across crime types.
+To explore this.
+Figure 6 summarises the crime type profile of each cluster.
+
+In February, before the restrictions on mobility and social interaction, the crime type profiles of each cluster were already distinct.
+Higher crime clusters (e.g.
+E, and F) were weighted heavily towards shoplifting and theft.
+Low crime and stable clusters, by contrast, had higher proportions of criminal damage and arson, burglary and violence and sexual offences.
+
+Overall, we observe a degree stability in the distribution of crime types both within and between clusters.
+The largest, stable clusters (A and B) have remarkably similar crime type profiles.
+Before lockdown, for each of these clusters, the most prevalent crimes were violence and sexual offences, both comprising 41% of total crime.
+This proportion increase on lockdown commencement, largely at the expense of burglary and vehicle crime, which decreased as a proportion of total crime.
+In clusters A and B, and indeed across all clusters, the proportion of total crime attributable to public order also increased.
+These increases may reflect a seasonal effect: crimes such as violence and sexual offences and public order tend to increase between February and August in typical times, while vehicle crime and burglary are usually stable (@langton2021cs; see Appendix).
+This does not, by any means, indicate that these crimes increased during lockdown -- counts declined considerably -- but rather, these crimes declined less steeply relative to other offence categories.
+In the case of public order, this may be a result of lockdown-specific activity.
+Public order includes offences relating to processions and assemblies, and thus may capture gatherings which violated COVID-19 guidelines [@crown2021], and protests such as those relating to Black Lives Matter.
+Increases in public order offences across these areas would slow and work against a fall in a more typical public order crimes.
+
+Notable disparities between clusters arise with regards to theft from the person and other theft in cluster F, which as reported earlier, drove a considerable part of the nationwide decline in crime.
+Theft in these LSOA was squeezed upon lockdown commencement.
+In February, theft crime types comprised 53% of total crime, but by April, this figure had dropped to 13%.
+This occurred despite these crime types typically remaining stable during equivalent times of year (@langton2021cs; see Appendix).
+Somewhat surprisingly, given the spatial distribution and opportunity structures of clusters E and F in city centres, shoplifting did not experience a similar squeeze in these LSOA.
 

### 2021-03-05 draft_paper.Rmd
writing

@@
 Efforts to slow the spread of COVID-19 through social distancing and restrictions on citizen mobility have induced dramatic and unprecedented changes in police-recorded crime.
-Studies from international case study sites including the United States, United Kingdom and Australia have reported a decline in most crime types following the imposition of "stay at home" (or equivalent) orders [@estevez2020crime].
-Generally speaking, findings have aligned with expectations from opportunity perspectives on crime.
+Studies from international case study sites have reported a decline in most crime types following the imposition of "stay at home" (or equivalent) orders [@estevez2020crime].
+Findings have tended to align with expectations from opportunity perspectives on crime.
 Lockdown mobility patterns disrupted the typical convergence of motivated offenders, suitable targets and capable guardians, reducing opportunities for victimization [@stickle2020crime].
@@
 
-While existing studies have provided insight into the opportunity structures of crime, and the nuances of using police data to study lockdown as a natural experiment [@stickle2020crime; @langton2021cs] research has almost exclusively been undertaken using macro-level units of analysis, such as cities or countries.
+While existing studies have provided insight into the opportunity structures of crime, and the nuances of using police data to study lockdown as a natural experiment, research has almost exclusively been undertaken using macro-level units of analysis, such as cities or countries.
 Less is known about the localized drivers of the lockdown crime drop or the degree to which macro-level trends are masking geographic inequalities in victimization.
@@
 
-- Recent months have seen a wave of interest in describing and explaining changes in  police-recorded crime in response to restrictions on mobility brought about by efforts to limit the spread of COVID-19.
-- Generally speaking, findings have been consistent with theoretical expectations from an opportunity perspective, namely, that lockdown restrictions disrupted the convergence of motivated offenders, (in)capable guardians and suitable targets, bringing about widespread, macro-level declines in crime.
-- Some notable cases demonstrate that changes may be specific to crime type, although some studies using longer time periods have shown that change has been relatively uniform across crime types, with some exceptions (Langton et al., 2021).
+Recent months have seen a wave of interest in describing and explaining changes in crime in response to restrictions on mobility brought about to limit the spread of COVID-19.
+Studies include case study sites in the United States [@campedelli2020exploring; @gerell2020minor; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show] and Mexico [@estevez2020crime].
+Generally speaking, findings are consistent with opportunity perspectives and the mobility theory on the pandemic [@langton2021cs].
+That is, legally-enforced restrictions on people's mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime.
+As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often to levels that would have been expected without the global pandemic and ensuing lockdowns.
+
+That said, little is known about the localized drivers of these macro-level trends.
+An exploration of the extent to which the macro lockdown crime drop has been experienced equitably across space is valuable for two principal reasons.
+First, we would expect only a small number of fine-grained geographic units to have driven the lockdown crime drop.
+In pre-pandemic era, longitudinal studies have consistently demonstrated that, in a crime drop scenario, most local areas remained remarkably stable, with a disproportionatly large volume of change attributable to just handful of meso or micro-level units (e.g.
+@weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local).
+Findings have contributed to the evidence-base around routine activities theory, reflecing the considerable variation in criminal opportunities across (even spatially proximate) places [@eck2000finding; @weisburd2012criminology].
+
+Secondly,
+
+from an opportunity and mobility perspective,
+
+preliminary evidence from the United States suggests that macro lockdown crime trends might mask underlying variation.
+Using police-recorded crime data in San Francisco and Oakland, @shayegh2020staying provided visual evidence indicating geographic variation in that the fall in crime rates ac
+
+Secondly, existing studies suggest small number of drivers.
+
+Thirdly, equity argument.
+
+@shayegh2020staying; 
+@felson2020routine
+
 - Findings hold important practical and theoretical value for crime science research, and yet, less is known about the localized drivers of macro-level trends. 
@@
 
-Given the unique longitudinal trends, spatial patterning and opportunity structure of the clusters identified using the aggregate notifiable offences measure, we might expect the clusters identified to have distinct crime type profiles.
-That is, some crime types may be more or less prevalent in certain clusters to begin with and / or there may be variation in the extent to which lockdown impeded opportunities across crime types.
-To explore this.
-Figure 6 summarises the crime type profile of each cluster.
-
+Given the unique longitudinal trends, spatial patterning and opportunity structure of the clusters identified using the aggregate notifiable offences measure, we might expect the clusters identified to have distinct crime type profiles (see Figure 6).
 In February, before the restrictions on mobility and social interaction, the crime type profiles of each cluster were already distinct.

### 2021-03-05 draft_paper.Rmd
heat maps tidy

@@
 @weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local).
-Findings have contributed to the evidence-base around routine activities theory, reflecing the considerable variation in criminal opportunities across (even spatially proximate) places [@eck2000finding; @weisburd2012criminology].
+Findings have contributed to the evidence-base around routine activities theory, reflecing the considerable variation and concentration in criminal opportunities across (even spatially proximate) places [@eck2000finding; @weisburd2012criminology].
 
@@
 
+```{r heatmaps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Comparing clusters between equivalent months in 2020 and 2019 according to (a) absolute difference in counts, (b) relative percentage difference in counts, (c) average difference in LSOA counts.", fig.align='center'}
+knitr::include_graphics("visuals/heatmaps_gg.png")
+```
+
 ## Characteristics of clusters

### 2021-03-09 draft_paper.Rmd
heat maps and writing

--- 2021-03-05 draft_paper.Rmd
@@
 Studies include case study sites in the United States [@campedelli2020exploring; @gerell2020minor; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show] and Mexico [@estevez2020crime].
-Generally speaking, findings are consistent with opportunity perspectives and the mobility theory on the pandemic [@langton2021cs].
+Generally speaking, findings are consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@langton2021cs; @halford2020crime].
 That is, legally-enforced restrictions on people's mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime.
@@
 That said, little is known about the localized drivers of these macro-level trends.
-An exploration of the extent to which the macro lockdown crime drop has been experienced equitably across space is valuable for two principal reasons.
-First, we would expect only a small number of fine-grained geographic units to have driven the lockdown crime drop.
-In pre-pandemic era, longitudinal studies have consistently demonstrated that, in a crime drop scenario, most local areas remained remarkably stable, with a disproportionatly large volume of change attributable to just handful of meso or micro-level units (e.g.
-@weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local).
-Findings have contributed to the evidence-base around routine activities theory, reflecing the considerable variation in criminal opportunities across (even spatially proximate) places [@eck2000finding; @weisburd2012criminology].
+To what extent have local areas experienced the lockdown crime drop in unison?
+This exploration represents a unique test of opportunity perspectives on crime, which would stipulate that only a small number of local areas will have driven the lockdown crime drop.
+In the pre-pandemic era, longitudinal studies have consistently demonstrated that, in a macro-level crime drop scenario, most meso or micro-level areas remained remarkably stable, with a disproportionately large volume of change attributable to just handful of units [e.g.
+@weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local].
+This is credited to highly localized opportunity structures, which can differ considerably within the same city, and even within the same neighborhood [@eck2000finding; @weisburd2012criminology].
+Here, we expect that areas typically characterized by plentiful opportunities for crime, which quickly become unavailable following the imposition of lockdown measures, will drive the lockdown crime drop.
 
-Secondly,
+Preliminary evidence from the United States certainly suggests that macro lockdown crime trends might mask underlying variation.
+Using police-recorded crime data in San Francisco and Oakland, @shayegh2020staying provided visual descriptive evidence which indicated some degree of geographic variation in pre and post-lockdown crime.
+A study of Detroit census block groups found that, amidst a fall in burglaries following a stay at home order, there was a shift in concentrations away from residential areas towards mixed and non-residential parts of the city [@felson2020routine].
 
-from an opportunity and mobility perspective,
-
-preliminary evidence from the United States suggests that macro lockdown crime trends might mask underlying variation.
-Using police-recorded crime data in San Francisco and Oakland, @shayegh2020staying provided visual evidence indicating geographic variation in that the fall in crime rates ac
-
-Secondly, existing studies suggest small number of drivers.
-
-Thirdly, equity argument.
-
-@shayegh2020staying; 
-@felson2020routine
-
-- Findings hold important practical and theoretical value for crime science research, and yet, less is known about the localized drivers of macro-level trends. 
-- Questions around the drivers of macro-level crime drops have been of interest for some time, following the widespread fall in crime since the 1990s [@farrell2014crime; @weisburd2012criminology; @weisburd2004trajectories; @bannister2018local].
-- At the city level, numerous studies have deployed longitudinal clustering methods to unmask localized variation in the crime drop.
-Findings have consistently found that macro-level declines in crime over many years (decades) tend to be driven by a disproportionately small number of places, with most areas remaining fairly stable.
-- Here, we are presented with a unique scenario.
-Rather than a gradual decline in police-recorded crime over many decades, we have witnessed a rapid decline over a matter of weeks and months.
-And yet, comparable questions remain.
-- To what extent have local areas experienced this decline in unison?
-- Can we attribute most of the macro-level count trends to a handful of meso-level areas? 
-- Have the dividends of the lockdown crime drop been shared equitably across space [@akmed2021mon]?
-- Here, we use 7-months of police-recorded crime in England and Wales, aggregated to meso-level local units, to disentangle the nationwide lockdown crime drop.
 - We describe areas in terms of their opportunity structure, spatial patterning and crime type profile.
@@
 To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite].
-For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable (see Appendix).
+For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable.
+Potential cluster solution options were also examined using Principal Component Analysis to establish their suitability.
 
 We visualize the final cluster solution in a manner which conveys the underlying distribution of observations comprising each cluster at each time point, rather than reporting a summary statistic (e.g.
-the mean trajectory) in isolation, as has been in the case in existing research.
+the mean trajectory) in isolation.
 For each cluster, we also overlay the equivalent trajectories for 2018 and 2019 as a reference point for comparison to a 'typical' year.
@@
 We deploy non-parametric k-means clustering on LSOA (meso) level geographic units (N = 33,075) comprising England and Wales to decompose the macro-level 'decline and resurgence' observed between February and August, 2020.
-Figure 3 visualizes the resulting 6-cluster solution determined by the optimal Calinski-Criterion statistic [@calinski1974dendrite].
+Based on an assessment of the Calinski-Criterion statistic [@calinski1974dendrite] and Principal Component Analysis (see Appendix), we determined the optimal solution to be 6 clusters (see Figure 3).
 Solid black lines represent the median count for each cluster at any given time point, with the black dotted line showing the mean.
@@
 That said, most 'action' appears to be occurring amongst a small subset of LSOA.
-Clusters E and F collectively comprise only 0.33% (N = 110) of LSOA in England and Wales.
+Clusters E and F collectively comprise only 0.33% (N = 110).
 Yet, their crime counts are much higher, and the decline between March and April is considerable.
 We might therefore describe these clusters as 'high crime, major drop': LSOA with plentiful opportunities for crime in typical times, and in turn, LSOA which are most sensitive to the restriction in opportunities which followed after the imposition of lockdown.
-Interestingly, clusters E and F represent the only clusters which remained below pre-pandemic levels by August.
-In both cases, the average counts in February 2020 were _higher_ than those observed in previous years, but following the sharp decline and subsequent increase by August, these averages were _lower_ than 2018 and 2019.
+To further decompose these clusters, we now disentangle the contribution of each cluster.
 
@@
 
-To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the month-by-month change in counts.
-Alongside these raw counts, we also include the percentage of total absolute change (i.e.
+To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the monthly change in counts and percentage of total absolute change (i.e.
 positive or negative) attributable to each cluster.
-We can use this visual to identify which clusters drove the initial decline and subsequent resurgence nationwide.
+We can use this visual to identify which clusters drove the initial decline and subsequent nationwide resurgence in crime.
 
@@
 This is suggestive of highly localized change in opportunities as lockdown rules were relaxed: change which is aggregated away at the macro-level.
-Stable, low crime areas mitigated against further nationwide increases which may, if they had continued, actually resulted in a higher crime count in August than observed in previous years.
+Stable, low crime areas mitigated against further nationwide increases which may, if they had continued, resulted in a higher crime count in August than observed in previous years.
 
@@
 knitr::include_graphics("visuals/change_k6_gg.png")
+```
+
+These findings can be supplemented with a comparison to the previous year.
+Figure 5 compares the clusters identified in 2020 with the equivalent months in 2019.
+Here, it is clear that the 'high crime major drop' clusters (E and F) contributed the most to the lockdown crime drop relative to their size.
+By way of example, around 27,000 fewer crimes occurred in cluster A during April 2020 compared to April 2019 (Figure 5a).
+This translates to a 24% drop (Figure 5b), and an average fall per LSOA of 1.39 counts (Figure 5c).
+For cluster F, approximately 5,000 fewer crimes occurred (Figure 5a), which translates to an 87% drop (Figure 5b), and an average fall per LSOA of 543 counts (Figure 5c).
+In other words: whether comparing month-to-month change in 2020, or comparing equivalent months in 2019, the lockdown crime drop was driven by a small number of local areas, with most meso-level units remaining fairly stable, even amidst dramatic macro-level change.
+
+```{r heatmaps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Comparing clusters between equivalent months in 2020 and 2019 according to (a) absolute difference in counts, (b) percentage difference in counts, (c) average difference in counts.", fig.align='center'}
+knitr::include_graphics("visuals/heatmaps_gg.png")
 ```

### 2021-03-09 scripts/police_data_handling_cleaned.R

@@ -756,26 +756,29 @@ count_diff_gg <- ggplot(data = d2p, mapping = aes(x = month, y = Cluster, fill =
   # geom_text(mapping = aes(label = round(count_change, 2)), color = "white", size = 4) +
   scale_fill_viridis_c(direction = -1) +
   theme_bw() +
-  labs(x = NULL, y = NULL, fill = NULL) +
+  labs(x = NULL, y = NULL, fill = "Count   ") +
   theme(legend.position = "bottom",
         axis.text = element_text(size = 5),
         axis.title = element_text(size = 5),
         axis.ticks = element_line(size = 0.2),
         legend.key.height = unit(0.2, "cm"),
-        legend.text = element_text(size = 5))
+        legend.text = element_text(size = 5),
+        legend.title = element_text(size = 7))
 
 av_diff_gg <- ggplot(data = d2p, mapping = aes(x = month, y = Cluster, fill = av_change)) + 
   geom_tile() +
   # geom_text(mapping = aes(label = round(av_change, 2)), color = "white", size = 4) +
   scale_fill_viridis_c(direction = -1) +
   theme_bw() +
-  labs(x = NULL, y = NULL, fill = NULL) +
+  labs(x = NULL, y = NULL, fill = "Average \n count   ") +
   theme(legend.position = "bottom",
         axis.text = element_text(size = 5),
         axis.title = element_text(size = 5),
         axis.ticks = element_line(size = 0.2),
         legend.key.height = unit(0.2, "cm"),
-        legend.text = element_text(size = 5))
+        legend.text = element_text(size = 5),
+        legend.title = element_text(size = 7))
+
 
 
 # Data handling for percentage change between 2019 and 2020 by month and cluster.
@@ -793,13 +796,15 @@ perc_change_gg <- ggplot(data = d2p, mapping = aes(x = month, y = Cluster, fill 
   # geom_text(mapping = aes(label = round(percent_change, 2)), color = "white", size = 4) +
   scale_fill_viridis_c(direction = -1) +
   theme_bw() +
-  labs(x = NULL, y = NULL, fill = NULL) +
+  labs(x = NULL, y = NULL, fill = "%   ") +
   theme(legend.position = "bottom",
         axis.text = element_text(size = 5),
         axis.title = element_text(size = 5),
         axis.ticks = element_line(size = 0.2),
         legend.key.height = unit(0.2, "cm"),
[…]

### 2021-03-10 draft_paper.Rmd
background writing

@@
 ---
-title: 'Instability and inequality in the COVID-19 lockdown crime drop'
-output: pdf_document
+title: "Identifying local drivers of the COVID-19 lockdown crime drop in England and Wales"
+author:
+  - Samuel Langton
+  - Anthony Dixon
+  - Graham Farrell
+csl: apa.csl
+header-includes: \usepackage{float}
+output:
+  pdf_document: default
+  word_document:
+    reference_docx: reference_doc.docx
 bibliography: refs.bib
-csl: apa.csl
-header-includes:
-  - \usepackage{float}
 ---
@@
 ```{r setup, include=FALSE, echo = F}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, eval = T) 
 library(readr);library(kableExtra)
@@
 This exploration represents a unique test of opportunity perspectives on crime, which would stipulate that only a small number of local areas will have driven the lockdown crime drop.
-In the pre-pandemic era, longitudinal studies have consistently demonstrated that, in a macro-level crime drop scenario, most meso or micro-level areas remained remarkably stable, with a disproportionately large volume of change attributable to just handful of units [e.g.
-@weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local].
-This is credited to highly localized opportunity structures, which can differ considerably within the same city, and even within the same neighborhood [@eck2000finding; @weisburd2012criminology].
-Here, we expect that areas typically characterized by plentiful opportunities for crime, which quickly become unavailable following the imposition of lockdown measures, will drive the lockdown crime drop.
+In the pre-pandemic era, longitudinal studies have consistently demonstrated that, in a macro-level crime drop scenario, most meso or micro-level areas remained remarkably stable.
+Instead, a disproportionately large volume of the macro-level change is attributable to a small number of units [e.g.
+@weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local; @vandeviver2019stability].
+This finding can be credited to highly localized opportunity structures, which can differ considerably within the same city, and even within the same neighborhood [@eck2000finding; @weisburd2012criminology].
+A shift in these local structures, brought about either through intervention (e.g.
+hotspot policing) or rapid changes in routine activities [@griffiths2004communities] can bring about wider (macro) change in crime rates.
+
+In a lockdown scenario, we might expect these effects to be both exaggerated (in scale) and instantaneous (in time).
+For instance, public transport hubs typically act as major crime generators due to the vast congregation of ambient populations in time and space facilitating the convergence of motivated offenders and suitable targets [@newton2018macro].
+In lockdown, public transport services were either closed completely, or open only for a limited number of essential purposes.
+With the necessary convergence of offenders and victims disrupted, often quite literally overnight, we would expect these customarily high crime places to fall considerably, driving the macro-level trend.
+By contrast, areas typically devoid of criminal opportunity, with say, little or no ambient populations, may remain largely unaffected by lockdown restrictions on mobility, and in turn, contribute little to any macro-level change.
+In other words, while lockdowns are imposed at a city or national level, their effect on crime will likely be moderated by the opportunity structure of local areas.
 
 Preliminary evidence from the United States certainly suggests that macro lockdown crime trends might mask underlying variation.
-Using police-recorded crime data in San Francisco and Oakland, @shayegh2020staying provided visual descriptive evidence which indicated some degree of geographic variation in pre and post-lockdown crime.
-A study of Detroit census block groups found that, amidst a fall in burglaries following a stay at home order, there was a shift in concentrations away from residential areas towards mixed and non-residential parts of the city [@felson2020routine].
+Using police-recorded crime data in San Francisco and Oakland, @shayegh2020staying provided visual descriptive evidence which indicated a degree of geographic variation in pre and post-lockdown crime.
+In Detroit, a study using a small sample of block units found that, amidst a fall in burglaries following a stay at home order, there was a shift in concentrations away from residential areas towards mixed and non-residential parts of the city [@felson2020routine].
+In Chicago, there is evidence of variability in the extent to which lockdown policies impacted upon crime.
+A small proportion of communities drove the citywide decline, with most areas remaining largely unchanged, and some areas even _increasing_, bucking the trend entirely [@campedelli2020disentangling].
+Using regional units of analysis nested within the state of Queensland, Australia, @payneregional2021 found a degree of diversity in crime rate trends, suggesting that the lockdown crime drop was not 'universal'.
+In the United Kingdom, however, we have not seen a comprehensive decomposition of local variation and longitudinal (in)stability underpinning the lockdown crime drop and subsequent resurgence, or indeed an exploration of the opportunity structures characterizing the local areas which have driven macro-level changes.
 
-- We describe areas in terms of their opportunity structure, spatial patterning and crime type profile.
+In this paper, we aim to identify the localized drivers of the lockdown crime drop in England and Wales.
+We achieve this using 7-months of police-recorded crime data between February and August, 2020.
+First, we summarize the nationwide (macro) trend in terms of absolute crime counts and concentration.
+Second, using a non-parametric longitudinal clustering technique, we identify meaningful clusters of meso-level areas which unpick stable (and volatile) local areas underpinning the macro-level trend.
+Third, we describe the clusters identified in terms of their spatial patterning, opportunity structures, and crime type profile.
 
@@
 Each of these are now outlined in turn, followed by an outline of the methods deployed.
-Code to replicate the data downloads, handling, analyses and visualization reported here is openly available (https://github.com/langtonhugh/covid_spatial).
+Code to replicate the data downloads, handling, analyses and visualization reported here are openly available (https://github.com/langtonhugh/covid_spatial).
 
@@
 Here, it is clear that the 'high crime major drop' clusters (E and F) contributed the most to the lockdown crime drop relative to their size.
-By way of example, around 27,000 fewer crimes occurred in cluster A during April 2020 compared to April 2019 (Figure 5a).
+By way of example, around 27,000 fewer crimes occurred in cluster A--containing 19,162 LSOA--during April 2020 compared to April 2019 (Figure 5a).
 This translates to a 24% drop (Figure 5b), and an average fall per LSOA of 1.39 counts (Figure 5c).
-For cluster F, approximately 5,000 fewer crimes occurred (Figure 5a), which translates to an 87% drop (Figure 5b), and an average fall per LSOA of 543 counts (Figure 5c).
-In other words: whether comparing month-to-month change in 2020, or comparing equivalent months in 2019, the lockdown crime drop was driven by a small number of local areas, with most meso-level units remaining fairly stable, even amidst dramatic macro-level change.
+For cluster F, containing just 9 LSOA, approximately 5,000 fewer crimes occurred (Figure 5a), which translates to an 87% drop (Figure 5b), and an average fall per LSOA of 543 counts (Figure 5c).
+In other words: whether comparing month-to-month change in 2020, or comparing equivalent months in 2019, the lockdown crime drop was disproportionately driven by a small number of local areas, with most meso-level units remaining fairly stable, even amidst dramatic macro-level change.
 
-```{r heatmaps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Comparing clusters between equivalent months in 2020 and 2019 according to (a) absolute difference in counts, (b) percentage difference in counts, (c) average difference in counts.", fig.align='center'}
+```{r heatmaps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Comparing LSOA clusters between equivalent months in 2020 and 2019 according to (a) absolute difference in counts, (b) percentage difference in counts, (c) average difference in counts.", fig.align='center'}
 knitr::include_graphics("visuals/heatmaps_gg.png")
@@
 Without exception, the city centres and commercial districts of each city are characterised by 'high crime major drop' clusters E and F, which in turn often neighbor the 'mid crime mid drop' cluster D.
-By contrast, the 'low crime and stable' clusters tend to sit outside of the city centres: cluster B in the suburbs and the sparsely populated cluster A LSOA on the periphery of urban conurbations.
-For a detailed investigation of the spatial clusters identified for the whole of England and Wales, we refer readers to an openly available online map (anonymized link here).
+By contrast, the 'low crime and stable' clusters tend to sit outside of the city centres: cluster B in the suburbs and the sparsely populated cluster A on the periphery of urban conurbations.
+For a detailed and interactive investigation of the spatial clusters identified for the whole of England and Wales, we refer readers to an openly available online map (anonymized link here).
 

### 2021-03-11 draft_paper.Rmd
discussion start

@@
 
-- We have investigated the degree of stability and localised inequality in the lockdown crime drop between February and August in England and Wales using police-recorded crime data aggregated to local areas (LSOA).
-- Findings are largely consistent with broader findings around inequalities in the crime drop and the notion that a small number of fine-grained geographic areas tend to drive nationwide trends [@andresen2017trajectories; @weisburd2012criminology; @mcvie2020increasing].
+This paper has aimed to investigate the localized drivers of the COVID-19 crime drop in England and Wales.
+Using 7-months of police-recorded data on notifiable offences (excluding drugs), we decomposed the (macro) nationwide decline in crime according to clusters of (meso) local areas, each with distinct longitudinal trajectories.
+Findings provide substantial evidence to suggest that the nationwide decline in crime during lockdown has been driven by a disproportionately small number of localized geographic areas.
+We find that these localized drivers, dominating city centers, had pre-existing high levels of crime attributable to their unique opportunity structures, containing numerous nightlife facilities, commercial shops, public transport nodes and bicycle parking spaces.
+Clusters demonstrate a degree of stability in their crime type profiles, but the 'high crime major drop' areas appear to be weighted towards theft and shoplifting.
+These areas experienced a fairly slow resurgence upon the relaxation of lockdown rules, remaining below historical levels even by the time low and mid-crime clusters had returned to normality.
+
+The finding that a small number of fine-grained geographic units drove the macro-level crime drop, with most areas otherwise remaining fairly stable, is consistent with results from other study sites.
+Evidence from both Australia [@payneregional2021] and the United States [@campedelli2020disentangling] suggests that the lockdown crime drop was not uniform, but instead, the macro-level trend masked underlying variation.
+
+Here, we have quantified and described this localized variation for both the decline and resurgence stages of the pandemic.
+
+is also consistent with findings from the pre-pandemic era, in which
+
+we would argue that this study has extended these existing findings in three key areas. Firstly, we have examined both the lock
+
+The non-parametric longitudinal clustering technique deployed here has permitted, for the first time, a comprehensive evaluation of this using fine-grained spatial scales in England and Wales.
+
+ [@andresen2017trajectories; @weisburd2012criminology; @mcvie2020increasing].
 - Specifically, we find that most LSOA remained remarkably stable during the initial, major lockdown crime drop between March and April.

### 2021-03-11 draft_paper.Rmd
discussion

--- 2021-03-10 draft_paper.Rmd
@@
 Findings have tended to align with expectations from opportunity perspectives on crime.
-Lockdown mobility patterns disrupted the typical convergence of motivated offenders, suitable targets and capable guardians, reducing opportunities for victimization [@stickle2020crime].
+Lockdown mobility patterns appear to have disrupted the typical convergence of motivated offenders, suitable targets and capable guardians, reducing opportunities for victimization [@stickle2020crime].
 As lockdown restrictions were slowly repealed, routine activities converged back to pre-pandemic levels and we witnessed a resurgence in criminal activity [@langton2021cs].
 
-While existing studies have provided insight into the opportunity structures of crime, and the nuances of using police data to study lockdown as a natural experiment, research has almost exclusively been undertaken using macro-level units of analysis, such as cities or countries.
+While existing studies have provided insight into lockdown crime and mobility patterns, research has almost exclusively been undertaken using macro-level units of analysis, such as cities or countries.
 Less is known about the localized drivers of the lockdown crime drop or the degree to which macro-level trends are masking geographic inequalities in victimization.
-Pre-pandemic research examining the (in)stability of crime at fine-grained spatial scales [@weisburd2012criminology; @curman2015crime] and inequalities in the crime drop [@mcvie2020increasing; @ignatans2016whom; @akmed2021mon] would certainly suggest that local areas are unlikely to have experienced the macro-level lockdown crime trends in unison.
-Rather, we might expect specific places, typically associated with high ambient populations and plentiful opportunities for crime, to have driven the wider trend, with most localized areas remaining fairly stable.
+Pre-pandemic research examining the (in)stability of crime at fine-grained spatial scales [@weisburd2012criminology; @curman2015crime] and inequalities in the crime drop [@mcvie2020increasing; @ignatans2016whom; @akmed2021mon] would certainly suggest that local areas are unlikely to have experienced lockdown crime trends in unison.
+Rather, we might expect specific places, typically associated with high ambient populations and plentiful opportunities for crime, to have driven the wider trend, with most local areas remaining fairly stable.
 
-Examining the localized drivers of the lockdown crime drop represents the primary motivation of this paper.
-We disentangle the macro-level drop in police-recorded crime observed in England and Wales between February and August 2020.
+Examining the localized drivers of the lockdown crime drop (and subsequent resurgence) represents the primary motivation of this paper.
+We decompose the macro-level trend in police-recorded crime observed in England and Wales between February and August 2020.
 We deploy non-parametric longitudinal clustering to identify clusters of meso-level units which contributed disproportionately to the nationwide drop and subsequent resurgence in crime during the nationwide lockdown.
-The spatial patterning, opportunity structure and crime type profile of these local areas are quantified and summarized for their consistency with expectations for opportunity and mobility theories of crime [@halford2020crime].
+The spatial patterning, opportunity structure and crime type profile of these local areas are quantified and summarized for their consistency with expectations from opportunity and mobility theories of crime [@halford2020crime].
 
@@
 
-- We have investigated the degree of stability and localised inequality in the lockdown crime drop between February and August in England and Wales using police-recorded crime data aggregated to local areas (LSOA).
-- Findings are largely consistent with broader findings around inequalities in the crime drop and the notion that a small number of fine-grained geographic areas tend to drive nationwide trends [@andresen2017trajectories; @weisburd2012criminology; @mcvie2020increasing].
-- Specifically, we find that most LSOA remained remarkably stable during the initial, major lockdown crime drop between March and April.
-- The subsequent resurgence among low and mid-crime areas was limited by 'high crime' areas stabilizing into May.
-- But, the resurgence of high crime areas was lagged, and began towards the end of spring.
-- Further macro-level increases in crime at the end of summer were limited by the low crime and stable areas, mitigating the impact of increases in high crime clusters.
+This paper has aimed to investigate the localized drivers of the COVID-19 crime drop in England and Wales.
+Using 7-months of police-recorded data on notifiable offences (excluding drugs), we decomposed the (macro) nationwide decline in crime according to clusters of (meso) local areas, each with distinct longitudinal trajectories.
+Findings provide substantial evidence to suggest that the nationwide decline in crime during lockdown has been driven by a disproportionately small number of localized geographic areas.
+We find that these localized drivers, dominating city centers, had pre-existing high levels of crime, likely attributable to their unique opportunity structures, containing numerous nightlife facilities, commercial shops, public transport nodes and bicycle parking spaces.
+Clusters demonstrate a degree of stability in their crime type profiles, but the 'high crime major drop' areas appear to be weighted towards theft and shoplifting.
+Despite their dramatic declines, these areas experienced a fairly slow resurgence upon the relaxation of lockdown rules, remaining below historical levels even by the time low and mid-crime clusters had returned to normality.
+
+The finding that a small number of fine-grained geographic units drove macro-level crime trends, with most areas otherwise remaining fairly stable, is consistent with results from other study sites.
+Evidence from Queensland, Australia [@payneregional2021] and the Chicago, United States [@campedelli2020disentangling] certainly suggests that the lockdown crime drop was not uniform, but instead, the macro-level trend masked underlying variation.
+In the pre-pandemic era, studies have consistently found that local areas (e.g.
+street segments, neighborhoods) do not experience macro-level trends in unison, but rather, change tends to be attributable to a handful of areas [@andresen2017trajectories; @weisburd2012criminology].
+Here, for the first time, we have exploited the quasi-'natural experiment' conditions of the pandemic to decompose this variation during both the _decline_ and _resurgence_ stages of a macro-level trend.
+Interestingly, the degree of concentration, as measured by the generalized Gini coefficient, remained remarkably stable, even amidst this macro and meso-level change.
+This suggests that the law-like behavior of crime concentration [@weisburd2015law] holds even in this highly volatile and unusual unusual scenario.
+
+The opportunity characteristics of the clusters identified certainly warrants further discussion. We found that
+
 - The LSOA which drove the lockdown crime drop tended to have considerable crime attractors and generators, such as public transport nodes, an active nightlife, and plenty of shops and bicycle parking.

### 2021-03-17 scripts/police_data_handling_cleaned.R
edits

@@ -866,11 +866,13 @@ liver_sf <- lsoa_ew_valid_sf %>%
 cardiff_sf <- lsoa_ew_valid_sf %>% 
   filter(str_detect(string = lsoa_ew_valid_sf$geo_label, pattern = "Cardiff"))
 
-# Visual inspection tells us that Cardiff has an island LSOA to the south. This is Flat Holne Island - we
-# remove this for the visual, and because its policing/opportunity structure will perform independently 
-# to Cardiff. LSOA code identified interactively in QGIS.
-cardiff_sf <- cardiff_sf %>% 
+# Visual inspection tells us that Cardiff has an island LSOA to the south. This includes Flat Holm Island - we
+# remove this for the visual. LSOA code identified interactively in QGIS. Crude but for visual it's okay.
+cardiff_sf <- cardiff_sf %>%
   filter(lsoa_code != "W01001940")
+
+# Remove Flat Holne Island from Cardiff.
+# cardiff_sf <- rmapshaper::ms_filter_islands(input = cardiff_sf, min_area = 10000)
 
 # Create list.
 cities_list <- list(birm_sf, leeds_sf, sheff_sf, brad_sf, liver_sf, cardiff_sf)

### 2021-03-17 scripts/police_data_handling_cleaned_drugs.R
revert

@@ -1,0 +1,314 @@
+# Load packages.
+library(kml)
+library(cowplot)
+library(lorenzgini)
+library(spdep)
+library(readr)
+library(janitor)
+library(dplyr)
+library(tidyr)
+library(purrr)
+library(stringr)
+library(forcats)
+library(ggplot2)
+library(sf)
+library(factoextra)
+
+# Useful function.
+`%nin%` <- Negate(`%in%`)
+
+
+# Archive data downloaded January 2012, covering the 3-year period up to and including November 2020.
+# if(!file.exists('data')){dir.create("data")}
+# download.file(url = "https://data.police.uk/data/archive/2020-11.zip", destfile = "data/archive2020-11.zip")
+
+# Unzip.
+# unzip(zipfile = "data/archive2020-11.zip", exdir = "data")
+
+# Note the loading and handling data is largely determined by available memory. Many people may be able to
+# run things more efficiently (and in a more straightforward order) with a more powerful computer.
+
+# List all those 'street' files (rather than outcomes or stop and search).
+list_2020 <- paste("data/", list.files("data", pattern = glob2rx("2020*street.csv"),  recursive=TRUE), sep = "")
+list_2019 <- paste("data/", list.files("data", pattern = glob2rx("2019*street.csv"),  recursive=TRUE), sep = "")
+
+# Add December 2020 manually to avoid conflicts with the main paper.
+wd <- "C:/Users/langt/OneDrive - University of Leeds/Lenovo/Leeds_covid/spatial/dec_2020/"
+list_dec_2020 <- paste(wd, list.files(wd, pattern = glob2rx("2020*street.csv"),  recursive=TRUE), sep = "")
+
+# Read in .csv files for each year.
+data_2020     <- lapply(list_2020    , read_csv)
+data_dec_2020 <- lapply(list_dec_2020, read_csv)
+data_2019     <- lapply(list_2019    , read_csv)
+
+# Bind each in to data frames. For now, we keep the years separate.
+full_data_2020 <- data_2020 %>% 
+  bind_rows() %>% 
+  clean_names() %>% 
+  mutate(year = "2020") %>% 
+  filter(crime_type == "Burglary")
+
+full_dec_data_2020 <- data_dec_2020 %>% 
+  bind_rows() %>% 
+  clean_names() %>% 
+  mutate(year = "2020") %>% 
[…]

### 2021-03-18 draft_paper.Rmd
add maps city centre

@@
 ---
-title: "Identifying local drivers of the COVID-19 lockdown crime drop in England and Wales"
+title: "Local drivers of the COVID-19 lockdown crime drop in England and Wales"
 author:
@@
 
-While existing studies have provided insight into lockdown crime and mobility patterns, research has almost exclusively been undertaken using macro-level units of analysis, such as cities or countries.
+While existing studies have provided insight into the impact of lockdown on crime, research has almost exclusively been undertaken using macro-level units of analysis, such as cities or countries.
 Less is known about the localized drivers of the lockdown crime drop or the degree to which macro-level trends are masking geographic inequalities in victimization.
-Pre-pandemic research examining the (in)stability of crime at fine-grained spatial scales [@weisburd2012criminology; @curman2015crime] and inequalities in the crime drop [@mcvie2020increasing; @ignatans2016whom; @akmed2021mon] would certainly suggest that local areas are unlikely to have experienced lockdown crime trends in unison.
+Pre-pandemic studies examining the (in)stability of crime at fine-grained spatial scales [@weisburd2012criminology; @curman2015crime] and inequalities in the crime drop [@mcvie2020increasing; @ignatans2016whom; @akmed2021mon] would certainly suggest that local areas are unlikely to have experienced lockdown crime trends in unison.
 Rather, we might expect specific places, typically associated with high ambient populations and plentiful opportunities for crime, to have driven the wider trend, with most local areas remaining fairly stable.
@@
 
-Recent months have seen a wave of interest in describing and explaining changes in crime in response to restrictions on mobility brought about to limit the spread of COVID-19.
-Studies include case study sites in the United States [@campedelli2020exploring; @gerell2020minor; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show] and Mexico [@estevez2020crime].
-Generally speaking, findings are consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@langton2021cs; @halford2020crime].
-That is, legally-enforced restrictions on people's mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime.
-As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often to levels that would have been expected without the global pandemic and ensuing lockdowns.
+Recent months have seen a wave of interest in describing and explaining how crime has changed in response to restrictions on mobility brought about to limit the spread of COVID-19.
+Studies have been undertaken in the United States [@campedelli2020exploring; @gerell2020minor; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show] and Mexico [@estevez2020crime].
+Generally speaking, findings have been consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@langton2021cs; @halford2020crime].
+That is, legally-enforced restrictions on daily activities, mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime.
+As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often back to levels that would have been expected without the global pandemic and ensuing lockdowns.
 
 That said, little is known about the localized drivers of these macro-level trends.
-To what extent have local areas experienced the lockdown crime drop in unison?
-This exploration represents a unique test of opportunity perspectives on crime, which would stipulate that only a small number of local areas will have driven the lockdown crime drop.
-In the pre-pandemic era, longitudinal studies have consistently demonstrated that, in a macro-level crime drop scenario, most meso or micro-level areas remained remarkably stable.
+To what extent have local areas experienced lockdown trends in unison?
+This exploration represents a unique test of opportunity perspectives on crime, which would stipulate, for instance, that only a small number of local areas will have driven the lockdown crime drop.
+Prior to the COVID-19 pandemic, longitudinal studies have consistently demonstrated that, in a macro-level crime drop scenario, most meso or micro-level areas remained remarkably stable.
 Instead, a disproportionately large volume of the macro-level change is attributable to a small number of units [e.g.
-@weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local; @vandeviver2019stability].
+@weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local; @andresen2017trajectories].
 This finding can be credited to highly localized opportunity structures, which can differ considerably within the same city, and even within the same neighborhood [@eck2000finding; @weisburd2012criminology].
 A shift in these local structures, brought about either through intervention (e.g.
-hotspot policing) or rapid changes in routine activities [@griffiths2004communities] can bring about wider (macro) change in crime rates.
+hotspot policing) or rapid changes in routine activities [@griffiths2004communities] can bring about wider (macro) change in crime rates, even if most units nested within the macro region remain stable.
 
@@
 In lockdown, public transport services were either closed completely, or open only for a limited number of essential purposes.
-With the necessary convergence of offenders and victims disrupted, often quite literally overnight, we would expect these customarily high crime places to fall considerably, driving the macro-level trend.
+With the necessary convergence of offenders and victims disrupted, often quite literally overnight, we would expect these customarily high crime places to fall considerably, prompting a macro-level decline.
 By contrast, areas typically devoid of criminal opportunity, with say, little or no ambient populations, may remain largely unaffected by lockdown restrictions on mobility, and in turn, contribute little to any macro-level change.
-In other words, while lockdowns are imposed at a city or national level, their effect on crime will likely be moderated by the opportunity structure of local areas.
+In other words, while lockdowns have often been imposed equitably at a city or national level, the effect on crime will likely be moderated by the opportunity structure of local areas.
 
-Preliminary evidence from the United States certainly suggests that macro lockdown crime trends might mask underlying variation.
+Preliminary evidence from the United States certainly suggests that this may be the case.
 Using police-recorded crime data in San Francisco and Oakland, @shayegh2020staying provided visual descriptive evidence which indicated a degree of geographic variation in pre and post-lockdown crime.
-In Detroit, a study using a small sample of block units found that, amidst a fall in burglaries following a stay at home order, there was a shift in concentrations away from residential areas towards mixed and non-residential parts of the city [@felson2020routine].
-In Chicago, there is evidence of variability in the extent to which lockdown policies impacted upon crime.
-A small proportion of communities drove the citywide decline, with most areas remaining largely unchanged, and some areas even _increasing_, bucking the trend entirely [@campedelli2020disentangling].
+In Detroit, a study using a small sample of block units found that, amidst a fall in burglaries following stay at home orders, there was a shift in concentrations away from residential areas towards mixed and non-residential parts of the city [@felson2020routine].
+In Chicago, there was evidence of variability in the extent to which lockdown policies impacted upon crime.
+A small proportion of communities drove the citywide decline, with most areas remaining largely unchanged, and some areas even _increasing_, bucking the macro-level trend entirely [@campedelli2020disentangling].
 Using regional units of analysis nested within the state of Queensland, Australia, @payneregional2021 found a degree of diversity in crime rate trends, suggesting that the lockdown crime drop was not 'universal'.
@@
 
-In this paper, we aim to identify the localized drivers of the lockdown crime drop in England and Wales.
+In this paper, we aim to identify and describe the localized drivers of the lockdown crime drop in England and Wales.
 We achieve this using 7-months of police-recorded crime data between February and August, 2020.
-First, we summarize the nationwide (macro) trend in terms of absolute crime counts and concentration.
+First, we summarize the nationwide (macro) trend in terms of absolute crime counts in comparison to previous years.
 Second, using a non-parametric longitudinal clustering technique, we identify meaningful clusters of meso-level areas which unpick stable (and volatile) local areas underpinning the macro-level trend.
@@
 Individual records are time-stamped by month -- the temporal scale of this study.
-Due to ongoing data issues with Greater Manchester Police, we excluded data from this region, and therefore collated data from 42 out of 43 forces in England and Wales.
+Due to incomplete data releases from Greater Manchester Police, we excluded data from the Greater Manchester region, collating data from 42 out of 43 forces in England and Wales.
 
@@
 burglary, violence and sexual offences, theft from the person, vehicle crime).^[Further information on these crime categories can be obtained from https://www.police.uk/pu/contact-the-police/what-and-how-to-report/what-report/.] Records also include anti-social behaviour (ASB) which usually captures less serious offences such as nuisance behavior and is not a notifiable offence.
-Individual records were aggregated to create a count measure for notifiable offences (excluding drugs) by month either at the nationwide (macro) level, or the localized (meso) level, as detailed in the next section.
-There widespread recognition that drug crime trends, particularly during the COVID-19 lockdown, largely reflect policing proactivity rather than meaningful shifts in criminal behavior [@langton2021cs].
-Recognizing that aggregating data across crime types can mask variation [@andresen2017trajectories] we later decompose our main findings according to the twelve remaining notifiable offences, and refer to additional analyses in the Appendix.
+Individual records were aggregated to create a count measure for 'notifiable offences (excluding drugs)' by month at the nationwide (macro) level, and the localized (meso) level, as detailed in the next section.
+The decision to exclude drug offences follows widespread recognition that drug crime trends, particularly during the COVID-19 lockdown, largely reflect policing proactivity rather than meaningful shifts in criminal behavior [@langton2020open].
+Recognizing that aggregating data across crime types can mask variation [@andresen2017trajectories] we later decompose our main findings according to the twelve remaining notifiable offences, and report additional analyses broken down by crime type in the Appendix.
 
@@
 
-For nationwide (macro) descriptive statistics, which set the background for the main analyses, we use the geographic region of England and Wales, noting the exclusion of Greater Manchester.
+To provide national context to the main analysis, we use the (macro) geographic region of England and Wales, noting the exclusion of Greater Manchester.
 For localized (meso) analysis, we aggregate offences to Lower Super Output Area (LSOA).
@@
 In 2019, LSOA housed an average of 1,700 people.
-Data obtained from the open police data portal (see previous section) includes a pre-assigned field stating the LSOA in which the crime occurred, as recorded by the police.
+Data obtained from the open police portal (see previous section) include a pre-assigned field stating the LSOA in which the crime occurred as recorded by the police.
 Due to the spatial anonymization method used prior to data release, LSOA are the lowest level of aggregation at which we can reasonably assume spatial accuracy across multiple crime types [@tompson2015uk].
-Boundaries reflect 'on the ground' characteristics such as roads and railways [@onslsoachange], and as such, we consider LSOA to represent meaningful behavioral spaces to study the opportunity structures of crime.
-After removing crimes recorded by Greater Manchester Police, crimes recorded to have occurred within the Greater Manchester region, and four LSOAs which contained no crime between 2018 and 2020 (likely due to the spatial anonymization process), our final sample population for the meso-level analysis comprised 33,075 LSOA.
+LSOA boundaries reflect 'on the ground' characteristics such as roads and railways [@onslsoachange], and as such, we consider LSOA to represent meaningful behavioral spaces to study the opportunity structures of crime.
+After removing crimes recorded by Greater Manchester Police, crimes recorded to have occurred within the Greater Manchester region, and four LSOA which contained no crime between 2018 and 2020 (likely due to the spatial anonymization process), our final sample for the meso-level analysis comprised 33,075 LSOA.
 
@@
 Open Street Map is a crowdsourced geospatial database containing a vast array of features which can be used for explaining the temporal and spatial patterning of crime [@langton2020open].
-Geographic features are identified by pairs of keys and values which can be used to computationally query the API for data.
+Geographic features are identified by pairs of keys and values which can be used to computationally query the API for geospatial data.
 Based on existing research examining the opportunity structures of fine-grained spatial scales, we collated the coordinate locations of the following facilities:
@@
 nightlife, shops, public transport, bicycle parking).
-We hypothesize that LSOAs containing a high number of facility counts across each domain will have higher pre-pandemic levels of crime, due to the plentiful opportunities for criminality, and in turn, steeper declines in crime following lockdown due to such opportunities suddenly becoming unavailable.
-We expect areas with _low_ counts across these domains to have similarly low crime levels pre-lockdown, and thus will remain low and stable following lockdown commencement.
+We expect that LSOA containing a high number of facility counts across each domain will have higher pre-pandemic levels of crime, due to the plentiful opportunities for criminality, and in turn, steeper declines in crime following lockdown as a result of these opportunities suddenly becoming unavailable.
+We expect areas with low counts across these domains to have similarly low crime levels pre-lockdown, and thus will remain low and stable following lockdown commencement.
 
@@
 Analyses to examine the localized variation in the lockdown crime drop are conducted in three principal stages.
-First, an overview of nationwide (macro) trend is provided in terms of both absolute counts and the degree of concentration.
+First, an overview of the nationwide (macro) trend is provided in terms of absolute counts.
 Second, the macro-level trend is disentangled using non-parametric clustering techniques on the LSOA (meso) units (N = 33,075).
-Third, the characteristics of each cluster will be summarized in terms of their opportunity structures, spatial patterning and crime type profile.
+Third, the characteristics of each cluster are summarized in terms of their opportunity structures, spatial patterning and crime type profile.
+Each of these steps is now outlined in turn.
 
@@
 
-Macro-level descriptives of count trends and concentration for notifiable offences (excluding drugs) are reported for provide context from which we will subsequently unmask local (meso) variation.
-First, we visualize observed counts between February and August 2020 relative to the same periods in 2018 and 2019.
+Macro-level descriptives of count trends notifiable offences (excluding drugs) are reported for provide context from which we will subsequently unmask local (meso) variation.
+We visualize observed counts between February and August 2020 relative to the same periods in 2018 and 2019.
 In doing so, we can observe how nationwide crime trends changed in the face of lockdown measures [see also @langton2021cs].
-Second, we visualize generalized Gini coefficients [@bernasco2017more] to describe the degree of which offences became more (or less) concentrated during lockdown at LSOA level.
-Again, we report figures for 2018 and 2019 as a point of reference.
-We use the generalized version of the Gini coefficient to account for data sparsity, as for some crime types reported in the Appendix, the number of crimes falls below the number of LSOAs in the study region.
-Measures of counts and concentration set scene from which we can disentangle the underlying, meso-level variation.
+This sets the scene from which we can disentangle the underlying meso-level variation.
 
@@
 
-To quantify the degree of meso-level uniformity underpinning the macro-level trend, and identify the potential drivers of the lockdown crime drop, we deploy a longitudinal variant of k-means clustering [@genolini2011kml; @genolini2015kml].
+To quantify the degree of meso-level uniformity underpinning the macro-level trend, and identify the potential drivers of the lockdown crime drop (and resurgence), we deploy a longitudinal variant of k-means clustering [@genolini2011kml; @genolini2015kml].
 This non-parametric clustering technique has an established role in crime and place research for examining the longitudinal trajectories of local areas in a macro-level crime drop scenario [@curman2015crime; @andresen2017trajectories].
@@
 increasing, decreasing or stable) of clusters, rather than short-term volatility [@andresen2017trajectories].
-And yet, one key strength of k-means is that it is capable of identifying short-term fluctuation in longitudinal trends [@akmed2021mon].
-In this lockdown scenario, the availability of criminal opportunities were withdrawn quite literally overnight, and thus we might expect a similarly rapid and short-term change in crime at fine-grained spatial scales.
-This sets k-means apart from equivalent model-based methods restricted by polynomial terms, such as group-based trajectory modelling [@weisburd2004trajectories] and non-parametric techniques designed for long-term rather than short-term change [@akmed2021mon].
+To this end, it has demonstrated comparable value to model-based techniques such as group-based trajectory modelling [@curman2015crime].
+But, one key strength of k-means is that it is also capable of identifying short-term fluctuation in longitudinal trends [@akmed2021mon].
+In the lockdown scenario, criminal opportunities were withdrawn quite literally overnight, and thus we might expect a similarly rapid and short-term change in crime at meso spatial scales.
+The ability to unpick these rapid changes represents a key strength of k-means over non-parametric techniques such as anchored k-medoids, which are designed for long-term rather than short-term change [@akmed2021mon] or group-based trajectory modelling, which is limited by polynomial terms [@griffiths2004communities].
 
-We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOA in the study region (N = 33,075) between February and August 2020.
-To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite].
+We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offence (excluding drug) counts for LSOA in England and Wales (N = 33,075) between February and August 2020.
+To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite].
 For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable.
@@
 ```
-
-Figure 3 visualizes the overall level of concentration before and after lockdown, and in comparison to previous years, using the generalized Gini coefficient.
-Here, it is clear that notifiable offences, even amidst volatile and sharply changing levels in absolute counts (see Figure 2), experienced remarkably stable levels of concentration.
-There is a marginal dip in concentration during April, the first month of lockdown.
-This suggests that as crimes fell, those were occurring were _less_ concentrated.
-That is, crimes occurring during the height of lockdown were more spread out that we might have otherwise expected.
-That said, concentration quickly returned to normal levels.
-
-```{r nogini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Monthly generalized Gini cofficient trends for notitifable offences (excluding drugs) in England and Wales during the lockdown period.", fig.align='center'}
-knitr::include_graphics("visuals/no_gini_gg.png")
-```
-
-Based on the dramatic decline and subsequent resurgence of notifiable offences during the study period, we might expect some localized variability.
-Measures of concentration indicate that there has been stability at the 'global' level, but the evidence-base surrounding the law of crime concentration and micro-place (in)stability would certainly give reason to suspect that this masks underlying volatility [@weisburd2015law; @andresen2017trajectories].
-With this in mind, we now report the localized, meso-level analysis which unpicks the macro-level trend.
 
@@
 The provide local context to the cluster solutions identified through the k-means clustering, we visualize the spatial patterning of groupings for six major cities in England and Wales (see Figure 5).
-We find distinct geographic pattern to the clusters identified.
+We find distinct geographic patterns to the clusters identified.
+Maps denote a single 'city center' based on the LSOA with the highest count of Open Street Map features (i.e.
+the sum of nightlife, shops, public transport and bicycle parking).
 Here, it is worth noting that the Bradford region contains the city of Bradford, but also the satellite town of Keighley to the north west.
-The cluster map thus reflects these two urban centers.
 Without exception, the city centres and commercial districts of each city are characterised by 'high crime major drop' clusters E and F, which in turn often neighbor the 'mid crime mid drop' cluster D.
@@
 
-```{r  maps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Spatial distribution of clusters by major urban conurbation.", out.width= "100%", fig.align='center'}
+```{r  maps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Spatial distribution of clusters by major urban conurbation.
+LSOA with the highest count of Open Street Map features denotes city centre.", out.width= "100%", fig.align='center'}
 knitr::include_graphics("visuals/maps_gg.png")
@@
 
-We have now identified that, amidst dramatic nationwide change in notifiable offences during lockdown, and stability in overall crime concentrations, there has been considerable underlying volatility, with a small number of local areas driving the macro-level trend.
+We have now identified that, amidst dramatic nationwide change in notifiable offences during lockdown, there has been considerable underlying volatility, with a small number of local areas driving the macro-level trend.
 From a theoretical perspective, we expect the clusters to have differing opportunity structures.
@@
 
+The finding that a small number of fine-grained geographic units drove macro-level crime trends, with most areas otherwise remaining fairly stable, is consistent with results from other study sites.
+Evidence from Queensland, Australia [@payneregional2021] and the Chicago, United States [@campedelli2020disentangling] certainly suggests that the lockdown crime drop was not uniform, but instead, the macro-level trend masked underlying variation.
+In the pre-pandemic era, studies have consistently found that local areas (e.g.
+street segments, neighborhoods) do not experience macro-level trends in unison, but rather, change tends to be attributable to a handful of areas [@andresen2017trajectories; @weisburd2012criminology].
+Here, for the first time, we have exploited the quasi-'natural experiment' conditions of the pandemic to decompose this variation during both the _decline_ and _resurgence_ stages of a macro-level trend, over a short period of time.
+
+Findings suggest that the macro-level effect of lockdown on crime has been moderated by highly localized characteristics.
+Previously criminogenic areas, characterised by plentiful opportunities for crime, and often located in city centers, were the most susceptible to decline following the introduction of "stay at home" measures.
+Containing a disproportionately large number of nighttime economy facilities (e.g.
+pubs, nightclubs), commercial outlets (e.g.
+malls, convenience stores), public transport nodes (e.g.
+railway stations, bus stops) and bicycle parking spaces, these areas had typically acted as major crime attractors and generators.
+In March, overnight, as the hospitality sector closed its doors, non-essential shops closed, and transport either closed to became unnecessary, these key features became unavailable or unusable.
+In turn, these areas contributed a considerable (and disproportionate) amount to the nationwide decline in crime.
+
+That being said, it is notable that the local drivers of the lockdown crime drop experienced only a modest resurgence upon the easing of lockdown restrictions.
+By August, the clusters identified as 'high crime major drop' remained well below the crime counts observed in 2018 and 2019, by which point the low and mid-crime clusters had returned to historical levels.
+This can likely be attributed to crime type profile of clusters.
+The drivers of the lockdown crime drop were weighted heavily towards theft (theft from the person, other theft) and shoplifting compared to the low and mid-crime clusters.
+Instead, low and mid-crime areas were characterized by a larger proportion of violence and sexual offences.
+Breaking nationwide trends down by crime type, we observe starkly different resurgence trends across these crime types.
+Types of theft and shoplifting had still not returned to normality by August, while violence and sexual offences had converged back to normality.
+This is evidenced by a descriptive comparison to 2018 and 2019 (see Appendix) and more advanced forecasting techniques which account for long-term trends [see @langton2021cs].
+This may reflect the asymmetric manner of lockdown in England and Wales.
+While it came into force instantaneously and equitably across the country, lockdown was lifted in stages.
+For instance, a select number of commercial outlets (e.g.
+gardening centers) were reopened in May, but restrictions on the hospitality sector and public transport remained in place until July.
+
+Here, we would note some weaknesses in the study.
+First, the crime measure of 'notifiable offences (excluding drugs)' aggregates away some detail in longitudinal trends.
+The crime type profile findings certainly go some way in addressing this, and we have now set a baseline from which further investigations can be conducted.
+However, further insight could be gained from replicating this analyses for major crime types which we know to have distinct opportunity structures, such as burglary and theft.
+Here, we would note that the generalized Gini coefficients by crime type (see Appendix) certainly suggest that there may be differences between crime types.
+Secondly, the Open Street Map API used to quantify the prevalence of criminal opportunities in each cluster is imperfect.
+Studies have found an association between contributions to Open Street Map and contextual characteristics such as poverty and distance from the city center -- factors that are also known to be associated with crime.
+Due to computational issues and API query limits, some features were also unobtainable on a national scale, such as the residential buildings footprints.
+While we would argue that Open Street Map is unique in its ability to offer open data on opportunity structures nationwide, we would encourage future research to verify these findings using other data sources.
+
+# Conclusion
+
 This paper has aimed to investigate the localized drivers of the COVID-19 crime drop in England and Wales.
@@
 Despite their dramatic declines, these areas experienced a fairly slow resurgence upon the relaxation of lockdown rules, remaining below historical levels even by the time low and mid-crime clusters had returned to normality.
-
-The finding that a small number of fine-grained geographic units drove macro-level crime trends, with most areas otherwise remaining fairly stable, is consistent with results from other study sites.
-Evidence from Queensland, Australia [@payneregional2021] and the Chicago, United States [@campedelli2020disentangling] certainly suggests that the lockdown crime drop was not uniform, but instead, the macro-level trend masked underlying variation.
-In the pre-pandemic era, studies have consistently found that local areas (e.g.
-street segments, neighborhoods) do not experience macro-level trends in unison, but rather, change tends to be attributable to a handful of areas [@andresen2017trajectories; @weisburd2012criminology].
-Here, for the first time, we have exploited the quasi-'natural experiment' conditions of the pandemic to decompose this variation during both the _decline_ and _resurgence_ stages of a macro-level trend.
-Interestingly, the degree of concentration, as measured by the generalized Gini coefficient, remained remarkably stable, even amidst this macro and meso-level change.
-This suggests that the law-like behavior of crime concentration [@weisburd2015law] holds even in this highly volatile and unusual unusual scenario.
-
-The opportunity characteristics of the clusters identified certainly warrants further discussion. We found that
-
-- The LSOA which drove the lockdown crime drop tended to have considerable crime attractors and generators, such as public transport nodes, an active nightlife, and plenty of shops and bicycle parking.
-- Thus, while the opportunity structures of these areas makes them high crime areas in typical times, in a lockdown scenario, these are the areas which are most impacted by changes in mobility, driving crime down.
-- The crime type profile of clusters demonstrates considerable stability.
 
@@
 
-```{r longterm, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Select crime types with long-term increases in counts.", out.width= "90%", fig.align='center'}
-knitr::include_graphics("visuals/long_term_gg.png")
-```
-
 # References

### 2021-03-18 scripts/police_data_handling_cleaned.R

@@ -550,6 +550,7 @@ kmeans_violin_full_gg <- ggplot() + theme_bw() +
   geom_violin(data = filter(tc_clusters_agg_df, year == "2020"),
               mapping = aes(x = month_name, y = ew_crime_count, fill = traj_titles),
               alpha = 0.3, colour = "transparent", adjust = 2) + 
+  # scale_fill_brewer(guide = FALSE, palette = "Spectral", direction = -1) +
   scale_fill_hue(guide = FALSE) +
   stat_summary(data = filter(tc_clusters_agg_df, year == "2018"),
                mapping = aes(x = month_name, y = ew_crime_count, group = traj_titles, colour = "blue"),
@@ -845,26 +846,50 @@ lsoa_ew_valid_sf <- sub_data_agg_full_df %>%
   st_as_sf()
 
 # Save.
-st_write(obj = lsoa_ew_valid_sf, dsn = "data/lsoa_ew_valid_k6_sf.shp")
+# st_write(obj = lsoa_ew_valid_sf, dsn = "data/lsoa_ew_valid_k6_sf.shp")
+
+# Add OSM data to define city centres.
+osm_df <- read_csv("data/osm_full.csv")
+
+# Create sums (see below for actual OSM statistics by cluster).
+osm_df <- osm_df %>% 
+  mutate(transport_total = trains + bus,
+         shops_total     = shops_total + conveniences) %>% 
+  mutate(sum = nightlife_total + transport_total + bikes + shops_total)
+
+lsoa_ew_valid_osm_sf <- lsoa_ew_valid_sf %>% 
+  left_join(osm_df, by = c("lsoa_code" = "geo_code", "geo_label" = "geo_label"))
 
 # Selection of cities. Top 5 by population in England minus London plus Cardiff.
-birm_sf <- lsoa_ew_valid_sf %>% 
-  filter(str_detect(string = lsoa_ew_valid_sf$geo_label, pattern = "Birmingham"))
+birm_sf <- lsoa_ew_valid_osm_sf %>% 
+  filter(str_detect(string = lsoa_ew_valid_osm_sf$geo_label, pattern = "Birmingham")) %>% 
+  mutate(centre1 = as.integer(sum == max(sum)),
+         centre2 = as.integer(trains == max(trains)))
 
-leeds_sf <- lsoa_ew_valid_sf %>% 
[…]

### 2021-03-18 draft_paper.Rmd
drafting edits

@@
 ---
-title: "Local drivers of the COVID-19 lockdown crime drop in England and Wales"
+title: "Spatial patterning of the COVID-19 lockdown crime drop in England and Wales"
 author:
@@
 Lockdown mobility patterns appear to have disrupted the typical convergence of motivated offenders, suitable targets and capable guardians, reducing opportunities for victimization [@stickle2020crime].
-As lockdown restrictions were slowly repealed, routine activities converged back to pre-pandemic levels and we witnessed a resurgence in criminal activity [@langton2021cs].
+As lockdown restrictions were slowly repealed, routine activities converged back to pre-pandemic levels and many countries witnessed a resurgence in criminal activity [e.g.
+@langton2021cs].
 
@@
 Recent months have seen a wave of interest in describing and explaining how crime has changed in response to restrictions on mobility brought about to limit the spread of COVID-19.
-Studies have been undertaken in the United States [@campedelli2020exploring; @gerell2020minor; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show] and Mexico [@estevez2020crime].
+Studies have been undertaken in the United States [@campedelli2020exploring; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show], Sweden [gerell2020minor], China (@borrion2020measuring) and Mexico [@estevez2020crime].
 Generally speaking, findings have been consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@langton2021cs; @halford2020crime].
@@
 With the necessary convergence of offenders and victims disrupted, often quite literally overnight, we would expect these customarily high crime places to fall considerably, prompting a macro-level decline.
-By contrast, areas typically devoid of criminal opportunity, with say, little or no ambient populations, may remain largely unaffected by lockdown restrictions on mobility, and in turn, contribute little to any macro-level change.
+By contrast, areas typically devoid of crime opportunity, with say, little or no ambient populations, may remain largely unaffected by lockdown restrictions on mobility, and in turn, contribute little to any macro-level change.
 In other words, while lockdowns have often been imposed equitably at a city or national level, the effect on crime will likely be moderated by the opportunity structure of local areas.
@@
 Using regional units of analysis nested within the state of Queensland, Australia, @payneregional2021 found a degree of diversity in crime rate trends, suggesting that the lockdown crime drop was not 'universal'.
-In the United Kingdom, however, we have not seen a comprehensive decomposition of local variation and longitudinal (in)stability underpinning the lockdown crime drop and subsequent resurgence, or indeed an exploration of the opportunity structures characterizing the local areas which have driven macro-level changes.
+
+In England and Wales, early descriptive evidence from the first three months of lockdown suggests that previously high-crime areas may have experienced the steepest proportional declines [@dixon2020crime].
+Increases in fly-tipping have also been linked to a small number of councils, with trends varying considerably between regions [@dixon2020fly].
+However, there has not been a comprehensive decomposition of local variation and longitudinal (in)stability underpinning the nationwide lockdown crime drop and subsequent resurgence, or indeed an exploration of the opportunity structures characterizing the local areas which have driven macro-level changes.
 
@@
 Each of these are now outlined in turn, followed by an outline of the methods deployed.
-Code to replicate the data downloads, handling, analyses and visualization reported here are openly available (https://github.com/langtonhugh/covid_spatial).
+Code to replicate the data downloads, handling, analyses and visualization reported here are openly available (anonymized GitHub repo link here).
 
@@
 But, one key strength of k-means is that it is also capable of identifying short-term fluctuation in longitudinal trends [@akmed2021mon].
-In the lockdown scenario, criminal opportunities were withdrawn quite literally overnight, and thus we might expect a similarly rapid and short-term change in crime at meso spatial scales.
+In the lockdown scenario, crime opportunities were withdrawn quite literally overnight, and thus we might expect a similarly rapid and short-term change in crime at meso spatial scales.
 The ability to unpick these rapid changes represents a key strength of k-means over non-parametric techniques such as anchored k-medoids, which are designed for long-term rather than short-term change [@akmed2021mon] or group-based trajectory modelling, which is limited by polynomial terms [@griffiths2004communities].
@@
 Study regions are defined by based on the city names appearing in LSOA name.
+For the purposes of the visual, one LSOA in Cardiff containing Flat Holm Island, which is off the coast, was been removed.
 
@@
 
-The provide local context to the cluster solutions identified through the k-means clustering, we visualize the spatial patterning of groupings for six major cities in England and Wales (see Figure 5).
+To provide local context to the cluster solutions identified through the k-means clustering, we visualize the spatial patterning of groupings for six major cities in England and Wales (see Figure 5).
 We find distinct geographic patterns to the clusters identified.
-Maps denote a single 'city center' based on the LSOA with the highest count of Open Street Map features (i.e.
+The maps denote a single 'city center' based on the LSOA with the highest count of Open Street Map features (i.e.
 the sum of nightlife, shops, public transport and bicycle parking).
+Because LSOA are uniform by residential population, geographically small LSOA tend to be areas with higher population density.
 Here, it is worth noting that the Bradford region contains the city of Bradford, but also the satellite town of Keighley to the north west.
-Without exception, the city centres and commercial districts of each city are characterised by 'high crime major drop' clusters E and F, which in turn often neighbor the 'mid crime mid drop' cluster D.
-By contrast, the 'low crime and stable' clusters tend to sit outside of the city centres: cluster B in the suburbs and the sparsely populated cluster A on the periphery of urban conurbations.
+
+Without exception, the city centers and commercial districts of each city are characterised by 'high crime, major drop' clusters E and F, which in turn often neighbor the 'mid crime, mid drop' cluster D.
+By contrast, the 'low crime and stable' clusters tend to sit outside of the city centres: cluster B in the suburbs and the sparsely populated cluster A on the periphery of each urban conurbation.
+As such, changes in crime during lockdown appear to have a distinct geographic pattern: the nationwide drop was not experienced uniformly across space, but rather, particular areas including city centers appear to have driven the macro-level trend, with outer suburbs remaining fairly stable.
 For a detailed and interactive investigation of the spatial clusters identified for the whole of England and Wales, we refer readers to an openly available online map (anonymized link here).
@@
 ```{r  maps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Spatial distribution of clusters by major urban conurbation.
-LSOA with the highest count of Open Street Map features denotes city centre.", out.width= "100%", fig.align='center'}
+LSOA with the highest count of Open Street Map features (in black) denotes city centre.", out.width= "100%", fig.align='center'}
 knitr::include_graphics("visuals/maps_gg.png")
@@
 Table 1 reports a series of descriptive statistics based on the opportunity structure of each cluster using the nationwide data obtained from the Open Street Map API.
-These findings largely support our expectations, namely, that large, stable clusters have few opportunities for crime, while high crime areas which were responsible for a disproportionately large amount of the crime drop have plentiful features which facilitate crime.
+These findings largely support our expectations, namely, that large, stable clusters have few opportunities for crime, while high crime areas which were responsible for a disproportionately large amount of the crime drop contain plentiful features which facilitate crime.
 
@@
 However, further insight could be gained from replicating this analyses for major crime types which we know to have distinct opportunity structures, such as burglary and theft.
-Here, we would note that the generalized Gini coefficients by crime type (see Appendix) certainly suggest that there may be differences between crime types.
-Secondly, the Open Street Map API used to quantify the prevalence of criminal opportunities in each cluster is imperfect.
+Here, we would note that the generalized Gini coefficients by crime type (see Appendix) certainly suggest that there may be differences between crime types, as does preliminary findings using decile clusters
+
+Secondly, the Open Street Map API used to quantify the prevalence of crime opportunities in each cluster is imperfect.
 Studies have found an association between contributions to Open Street Map and contextual characteristics such as poverty and distance from the city center -- factors that are also known to be associated with crime.

### 2021-03-20 draft_paper.Rmd
anony process

@@
 ---
-title: "Spatial patterning of the COVID-19 lockdown crime drop in England and Wales"
+title: "Spatial distribution of the COVID-19 lockdown crime drop in England and Wales"
 author:
@@
 output:
-  pdf_document: default
   word_document:
     reference_docx: reference_doc.docx
+  pdf_document: default
 bibliography: refs.bib
@@
 ```{r setup, include=FALSE, echo = F}
-knitr::opts_chunk$set(echo = FALSE, eval = T) 
+knitr::opts_chunk$set(echo = FALSE, eval = F) 
 library(readr);library(kableExtra)
@@
 Studies have been undertaken in the United States [@campedelli2020exploring; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show], Sweden [gerell2020minor], China (@borrion2020measuring) and Mexico [@estevez2020crime].
-Generally speaking, findings have been consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@langton2021cs; @halford2020crime].
+Generally speaking, findings have been consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@halford2020crime].
 That is, legally-enforced restrictions on daily activities, mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime.
-As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often back to levels that would have been expected without the global pandemic and ensuing lockdowns.
+As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often back to levels that would have been expected without the global pandemic and ensuing lockdowns [@langton2021cs]
 
@@
 
-In England and Wales, early descriptive evidence from the first three months of lockdown suggests that previously high-crime areas may have experienced the steepest proportional declines [@dixon2020crime].
+In England and Wales, early descriptive evidence from the first three months of lockdown suggests that previously high-crime areas may have experienced the steepest relative declines compared to previous years [@dixon2020crime].
 Increases in fly-tipping have also been linked to a small number of councils, with trends varying considerably between regions [@dixon2020fly].
@@
 Each of these are now outlined in turn, followed by an outline of the methods deployed.
-Code to replicate the data downloads, handling, analyses and visualization reported here are openly available (anonymized GitHub repo link here).
+Code to replicate the data downloads, handling, analyses and visualization reported here are openly available (https://anonymous.4open.science/r/46199882-cc5a-449d-9946-300b3132ab2e/).
 
@@
 As such, changes in crime during lockdown appear to have a distinct geographic pattern: the nationwide drop was not experienced uniformly across space, but rather, particular areas including city centers appear to have driven the macro-level trend, with outer suburbs remaining fairly stable.
-For a detailed and interactive investigation of the spatial clusters identified for the whole of England and Wales, we refer readers to an openly available online map (anonymized link here).
+For a detailed and interactive investigation of the spatial clusters identified for the whole of England and Wales, we refer readers to an openly available online map (anonymous for review).
 
@@
 However, further insight could be gained from replicating this analyses for major crime types which we know to have distinct opportunity structures, such as burglary and theft.
-Here, we would note that the generalized Gini coefficients by crime type (see Appendix) certainly suggest that there may be differences between crime types, as does preliminary findings using decile clusters
-
+Here, we would note that the generalized Gini coefficients by crime type (see Appendix) certainly suggest that there may be differences between crime types, as have preliminary findings using decile clusters [@dixon2020crime].
 Secondly, the Open Street Map API used to quantify the prevalence of crime opportunities in each cluster is imperfect.

### 2021-03-22 draft_paper.Rmd
submission docs

@@
 author:
-  - Samuel Langton
-  - Anthony Dixon
-  - Graham Farrell
+- Samuel Langton (University of Leeds)
+- Anthony Dixon (University of Leeds) 
+- Graham Farrell (University of Leeds)
 csl: apa.csl
-header-includes: \usepackage{float}
 output:
@@
     reference_docx: reference_doc.docx
-  pdf_document: default
+header-includes: \usepackage{float}
 bibliography: refs.bib
@@
 ```
+
+**Abstract**
+
+Studies from international case study sites have documented the impact of the COVID-19 pandemic and ensuing lockdown policies on crime.
+Restrictions on mobility and social interaction have brought about widespread falls in police-recorded crime, swiftly followed by a resurgence as lockdowns were lifted.
+Less is known about the localized spatial patterning of these macro-level trends.
+This study uses police-recorded data on notifiable offences to decompose the nationwide lockdown trend in England and Wales.
+We find that local, meso-level areas did not experience the decline and subsequent 'bounce back' in unison.
+Rather, most areas remained remarkably stable, with a small number of meso units contributing a disproportionate amount to the macro-level trend.
+We find that these areas often occupy city centers, had plentiful opportunities for crime pre-pandemic and were dominated by theft and shoplifting offences.
+
+**Keywords:** COVID-19, clustering, kmeans, routine activities, opportunity theory, pandemic.
+
+**Corresponding author:** Graham Farrell, School of Law, University of Leeds, United Kingdom.
+
+Email: g.farrell@leeds.ac.uk.
+
+**Competing interests:** The authors declare that they have no competing interests.
+
+**Availability of data and materials:** The data are publicly available and retrievable from the open online data portal for England, Wales and Northern Ireland.
+Code to replicate data downloads, handling and analyses is openly available (https://anonymous.4open.science/r/46199882-cc5a-449d-9946-300b3132ab2e/).
+
+\newpage
 
@@
 Recent months have seen a wave of interest in describing and explaining how crime has changed in response to restrictions on mobility brought about to limit the spread of COVID-19.
-Studies have been undertaken in the United States [@campedelli2020exploring; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show], Sweden [gerell2020minor], China (@borrion2020measuring) and Mexico [@estevez2020crime].
+Studies have been undertaken in the United States [@campedelli2020exploring; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show], Sweden [@gerell2020minor], China [@borrion2020measuring] and Mexico [@estevez2020crime].
 Generally speaking, findings have been consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@halford2020crime].
 That is, legally-enforced restrictions on daily activities, mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime.
-As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often back to levels that would have been expected without the global pandemic and ensuing lockdowns [@langton2021cs]
+As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often back to levels that would have been expected without the global pandemic and ensuing lockdowns [@langton2021cs].
 
@@
 We used a study period spanning February to August 2020 in order to capture the first six months of the nationwide lockdown (March to August) and the one month preceding the change (February).
+Here, we note that lockdown was initiated on 23 March, making April the first full month of measures.
 For reference and comparison to historical trends, we obtained data for the same months in 2018 and 2019.
@@
 To summarize the opportunity structure of local areas we required a nationwide dataset of theoretically relevant facilities and urban features which could be aggregated at the LSOA level.
-To this end, we obtained point-level data from the Application Programming Interface (API) for Open Street Map via the `osmdata` package [@padgham2017osmdata] in `R` [@team2013r].
+To this end, we obtained point-level data from the Application Programming Interface (API) for Open Street Map via the _osmdata_ package [@padgham2017osmdata] in _R_ [@team2013r].
 Open Street Map is a crowdsourced geospatial database containing a vast array of features which can be used for explaining the temporal and spatial patterning of crime [@langton2020open].
@@
 We visualize observed counts between February and August 2020 relative to the same periods in 2018 and 2019.
-In doing so, we can observe how nationwide crime trends changed in the face of lockdown measures [see also @langton2021cs].
+In doing so, we can observe how crime trends changed in the face of lockdown measures in England and Wales [see also @langton2021cs].
 This sets the scene from which we can disentangle the underlying meso-level variation.
@@
 
-We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offence (excluding drug) counts for LSOA in England and Wales (N = 33,075) between February and August 2020.
+We deploy k-means using the _kml_ package [@kmlpack] in R on notifiable offence (excluding drug) counts for LSOA in England and Wales (N = 33,075) between February and August 2020.
 To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite].
@@
 
-To set the context for the localized analysis, Figure 2 visualizes crime counts between February and August 2020 for notifiable offences excluding drugs before and after lockdown.
-This descriptive analysis broadly confirms what has been demonstrated using model-based forecasting analyses elsewhere (Langton et al., 2021) although we note that this descriptive comparison does not account for long-term trends.
+To set the context for the localized analysis, Figure 1 visualizes crime counts between February and August 2020 for notifiable offences excluding drugs before and after lockdown.
 In April, the first full month of lockdown in England and Wales, we observe a nationwide decline in notifiable offences in comparison to previous years.
-Upon the relaxation of lockdown rules, crime resurged back to normality, and by August, crime had returned to within a range we might have expected without the nationwide lockdown.
+Upon the relaxation of lockdown rules, crime resurged back to normality, and by August, crime had returned to within a range we might have expected without the nationwide lockdown [see also @langton2021cs].
 This trend represents the 'global' trend which we will subsequently disentangle using localized analyses.
 
-```{r nocounts, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Notitifable offence (excluding drugs) counts in England and Wales during the lockdown period.", fig.align='center'}
+```{r nocounts, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Notifiable offence (excluding drugs) end of month counts in England and Wales.
+April was the first full month of lockdown.", fig.align='center'}
 knitr::include_graphics("visuals/no_raw_counts_gg.png")
 ```
+
+[Insert Figure 1 here. Caption: Notifiable offence (excluding drugs) end of month counts in England and Wales. April was the first full month of lockdown.]
 
@@
 We deploy non-parametric k-means clustering on LSOA (meso) level geographic units (N = 33,075) comprising England and Wales to decompose the macro-level 'decline and resurgence' observed between February and August, 2020.
-Based on an assessment of the Calinski-Criterion statistic [@calinski1974dendrite] and Principal Component Analysis (see Appendix), we determined the optimal solution to be 6 clusters (see Figure 3).
+Based on an assessment of the Calinski-Criterion statistic [@calinski1974dendrite] and Principal Component Analysis, we determined the optimal solution to be 6 clusters (see Figure 2).
 Solid black lines represent the median count for each cluster at any given time point, with the black dotted line showing the mean.
@@
 Together, these clusters comprise 89% of LSOA in England and Wales.
-Even amidst the stark macro-level decline in notifiable offences (see Figure 2) LSOA in these clusters only experienced marginal average dips in crime.
+Even amidst the stark macro-level decline in notifiable offences (see Figure 1) LSOA in these clusters only experienced marginal average dips in crime.
 The third largest cluster, cluster C, comprises 10% of LSOA in England and Wales.
@@
 
+[Insert Figure 2 here. Caption: K-means cluster solutions for LSOA notifiable offences (excluding drugs). Distributions refer to 2020 only.]
+
 That said, most 'action' appears to be occurring amongst a small subset of LSOA.
@@
 
-To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the monthly change in counts and percentage of total absolute change (i.e.
+To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 3 plots the monthly change in counts and percentage of total absolute change (i.e.
 positive or negative) attributable to each cluster.
@@
 That said, the figure demonstrates that this decline did not occur equitably across local areas.
-Consistent with the cluster solutions trends in Figure 3, a small number of LSOA (meso) units appear to have contributed disproportionately to the nationwide (macro) trend.
+Consistent with the cluster solutions trends in Figure 2, a small number of LSOA (meso) units appear to have contributed disproportionately to the nationwide (macro) trend.
 For instance, clusters A and B, which comprise 58% and 31% of LSOA in the country, accounted for only 20% and 27% of the total decline between March and April.
@@
 The 'low crime and stable' clusters (A and B) actually _declined again_ between July and August, and the higher crime clusters continued to increase.
-The change is almost imperceptible when viewing average counts in these LSOA (see Figure 3) but the sheer size of these units, collectively comprising 89% of England and Wales, has a major impact on nationwide trends.
+The change is almost imperceptible when viewing average counts in these LSOA (see Figure 2) but the sheer size of these units, collectively comprising 89% of England and Wales, has a major impact on nationwide trends.
 This is suggestive of highly localized change in opportunities as lockdown rules were relaxed: change which is aggregated away at the macro-level.
@@
 
+[Insert Figure 3 here. Caption: Counts and percentage of nationwide change between months attributable to each cluster.]
+
 These findings can be supplemented with a comparison to the previous year.
-Figure 5 compares the clusters identified in 2020 with the equivalent months in 2019.
+Figure 4 compares the clusters identified in 2020 with the equivalent months in 2019.
 Here, it is clear that the 'high crime major drop' clusters (E and F) contributed the most to the lockdown crime drop relative to their size.
-By way of example, around 27,000 fewer crimes occurred in cluster A--containing 19,162 LSOA--during April 2020 compared to April 2019 (Figure 5a).
-This translates to a 24% drop (Figure 5b), and an average fall per LSOA of 1.39 counts (Figure 5c).
-For cluster F, containing just 9 LSOA, approximately 5,000 fewer crimes occurred (Figure 5a), which translates to an 87% drop (Figure 5b), and an average fall per LSOA of 543 counts (Figure 5c).
+By way of example, around 27,000 fewer crimes occurred in cluster A--containing 19,162 LSOA--during April 2020 compared to April 2019 (Figure 4a).
+This translates to a 24% drop (Figure 4b), and an average fall per LSOA of 1.39 counts (Figure 4c).
+For cluster F, containing just 9 LSOA, approximately 5,000 fewer crimes occurred (Figure 4a), which translates to an 87% drop (Figure 4b), and an average fall per LSOA of 543 counts (Figure 4c).
 In other words: whether comparing month-to-month change in 2020, or comparing equivalent months in 2019, the lockdown crime drop was disproportionately driven by a small number of local areas, with most meso-level units remaining fairly stable, even amidst dramatic macro-level change.
@@
 ```
+
+[Insert Figure 4 here.
+Caption: Comparing LSOA clusters between equivalent months in 2020 and 2019 according to (a) absolute difference in counts, (b) percentage difference in counts, (c) average difference in counts.]
 
@@
 ```
+
+[Insert Figure 5 here. Caption: Spatial distribution of clusters by major urban conurbation. LSOA with the highest count of Open Street Map features (in black) denotes city centre.]
 
@@
 
+[Insert Table 1 here. Caption: Descriptive statistics of facilities and features in each cluster. Sourced from Open Street Map.]
+
 ### Crime type profile
@@
 
-```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type characteristics of each cluster solution.", out.width= "100%", fig.align='center'}
+```{r crimeprofile, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Crime type profiles of each cluster solution.", out.width= "100%", fig.align='center'}
 knitr::include_graphics("visuals/traj_crimes_monthly_k6_gg.png")
 ```
+
+[Insert Figure 6 here. Caption: Crime type profiles of each cluster solution.]
 
@@
 We find that these localized drivers, dominating city centers, had pre-existing high levels of crime, likely attributable to their unique opportunity structures, containing numerous nightlife facilities, commercial shops, public transport nodes and bicycle parking spaces.
-Clusters demonstrate a degree of stability in their crime type profiles, but the 'high crime major drop' areas appear to be weighted towards theft and shoplifting.
-Despite their dramatic declines, these areas experienced a fairly slow resurgence upon the relaxation of lockdown rules, remaining below historical levels even by the time low and mid-crime clusters had returned to normality.
+Clusters demonstrate a degree of stability in their crime type profiles, but the 'high crime, major drop' areas appear to be weighted towards theft and shoplifting.
+Despite their dramatic declines, these areas experienced a fairly shallow resurgence upon the relaxation of lockdown rules, remaining below historical levels even by the time low and mid-crime clusters had returned to normality.
 
@@
 
-```{r crimegini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Monthly generalized Gini cofficient trends by offence type in England and Wales during the lockdown period.", fig.align='center'}
+[Insert Figure 7 here. Caption: Counts by offence type in England and Wales during the lockdown period.]
+
+```{r crimegini, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Monthly generalized Gini coefficient trends by offence type in England and Wales during the lockdown period.", fig.align='center'}
 knitr::include_graphics("visuals/gini_gg.png")
@@
 
+[Insert Figure 8 here. Caption: Monthly generalized Gini coefficient trends by offence type in England and Wales during the lockdown period.]
+
 # References

### 2021-03-22 scripts/police_data_handling_cleaned.R

@@ -280,7 +280,7 @@ total_crime_agg_df <- sub_data_agg_full_df %>%
 # Single plot for notifiable offences.
 no_raw_counts_gg <- ggplot(data = total_crime_agg_df) +
   geom_line(mapping = aes(x = month, y = ew_crime_count, group = year, colour = year), size = 0.8) +
-  geom_vline(xintercept = 1.7, linetype = "dotted") +
+  # geom_vline(xintercept = 1.7, linetype = "dotted") +
   scale_x_discrete(labels = str_extract(month.name[2:9], "^.{3}")) +
   scale_color_manual(values = rev(c("black", "darkgrey", "lightgrey"))) +
   labs(x = NULL, y = "Count", colour = NULL) +
